---
title: "R Notebook"
output: html_notebook
---


```{r}
install.packages("devtools")
devtools::install_github("naef-lab/dryR")
install.packages('simphony')
install.packages('stringr')

library(tidyverse)
library("dryR")
library(stringr)
library(ggplot2)
library(ggpubr)

```

```{r}
processed_data <- read_csv("Bignon_RNA-seq_processed.csv")
processed_data
```

```{r}

# prepare arguments
countData = simData[["countData"]]
group     = simData[["group"]]
time      = simData[["time"]]

# prefilter
keep <- rowSums(countData) > 0
countData <- countData[keep,]

# run the analysis for count data (e.g. RNA-Seq data)
dryList   = dryseq(countData,group,time)

# explore the results
dryList[["results"]]     # data frame summarizing results
dryList[["parameters"]]  # coefficients: phase, amplitude and mean for each group
dryList[["ncounts"]]     # normalized counts
dryList[["counts"]]      # raw counts
dryList[["cook"]]        # cook's distance for outlier detection
dryList[["BICW_rhythm"]] # BICW for each rhythmic model
dryList[["BICW_mean"]]   # BICW for each mean model

# generate a pdf with a global summary of all models
plot_models_rhythm(dryList, "./")

# plot a feature of interest
dry_plot(dryList, "feature_113")

```

### Generate processed count matrix

```{r}
processed_data
count_data_drylm <- processed_data |>
    column_to_rownames("Gene_ID") |>
    select(-c(Gene_Name, description, Gene_type))

# Check for missing values or empty rows
table(rowSums(count_data_drylm) == 0) 
table(is.na(count_data_drylm))

group_drylm <- as.data.frame(t(count_data_drylm)) |>
    rownames_to_column("Label") |>
    select(Label) |> t()

time_drylm <- str_extract(group_drylm, 'ZT(\\d*)_', group = 1) |> strtoi (base=10)

# test <- 'ZT0_KO3'
group_drylm_new <- str_extract(group_drylm, 'ZT\\d*_(\\D*)', group = 1)
group_drylm_new

replicate_drylm <- c(1:5)
replicate_drylm <- rep(replicate_drylm, times=12)


```

```{r}

drylm_result = drylm(count_data_drylm,group_drylm_new,time_drylm, period = 24)

# explore the results
drylm_result[["results"]]     # data frame summarizing results
drylm_result[["parameters"]]  # coefficients: phase, amplitude and mean for each group
drylm_result[["ncounts"]]     # normalized counts
drylm_result[["counts"]]      # raw counts (NULL)
drylm_result[["cook"]]        # cook's distance for outlier detection (NULL)
drylm_result[["BICW_rhythm"]] # BICW for each rhythmic model
drylm_result[["BICW_mean"]]   # BICW for each mean model

# generate a pdf with a global summary of all models
plot_models_rhythm(drylm_result, "./")

# plot a feature of interest
dry_plot(drylm_result, "feature_113")



```

This is what the Bignon paper did, they used drylm on logCPM normalised (?) data.

### Generate raw count matrix

```{r}
install.packages("BiocManager")
BiocManager::install("GEOquery")
library(GEOquery)
library(stringr)
library(tidyr)

gse <- getGEO(filename = "Count data/GSE216252_series_matrix.txt.gz", GSEMatrix = TRUE,
    getGPL = FALSE)

meta <- gse[[1]]
gsm <- gse[[2]]
label_df <- as.data.frame(t(rbind(gsm, meta))) |> separate_wider_delim(cols = meta, delim = ", ", names = c("sample", "genotype", "timepoint", "replicate")) |> separate_wider_delim(cols = replicate, delim = " ", names = c("rep", "replicate")) |> separate_wider_delim(cols = genotype, delim = " ", names = c("genotype", "gen"))

label_df <- label_df |>
  mutate(genotype_new = recode(genotype, control = "C", cKO = "KO")) |> 
  mutate(timepoint_new = str_remove(timepoint, "ZT")) |>
  dplyr::select(gsm, genotype_new, timepoint_new, replicate)


```

```{r}
### run whole chunk
setwd("Count data/GSE216252_RAW")

file_list <- list.files(".")
count_matrix <- read.delim(file_list[1]) |> dplyr::select(Gene_ID)

for (file in file_list){
  gsm <- str_extract(file, '([[:alnum:]]*)_\\S*', group = 1)
  count_matrix <- read.delim(file) |> 
      rename_with(~gsm, 2) |> 
    full_join(count_matrix)
}

```

```{r}
summary(count_matrix)
sum(is.na(count_matrix))
count_matrix
```

```{r}
# sort gsm (i.e.) in count_matrix
count_data_dryseq <- count_matrix[1:(nrow(count_matrix)-5),] |> remove_rownames() |>
  column_to_rownames("Gene_ID") 
count_data_dryseq <- count_data_dryseq[,order(colnames(count_data_dryseq))] |> as.matrix()

tail(count_data_dryseq)
  
is.unsorted(label_df$gsm) # check if the metadata is sorted

group_dryseq <- label_df$genotype_new 
time_dryseq <- strtoi(label_df$timepoint_new, base=10)
```


```{r}
# prefilter
keep <- rowSums(count_data_dryseq) > 0
count_data_dryseq <- count_data_dryseq[keep,]
nrow(count_data_dryseq)

dryseq_result = dryseq(count_data_dryseq, group_dryseq, time_dryseq)

# explore the results
dryseq_result[["results"]]     # data frame summarizing results
dryseq_result[["parameters"]]  # coefficients: phase, amplitude and mean for each group
dryseq_result[["ncounts"]]     # normalized counts
dryseq_result[["counts"]]      # raw counts
dryseq_result[["cook"]]        # cook's distance for outlier detection
dryseq_result[["BICW_rhythm"]] # BICW for each rhythmic model
dryseq_result[["BICW_mean"]]   # BICW for each mean model

# generate a pdf with a global summary of all models
plot_models_rhythm(dryseq_result, "./")

# plot a feature of interest
dry_plot(dryseq_result, "ENSMUSG00000000440")
dry_plot(dryseq_result, "ENSMUSG00000000441")
dry_plot(dryseq_result, "ENSMUSG00000000486")

dryseq_parameters <- dryseq_result[["parameters"]]
table(dryseq_parameters$chosen_model)
```
    1     2     3     4     5 
26307  1369   786  4222   240

Bruhhh it seems that the RNA-seq data is too skewed (with a few extremely big values), and it breaks the negative binomial (NB) algorithm in dryseq...

HUH why did it suddenly work again when I changed to LMB mac (R version issue? or did I not filter away the rowsum=0???)


```{r}
count_data_dryseq_sorted <- as.data.frame(count_data_dryseq[order(rowSums(count_data_dryseq), decreasing=TRUE),])

BiocManager::install("AnnotationHub")
BiocManager::install("ensembldb")
library(AnnotationHub)
library(AnnotationDbi)
library(ensembldb)

ah <- AnnotationHub()

# search for id of desired dataset
dbResult <- query(ah, c("EnsDb", "Mus musculus", "102")) 
dbResult[1]

# load mouse transcript dataset
MouseEnsDb <- ah[[dbResult$ah_id]]
MouseEnsDb

# output data of interest in form of data.frame
annotations <- genes(MouseEnsDb, return.type = "data.frame")
annotations

annotation_selection <- annotations |> 
    dplyr::select(gene_id, gene_name, description, entrezid) |>
    dplyr::filter(gene_id %in% rownames(count_data_dryseq_sorted))

length(annotation_selection$entrezid)
length(unique(annotation_selection$entrezid))
sum(is.na(annotation_selection$entrezid))



count_data_dryseq_sorted_annot <- count_data_dryseq_sorted |> 
    rownames_to_column("gene_id") |>
    left_join(annotation_selection, "gene_id") |>
    relocate(c(gene_name, description, entrezid), .after=gene_id)
```

- The extremely high read are all related to kidney functioning / respiration
bruhhh okay let's go with drylm first

### Drylm result analysis
```{r}
BiocManager::install("limma")
library(limma)

drylm_parameters <- drylm_result[["parameters"]]
#tail(drylm_result[["parameters"]])

table(drylm_parameters$chosen_model)
nrow(drylm_parameters)
nrow(count_data_drylm)
nrow(count_data_dryseq)
sum(rowSums(count_data_dryseq) > 100)

```
   1    2    3    4    5 
7021 1624 1067 5252  582 


Why the number of transcripts in in drylm dataset is so low?? what did they remove??

```{r}
# plot a feature of interest
# model 3: gain-of rhythm
dry_plot(drylm_result, "ENSMUSG00000022698")
dry_plot(drylm_result, "ENSMUSG00000054893")

# model 5: altered rhythm
dry_plot(drylm_result, "ENSMUSG00000059824")
dry_plot(drylm_result, "ENSMUSG00000028957")

# model 2: loss of rhythm
dry_plot(drylm_result, "ENSMUSG00000026077")
```
```{r}
ggpaired(drylm_parameters[1:100,], cond1 = "mean_C", cond2 = "mean_KO")
#plot(drylm_parameters$mean_C, drylm_parameters$mean_KO)
```

Oh it seems like the reason they choose limma over DEseq is because limma accepts normalised values but DEseq only accepts count data!

```{r}
drylm_parameters_limma <- dplyr::select(drylm_parameters, c(mean_C, mean_KO))
group_limma <- factor(c("C", "KO"))
design <- model.matrix(~ group_limma)

fit <- lmFit(drylm_parameters_limma, design)
# fit <- eBayes(fit)
# topTable(fit, coef = "groupTreatment", number = Inf)

count_data_drylm

```

__You cannot perform a significance test on two means unless you know the variability behind those means.__ HUHHHHHH

```{r}
meta_drylm <- rbind(gse[[2]], group_drylm, group_drylm_new, time_drylm, replicate_drylm)
meta_drylm <- as.data.frame(t(meta_drylm)) |> dplyr::rename(gene_id = V1, label = Label, group = group_drylm_new, time = time_drylm, replicate = replicate_drylm)
meta_drylm

# calculate mean value over timepoints


```


### Comparing the rhythmic transcripts from two methods

drylm:
   1    2    3    4    5 
7021 1624 1067 5252  582 

dryseq:
    1     2     3     4     5 
26307  1369   786  4222   240

1 nonrhythmic
2 loss of rhythm
3 gain
4 unaltered rhythm
5 altered 

```{r}

common_2 <- drylm_parameters |> 
    rownames_to_column("gene_id") |> 
    dplyr::filter(chosen_model == 2) |>
    dplyr::filter(gene_id %in% rownames(dplyr::filter(dryseq_parameters, chosen_model == 2)))
 
nrow(common_2)

common_3 <- drylm_parameters |> 
    rownames_to_column("gene_id") |> 
    dplyr::filter(chosen_model == 3) |>
    dplyr::filter(gene_id %in% rownames(dplyr::filter(dryseq_parameters, chosen_model == 3)))
 
nrow(common_3)

common_4 <- drylm_parameters |> 
    rownames_to_column("gene_id") |> 
    dplyr::filter(chosen_model == 4) |>
    dplyr::filter(gene_id %in% rownames(dplyr::filter(dryseq_parameters, chosen_model == 4)))
 
nrow(common_4)

common_5 <- drylm_parameters |> 
    rownames_to_column("gene_id") |> 
    dplyr::filter(chosen_model == 5) |>
    dplyr::filter(gene_id %in% rownames(dplyr::filter(dryseq_parameters, chosen_model == 5)))
 
nrow(common_5)

dryseq_result


```

erm two methods not completely the same but ok-ish overlap



