---
title: "Aviram pipeline analysis"
author: "Elysia Luo"
date: "2026-02-05"
output: html_document
---

```{r}
mamba_folder <- "/Users/elysialuo/miniforge3"
env_name <- "circadian"
.libPaths(paste(mamba_folder, "envs", env_name, "lib/R/library", sep = "/"))

library(tidyverse)
library(dryR)
library(stringr)
library(ggplot2)
library(ggpubr)

library(MASS)
library(janitor)
library(naniar)
library(patchwork)
library(scales)
library(cluster)

library(RColorBrewer)
library(ggrepel)

library(GEOquery)
library(stringr)
library(tidyr)

library(AnnotationHub)
library(AnnotationDbi)
library(ensembldb)

library(limma)
library(edgeR)
library(RUVSeq)

library(limorhyde)
library(annotate)
library(data.table)
library(org.Mm.eg.db)
library(foreach)

library(VennDiagram)

library('org.Mm.eg.db')
library(GEOquery)
library(stringr)
library(tidyr)
library(plotly)
library("RColorBrewer")
```

```{r}
getwd()
```

```{r}
ah <- AnnotationHub()

# search for id of desired dataset
dbResult <- query(ah, c("EnsDb", "Mus musculus", "102")) 
dbResult[1]

# load mouse transcript dataset
MouseEnsDb <- ah[[dbResult$ah_id]]
MouseEnsDb

# output data of interest in form of data.frame
annotations <- genes(MouseEnsDb, return.type = "data.frame")
annotations
```


```{r}
# read in raw count matrix from Aviram
gse <- getGEO(filename = "../../data/Aviram_raw_data/GSE171975_series_matrix.txt", GSEMatrix = TRUE, getGPL = FALSE)

metadata <- pData(gse) |> 
  dplyr::mutate("time point:ch1" = gsub("\\D", "", `time point:ch1`)) |> 
  dplyr::rename(genotype = "genotype:ch1", time = "time point:ch1") |>
  dplyr::select(title, geo_accession, genotype, time) |> 
  dplyr::filter(genotype != "Bmal1-/-") 

metadata <- metadata[order(as.integer(metadata$time)),]
metadata <- metadata[order(metadata$genotype, decreasing = TRUE),]

raw_count_WT <- read_csv("../../data/Aviram_raw_data/Aviram_WT_raw.csv") |> dplyr::rename(gene_name = ...1)
raw_count_Per <- read_csv("../../data/Aviram_raw_data/Aviram_PerKO_raw.csv") |> dplyr::rename(gene_name = ...1) 

raw_count <- full_join(raw_count_WT, raw_count_Per) |> replace_na_with(0) |> column_to_rownames("gene_name") 

raw_count <- raw_count[,metadata$title]

```

*requires second thought!! should i use full-join or inner-join?*
- Treating NA → 0 assumes it’s biological, which can bias downstream normalization.


### Normalisation

```{r}
# 1. Convert to CPM

# Compute CPM
raw_count_cpm <- raw_count |>
  dplyr::mutate(across(.cols = everything(), .fns = ~ . * 1000000 / sum(.), .names = "{col}"))

# Filter for at least CPM >= 1 in one sample
raw_count_cpm_filtered <- dplyr::filter(raw_count_cpm, apply(raw_count_cpm, 1, max) >= 1)
raw_count_cpm_filtered
```

```{r}
# 2. TMM normalisation

# Filter raw count based on cpm
raw_count_filtered <- raw_count[rownames(raw_count) %in% rownames(raw_count_cpm_filtered),]
nrow(raw_count_filtered)

# Normalise library size of filtered raw count
raw_count_filtered_norm <- DGEList(counts=raw_count_filtered) |> normLibSizes()
str(raw_count_filtered_norm)
raw_count_filtered_norm$samples

# log2 cpm with normalised library size
raw_filtered_norm_logcpm <- cpm(raw_count_filtered_norm, log=TRUE)
raw_filtered_norm_logcpm_prior <- cpm(raw_count_filtered_norm, log=TRUE, prior.count=2)

```


```{r}
cov <- sd(raw_count_filtered_norm$samples$lib.size) / mean(raw_count_filtered_norm$samples$lib.size) * 100
cov
```

Aviram's count matrix seems already filtered???

coefficient of variance = 22.2% => ok!!


```{r}
### RUV
x <- str_c(metadata$genotype, str_pad(metadata$time, 2, pad="0"), sep="_")

differences <- makeGroups(x)  # -1 assigned to missing sample (PerKO20 and PerKO8)
controls <- rownames(raw_filtered_norm_logcpm_prior)

# k=2, prior
RUVs_output_prior <- RUVs(raw_filtered_norm_logcpm_prior, controls, k=2, differences, isLog=TRUE)
raw_filtered_norm_logcpm_prior_RUVcount_k2 <- RUVs_output_prior$normalizedCounts
as.data.frame(raw_filtered_norm_logcpm_prior)
as.data.frame(raw_filtered_norm_logcpm_prior_RUVcount_k2)
W1 <- RUVs_output_prior$W[,1]
W2 <- RUVs_output_prior$W[,2]

```

```{r}
par(mfrow = c(2,2))

## filtered, unnormalised
plotRLE(as.matrix(raw_count_filtered), outline=FALSE, exprs_logged = FALSE,
        ylim=c(-0.9, 0.9), main = 'Unnormalised (filtered)', xlab = "Samples", ylab = "Relative log")

## filtered, TMM normalised
plotRLE(raw_filtered_norm_logcpm_prior, outline=FALSE, exprs_logged = TRUE,
        ylim=c(-0.9, 0.9), main = 'TMM normalised (filtered)', xlab = "Samples", ylab = "Relative log")

## filtered, TMM & RUV normalised
plotRLE(RUVs_output_prior$normalizedCounts, outline=FALSE, exprs_logged = TRUE,
        ylim=c(-0.9, 0.9), main = 'TMM + RUV normalised (filtered)', xlab = "Samples", ylab = "Relative log")

```

```{r}
# PCA plot: CPM filtered raw count
pca_test_raw <- prcomp(t(raw_count_filtered), center=TRUE, scale=TRUE)
pca_df_raw <- as.data.frame(pca_test_raw$x)
pca_df_raw <- left_join(rownames_to_column(pca_df_raw, "title"), metadata, by = "title")
pca_df_raw$lightdark <- "light"
pca_df_raw$lightdark[pca_df_raw$time %in% c("0", "16", "20", "24", "40", "44")] <- "dark"

ggplot(pca_df_raw, aes(PC1, PC2, color = lightdark, shape = genotype, label = time)) +
  geom_point(size = 2)  +
  geom_text(hjust=0, vjust=0, color = "black", size = 2.5) +
  theme_classic() +
  labs(title = "PCA: Aviram unnormalised, filtered",
       x = paste0("PC1 (", round(summary(pca_test_raw)$importance[2,1] * 100, 1), "%)"),
       y = paste0("PC2 (", round(summary(pca_test_raw)$importance[2,2] * 100, 1), "%)"))
```

```{r}
## PCA plot: TMM_log_cpm normalised

pca_test <- prcomp(t(raw_filtered_norm_logcpm_prior), center=TRUE, scale=TRUE)
pca_df_TMM <- as.data.frame(pca_test$x)
pca_df_TMM <- left_join(rownames_to_column(pca_df_TMM, "title"), metadata, by = "title")
pca_df_TMM$lightdark <- "light"
pca_df_TMM$lightdark[pca_df_TMM$time %in% c("0", "16", "20", "24", "40", "44")] <- "dark"


ggplot(pca_df_TMM, aes(PC1, PC2, color = lightdark, shape = genotype, label = time)) +
  geom_point(size = 2) +
  geom_text(hjust=0, vjust=0, color = "black", size = 2.5) +
  theme_classic() +
  labs(title = "PCA: Aviram TMM-normalised only",
       x = paste0("PC1 (", round(summary(pca_test)$importance[2,1] * 100, 1), "%)"),
       y = paste0("PC2 (", round(summary(pca_test)$importance[2,2] * 100, 1), "%)"))

```

```{r}
## PCA plot: RUV normalised, k=2, prior

pca_test <- prcomp(t(raw_filtered_norm_logcpm_prior_RUVcount_k2), center=TRUE, scale=TRUE)
pca_df_RUV <- as.data.frame(pca_test$x)
pca_df_RUV <- left_join(rownames_to_column(pca_df_RUV, "title"), metadata, by = "title")
pca_df_RUV$lightdark <- "'light'"
pca_df_RUV$lightdark[pca_df_RUV$time %in% c("0", "16", "20", "24", "40", "44")] <- "'dark'"

ggplot(pca_df_RUV, aes(PC1, PC2, color = lightdark, shape = genotype, label = time)) +
  geom_point(size = 2) +
  geom_text(hjust=0, vjust=0, color = "black", size = 2.5) +
  theme_classic() +
  labs(title = "PCA: Aviram TMM + RUV (k=2, prior)",
       x = paste0("PC1 (", round(summary(pca_test)$importance[2,1] * 100, 1), "%)"),
       y = paste0("PC2 (", round(summary(pca_test)$importance[2,2] * 100, 1), "%)"))


```

ehhh still cannot separate perDKO --> *actually it makes sense!! bcuz it's in DD and the mutant don't have functional clock anymore!!*



```{r}
# Limorhyde

period = 24
qvalRhyCutoff = 0.15
qvalDrCutoff = 0.1

### should i use unRUV matrix???? yes
y = raw_filtered_norm_logcpm_prior #_RUVcount_k2

metadata_lmr <- metadata |> dplyr::select(title, genotype, time) |> 
  dplyr::mutate(time = as.integer(time), cond = factor(genotype)) |> 
  rownames_to_column("sample_alias")


metadata_lmr <- cbind(metadata_lmr, limorhyde(metadata_lmr$time, 'time_'), W1, W2) |> as.data.table()


```


```{r}
condNow <- "WT"

design = model.matrix(~ time_cos + time_sin, data = metadata_lmr[cond == condNow, ])
fit = lmFit(y[, metadata_lmr$cond == condNow], design)
fit = eBayes(fit, trend = TRUE)
rhyNow = data.table(topTable(fit, coef = 2:3, number = Inf), keep.rownames = TRUE)
setnames(rhyNow, 'rn', 'gene_name')
rhyNow[, cond := condNow]

rhyLimmaSummary = rhyNow[, .(P.Value = min(P.Value)), by = gene_name]
rhyLimmaSummary[, adj.P.Val := p.adjust(P.Value, method = 'BH')]
setorderv(rhyLimmaSummary, 'adj.P.Val')

rhyLimmaSummary[rhyLimmaSummary$adj.P.Val < 0.05, ]
```

```{r}
# Extract coefficients from your limma fit
coefs = coef(fit)[, c("(Intercept)", "time_cos", "time_sin")]

# Compute (log2 *relative*) amplitude for each gene
amplitude_lmr = sqrt(coefs[, "time_cos"]^2 + coefs[, "time_sin"]^2)
mean = coefs[, "(Intercept)"]

# Compute (absolute) amplitude
amplitude_lmr_abs = 2^(mean + amplitude_lmr) - 2^mean

cbind(amplitude_lmr, coefs[, "(Intercept)"])

hist(amplitude_lmr, ylim = c(0,100))
```

```{r}
as.data.frame(raw_filtered_norm_logcpm_prior[amplitude_lmr > 1,])
amplitude_lmr[amplitude_lmr > 1]
sum(amplitude_lmr > 0.58)

```

### JTK
```{r}
library(dataPreparation)
metadata
table(metadata$time[metadata$genotype == "WT"])

# log-TMM & RUV normalised matrix, control only
RUVs_output_prior$normalizedCounts[,1:48]

JTK_matrix_RUV <- as.data.frame(RUVs_output_prior$normalizedCounts[,1:48])
# JTK_matrix_RUV[rowSums(JTK_matrix_RUV)>0,]
# JTK_matrix_RUV[raw_count_WT$gene_name, ]

JTK_matrix_RUV <- JTK_matrix_RUV |> rownames_to_column('Probe')

write.table(JTK_matrix_RUV, file = 'JTK_Cycle/Aviram_JTK/Aviram_JTK_RUV_data.txt', sep = '\t', row.names = FALSE, col.names = TRUE, quote=FALSE)

write.table(JTK_matrix_RUV['Probe'], file = 'JTK_Cycle/Aviram_JTK/Aviram_JTK_RUV_annot.txt', sep = '/t', row.names = FALSE, col.names = TRUE, quote=FALSE)



```
12377 => 11398 after filtering out the zeros in WT
original matrix: 11727

--- running JTK cycle ---

```{r}
JTK_output_RUV <- read.table(file = 'JTK_Cycle/Aviram_JTK/JTK.Aviram_JTK_RUV_run.txt', header = TRUE ) |> dplyr::rename('gene_name' = 'Probe')

# Lable as rythmic & non-rhythmic using limorhyde's amplitude

JTK_output_RUV$amp_lmr <- amplitude_lmr[JTK_output_RUV$gene_name]
JTK_output_RUV$amp_lmr_abs <- amplitude_lmr_abs[JTK_output_RUV$gene_name]
JTK_output_RUV <- dplyr::relocate(JTK_output_RUV, amp_lmr, amp_lmr_abs, .after = AMP)

JTK_output_RUV$Rhythmicity_lmrAMP <- 'Ambiguous'
JTK_output_RUV$Rhythmicity_lmrAMP[JTK_output_RUV$BH.Q > 0.1 & JTK_output_RUV$amp_lmr < 0.58] <- 'Non-rhythmic'
JTK_output_RUV$Rhythmicity_lmrAMP[JTK_output_RUV$BH.Q < 0.05 & JTK_output_RUV$amp_lmr > 0.58] <- 'Rhythmic'
JTK_output_RUV <- dplyr::relocate(JTK_output_RUV, Rhythmicity_lmrAMP, .before = BH.Q)
table(JTK_output_RUV$Rhythmicity_lmrAMP)


# nonsense threshold for absolute values...
# JTK_output_RUV$Rhythmicity_lmrAMP_abs <- 'Non-rhythmic'
# JTK_output_RUV$Rhythmicity_lmrAMP_abs[JTK_output_RUV$BH.Q < 0.05 & JTK_output_RUV$amp_lmr_abs > 0.2] <- 'Rhythmic'
# JTK_output_RUV <- dplyr::relocate(JTK_output_RUV, Rhythmicity_lmrAMP_abs, .before = BH.Q)
# table(JTK_output_RUV$Rhythmicity_lmrAMP_abs)


```

rAMP > 0.3:
Non-rhythmic     Rhythmic 
       10827         1550 
       
rAMP > 0.58:
Non-rhythmic     Rhythmic 
       12040          337

huh why so much more than Bignon's??? --> *probably due to longer experiment duration*
       

### Limorhyde x Limma DE analysis

```{r}
# with interaction

design = model.matrix(~ cond + time_cos + time_sin + cond:time_cos + cond:time_sin + W1 + W2, data = metadata_lmr)

qr(design)$rank  # check if W1/2 are colinear with cond
cor(metadata_lmr[, c("W1", "W2", "time_cos", "time_sin")])  # check if W1/2 is eating away time information


fit = lmFit(y, design)
fit = eBayes(fit, trend = TRUE)
deLimma = data.table(topTable(fit, coef = 2, number = Inf), keep.rownames = TRUE)
setnames(deLimma, 'rn', 'gene_name')

# record mean level of WT and KO!
mean_wt <- fit$coefficients[,1]
mean_ko <- fit$coefficients[,1] + fit$coefficients[,2]

# deLimma = deLimma[!(gene_id %in% drLimma[adj.P.Val <= qvalDrCutoff]$gene_id)] ## don't ignore differentially rhythmic genes
deLimma[, adj.P.Val := p.adjust(P.Value, method = 'BH')]
setorderv(deLimma, 'adj.P.Val')

deLimma

deLimma$diffExpressed <- 'NO'
deLimma$diffExpressed[deLimma$logFC>1 & deLimma$adj.P.Val<0.05] <- 'UP'
deLimma$diffExpressed[deLimma$logFC<(-1) & deLimma$adj.P.Val<0.05] <- 'DOWN'
table(deLimma$diffExpressed)
```
logFC > ±0.8:
 DOWN    NO    UP 
  582 11514   281 

logFC > ±1:
 DOWN    NO    UP 
  309 11946   122 

## Rhythmicity vs DE

```{r}
JTK_output_RUV

deLimma

JTK_deLimma <- left_join(deLimma, JTK_output_RUV, by = "gene_name") |> 
  dplyr::mutate(Rhythmicity_fac = factor(Rhythmicity_lmrAMP)) |>
  dplyr::mutate(diffExpressed_fac = factor(diffExpressed)) |> 
  dplyr::relocate(Rhythmicity_fac, .after = "gene_name") |> 
  dplyr::relocate(diffExpressed_fac, .after = "gene_name")

table(JTK_deLimma[, c('diffExpressed', 'Rhythmicity_lmrAMP')])
```

p>0.1, p<0.05            
             Rhythmicity_lmrAMP
diffExpressed Ambiguous Non-rhythmic Rhythmic
         DOWN        26          260       23
         NO        3755         7896      295
         UP          27           76       19


```{r}
table <- table(JTK_deLimma[, c('diffExpressed', 'Rhythmicity_lmrAMP')])

DEratio <- (table[1, ] + table[3,]) / colSums(table)
DEratio
```

DEratio:
(AMP > 0.3)
Non-rhythmic     Rhythmic 
  0.06686986   0.08967742 


(AMP > 0.5)
Non-rhythmic     Rhythmic 
   0.0663631    0.1467181 
   
 Ambiguous Non-rhythmic     Rhythmic 
  0.01391807   0.04081633   0.12462908

```{r}
volcano_data_rhythmic <- JTK_deLimma |> dplyr::filter(Rhythmicity_lmrAMP=='Rhythmic')
volcano_data_nonrhythmic <- JTK_deLimma |> dplyr::filter(Rhythmicity_lmrAMP=='Non-rhythmic')

ggplot(data=volcano_data_rhythmic, aes(x=logFC, y=-log10(adj.P.Val), col = diffExpressed_fac)) +
  geom_vline(xintercept = c(-0.8, 0.8), col='gray', linetype='dashed') +
  geom_hline(yintercept=c(-log10(0.05)), col='gray', linetype='dashed') +
  geom_point(size=1) +
  scale_color_manual(values = c("#00AFBB", "grey", "#FC4E07"), 
                     labels = c("Downregulated", "Not significant", "Upregulated")) +
  ggtitle("Rhythmic genes") +
  xlim(-5, 5) +
  ylim(-1, 50) +
  geom_label_repel(data = dplyr::filter(volcano_data_rhythmic, abs(logFC) > 1.8 | -log10(adj.P.Val) > 25), aes(, label = gene_name), hjust=0, vjust=0, color = "black", size = 2.5)
  

ggplot(data=volcano_data_nonrhythmic, aes(x=logFC, y=-log10(adj.P.Val
), col = diffExpressed_fac)) +
  geom_vline(xintercept = c(-0.8, 0.8), col='gray', linetype='dashed') +
  geom_hline(yintercept=c(-log10(0.05)), col='gray', linetype='dashed') +
  geom_point(size=1) +
  scale_color_manual(values = c("#00AFBB", "grey", "#FC4E07"), 
                     labels = c("Downregulated", "Not significant", "Upregulated")) +
  ggtitle("Non-rhythmic genes") +
  xlim(-5, 5) +
  ylim(-1, 50) +
  geom_label_repel(data = dplyr::filter(volcano_data_nonrhythmic, logFC < -2 | logFC > 1.2 | -log10(adj.P.Val) > 35), aes(, label = gene_name), hjust=0, vjust=0, color = "black", size = 2.5)

```

Significantly more down-regulation! 


### Annotation of rhythmic genes

```{r}
DE_rhythmic_genes <- JTK_deLimma[JTK_deLimma$BH.Q < 0.05 & 
                                   JTK_deLimma$amp_lmr > 0.58 & 
                                   abs(JTK_deLimma$logFC) > 1.0 & 
                                   JTK_deLimma$adj.P.Val < 0.05]

DE_rhythmic_genes_annot <- right_join(annotations[match(DE_rhythmic_genes$gene_name, annotations$gene_name),], DE_rhythmic_genes) |>
  arrange(desc(abs(logFC))) |> as.matrix()

write.table(DE_rhythmic_genes_annot, file = "../../results/Aviram_DE_rhythmic_genes_annot.tsv", sep = "\t", row.names = FALSE)

sum(annotations$gene_name == "Aim1l")


```



```{r}
JTK_deLimma
ggplot(JTK_deLimma, aes(x = amp_lmr, y = logFC)) +
  geom_point(alpha = 0.5) +
  theme_minimal() +
  xlab("Rhythmic amplitude (from Limorhyde)") +
  ylab("log2 Fold Change (DE)") +
  ggtitle("DE vs rhythmic amplitude")


cor(JTK_deLimma$logFC, JTK_deLimma$amp_lmr)
```

**significantly more down-skewed!


```{r}
JTK_deLimma_mean <- JTK_deLimma |>
  left_join(rownames_to_column(as.data.frame(mean_wt), "gene_name")) |>
  left_join(rownames_to_column(as.data.frame(mean_ko), "gene_name")) |>
  mutate(abs_diff = 2^mean_ko - 2^mean_wt) |> 
  dplyr::relocate(mean_wt, mean_ko, abs_diff, .after = gene_name)
mean_wt[geneSymbolsNow]
mean_ko[geneSymbolsNow]
```


```{r}
ggplot(JTK_deLimma[JTK_deLimma$Rhythmicity_fac == "Rhythmic", ], aes(x = amp_lmr, y = logFC, color = Rhythmicity_fac)) +
  geom_point(alpha = 0.6) +
  scale_color_manual(values = c(
    "Non-rhythmic" = "transparent",
    "Rhythmic" = "royalblue"
  )) +
  theme_minimal(base_size = 14) +
  xlab("Rhythmic absolute amplitude") +
  ylab("log2 Fold Change") +
  xlim(0, 3) +
  ggtitle("logFC vs rhythmic amplitude by rhythmicity") +
  theme(
    legend.title = element_text(face = "bold"),
    legend.position = "right"
  ) +
  geom_label_repel(data = dplyr::filter(JTK_deLimma, (amp_lmr > 1.3 | abs(logFC) > 2.5) & amp_lmr > 0.3), aes(, label = gene_name), 
            hjust=0, vjust=0, color = "black", size = 2)


ggplot(JTK_deLimma_mean, aes(x = amp_lmr_abs, y = abs_diff)) +
  geom_point(alpha = 0.6, color = "coral") +
  # scale_color_manual(values = c(
  #   "Non-rhythmic" = "transparent",
  #   "Rhythmic" = "royalblue"
  # )) +
  theme_minimal(base_size = 14) +
  xlab("Absolute amplitude") +
  ylab("Absolute difference") +
  ggtitle("Absolute difference vs absolute amplitude") +
  theme(
    legend.title = element_text(face = "bold"),
    legend.position = "right"
  ) +
  ylim(-5000,5000) +
  geom_label_repel(data = dplyr::filter(JTK_deLimma_mean, amp_lmr_abs > 500), aes(, label = gene_name), 
            hjust=0, vjust=0, color = "black", size = 3)
```






```{r}
### Plot genes (using limorhyde's version now, to be modifie)

metadata_dt <- as.data.table(metadata)


geneSymbolsNow = c("Per1", "Per2", "Per3", "Cry1", "Cry2")
geneIdsNow = annotations[match(geneSymbolsNow, annotations$gene_name),]$gene_id
  
df = data.table(t(y[geneSymbolsNow, ]))
df[, title := colnames(y[geneSymbolsNow, ])]

df = merge(df, metadata_dt[, .(title, genotype, time)], by = 'title')
df = melt(df, measure.vars = geneSymbolsNow, variable.name = 'symbol',
          value.name = 'expr')
df[, symbol := factor(symbol, geneSymbolsNow)]
df[, time := as.integer(time)]
df[, genotype := factor(genotype, levels = c("WT", "PerDKO"))]

ggplot(df) +
  facet_grid(symbol ~ genotype, scales = 'free_y') +
  geom_point(aes(x = time, y = expr, shape = genotype, color = symbol), size = 2) +
  labs(x = 'Zeitgeber time (h)', y = 'Expression (log)') +
  scale_shape(solid = FALSE, guide = 'none') +
  scale_color_brewer(type = 'qual', palette = 'Dark2', guide = 'none') +
  scale_x_continuous(limits = c(0, 24), breaks = seq(0, 24, 4))


## This shows the log2 value!, hence the AMP ~ relative AMP
```


```{r}
geneSymbolsNow = c("Arntl", "Clock", "Npas2", "Nr1d1","Dbp", "Actb")
geneIdsNow = annotations[match(geneSymbolsNow, annotations$gene_name),]$gene_id
  
df = data.table(t(y[geneSymbolsNow, ]))
df[, title := colnames(y[geneSymbolsNow, ])]

df = merge(df, metadata_dt[, .(title, genotype, time)], by = 'title')
df = melt(df, measure.vars = geneSymbolsNow, variable.name = 'symbol',
          value.name = 'expr')
df[, symbol := factor(symbol, geneSymbolsNow)]
df[, time := as.integer(time)]
df[, genotype := factor(genotype, levels = c("WT", "PerDKO"))]

ggplot(df) +
  facet_grid(symbol ~ genotype, scales = 'free_y') +
  geom_point(aes(x = time, y = expr, shape = genotype, color = symbol), size = 2) +
  labs(x = 'Zeitgeber time (h)', y = 'Expression (log)') +
  scale_shape(solid = FALSE, guide = 'none') +
  scale_color_brewer(type = 'qual', palette = 'Dark2', guide = 'none') +
  scale_x_continuous(limits = c(0, 24), breaks = seq(0, 24, 4))

```

There's no Kcnj15 in these samples!!





### 3D plot!

```{r}

AMP_threshold_list <- seq(0, 2, by=0.05)
logFC_threshold_list <- seq(0, 5, by=0.1)


threshold_result <- data.frame(
  AMP = NA,
  logFC = NA,
  ratio = NA,
  rhythmic_DE = NA,
  DE = NA,
  rhythmic = NA,
  log_rhythmic_DE = NA
)


# 3d threshold plot data
for (logFC_threshold in logFC_threshold_list){
  for (AMP_threshold in AMP_threshold_list){
    N_rhythmic_DE <- nrow(JTK_deLimma[abs(JTK_deLimma$logFC) > logFC_threshold & JTK_deLimma$adj.P.Val < 0.05 & JTK_deLimma$BH.Q < 0.05 & JTK_deLimma$amp_lmr > AMP_threshold])
    N_rhythmic <- nrow(JTK_deLimma[JTK_deLimma$BH.Q < 0.05 & JTK_deLimma$amp_lmr > AMP_threshold])
    # print(c(AMP_threshold, N_rhythmic))
    N_DE <- nrow(JTK_deLimma[abs(JTK_deLimma$logFC) > logFC_threshold & JTK_deLimma$adj.P.Val < 0.05])
    
    if (N_rhythmic == 0){
      DE_ratio = NA
    } 
    else{
      DE_ratio <- N_rhythmic_DE/N_rhythmic
    }
    threshold_result <- rbind(threshold_result, c(AMP = AMP_threshold, logFC = logFC_threshold, ratio = DE_ratio, rhythmic_DE = N_rhythmic_DE, DE = N_DE, rhythmic = N_rhythmic, log10_rhythmic_DE = log10(N_rhythmic_DE)))
  }
}

threshold_result <- drop_na(threshold_result)
threshold_result_finite <- threshold_result[is.finite(threshold_result$log_rhythmic_DE), ]


threshold_result[threshold_result$AMP == 0.55 & threshold_result$logFC == 1.0, ]
```

```{r}
### Relaative 3D plot

fig <- plot_ly(threshold_result, x = ~logFC, y = ~AMP, z = ~ratio, color = ~rhythmic_DE, size = I(30), colors = rainbow(10))
fig <- fig %>% add_markers()
fig <- fig %>% layout(title = list(text="(Relative) DE ratio for various logFC and relative AMP thresholds", y = 0.95, x = 0.5, xanchor = 'center', yanchor = "top"),
  scene = list(xaxis = list(title = 'log2 Fold Change',
                                                range = c(0, 3.5)),
                                   yaxis = list(title = 'Amplitude (relative)',
                                                range = c(0, 1.0)),
                                   zaxis = list(title = 'Ratio of DE genes : Rhythmic genes',
                                                range = c(0, 0.9))))

fig


```

```{r}
### Relative 3D plot (log-count coloured)

fig <- plot_ly(threshold_result_finite, x = ~logFC, y = ~AMP, z = ~ratio, color = ~log_rhythmic_DE, size = I(30), colors = rainbow(5))
fig <- fig %>% add_markers()
fig <- fig %>% layout(title = list(text="(Relative) DE ratio for various logFC and relative AMP thresholds", y = 0.95, x = 0.5, xanchor = 'center', yanchor = "top"),
  scene = list(xaxis = list(title = 'log2 Fold Change <br> Magnitude',
                                                range = c(0, 2.5)),
               yaxis = list(title = 'Amplitude (relative)',
                                                range = c(0, 1.0)),
               zaxis = list(title = 'Ratio of DE genes : Rhythmic genes',
                                                range = c(0, 0.9)),
               aspectmode = "cube"))

colorbar(fig, title = "# rhythmic DE genes", tickmode = "array", tickvals = c(0,1,2,3), ticktext = c(1, 10, 100, 1000))



```


```{r}
AMP_abs_threshold_list <- seq(0, 300, by=10)
difference_threshold_list <- seq(0, 1100, by=20)


threshold_result_abs <- data.frame(
  AMP_abs = NA,
  difference = NA,
  ratio = NA,
  rhythmic_DE = NA,
  DE = NA,
  rhythmic = NA
)


# 3d threshold plot data
for (difference_threshold in difference_threshold_list){
  for (AMP_abs_threshold in AMP_abs_threshold_list){
    N_rhythmic_DE <- nrow(JTK_deLimma_mean[abs(JTK_deLimma_mean$abs_diff) > difference_threshold & 
                                             JTK_deLimma_mean$adj.P.Val < 0.05 & 
                                             JTK_deLimma_mean$BH.Q < 0.05 & 
                                             JTK_deLimma_mean$amp_lmr_abs > AMP_abs_threshold])
    N_rhythmic <- nrow(JTK_deLimma_mean[JTK_deLimma_mean$BH.Q < 0.05 & 
                                          JTK_deLimma_mean$amp_lmr_abs > AMP_abs_threshold])
    # print(c(AMP_threshold, N_rhythmic))
    N_DE <- nrow(JTK_deLimma_mean[abs(JTK_deLimma_mean$abs_diff) > difference_threshold & 
                                    JTK_deLimma_mean$adj.P.Val < 0.05])
    
    if (N_rhythmic == 0){
      DE_ratio = NA
    } 
    else{
      DE_ratio <- N_rhythmic_DE/N_rhythmic
    }
    threshold_result_abs <- rbind(threshold_result_abs, c(AMP_abs = AMP_abs_threshold, difference = difference_threshold, ratio = DE_ratio, rhythmic_DE = N_rhythmic_DE, DE = N_DE, rhythmic = N_rhythmic))
  }
}

threshold_result_abs <- drop_na(threshold_result_abs)

```


```{r}

fig <- plot_ly(threshold_result_abs, x = ~difference, y = ~AMP_abs, z = ~ratio, color = ~difference, size = I(30))
fig <- fig %>% add_markers()
fig <- fig %>% layout(title = list(text="(Absolute) DE ratio for various differences and absolute AMP thresholds", y = 0.95, x = 0.5, xanchor = 'center', yanchor = "top"),
  scene = list(xaxis = list(title = 'Absolute difference',
                                                range = c(0, 1100)),
                                   yaxis = list(title = 'Amplitude (absolute)',
                                                range = c(0, 300)),
                                   zaxis = list(title = 'Ratio of DE genes : Rhythmic genes',
                                                range = c(0, 1))))

fig


```