---
title: "Bignon Pipeline Analysis"
output: html_document
date: "2026-01-20"
---

```{r}
mamba_folder <- "/Users/elysialuo/miniforge3"
env_name <- "circadian"
.libPaths(paste(mamba_folder, "envs", env_name, "lib/R/library", sep = "/"))

library(tidyverse)
library(dryR)
library(stringr)
library(ggplot2)
library(ggpubr)

library(MASS)
library(janitor)
library(naniar)
library(patchwork)
library(scales)
library(cluster)

library(RColorBrewer)
library(ggrepel)

library(GEOquery)
library(stringr)
library(tidyr)

library(AnnotationHub)
library(AnnotationDbi)
library(ensembldb)

library(limma)
library(edgeR)
library(RUVSeq)

library(limorhyde)
library(annotate)
library(data.table)
library(org.Mm.eg.db)
library(foreach)

library(VennDiagram)
library(grid)
library(plotly)
library('org.Mm.eg.db')

library("clusterProfiler")

```

```{r}
getwd()
```

```{r}
### Anotation set up
ah <- AnnotationHub()

# search for id of desired dataset
dbResult <- query(ah, c("EnsDb", "Mus musculus", "102")) 
dbResult[1]

# load mouse transcript dataset
MouseEnsDb <- ah[[dbResult$ah_id]]
MouseEnsDb

# output data of interest in form of data.frame
annotations <- genes(MouseEnsDb, return.type = "data.frame")
annotations
```




```{r}
# read in raw count matrix from Nextflow pipeline
raw_count_assay <- read_rds("../../data/Bignon_QC_raw_data/salmon.merged.gene_counts.rds")
raw_count_unfiltered <- assays(raw_count_assay)$counts
```

```{r}
# read in Bignon's processed data for comparison
Bignon_processed_data <- read_csv("../../data/Bignon_processed_data//Bignon_RNA-seq_processed.csv") |>
  column_to_rownames("Gene_ID") |>
  dplyr::select(-c(Gene_Name, description, Gene_type))

```

```{r}
# Read in meta data (to modify)
gse <- getGEO(filename = "../../Data/Bignon_raw_data/GSE216252_series_matrix.txt.gz", GSEMatrix = TRUE,
    getGPL = FALSE)

meta <- gse[[1]]
gsm <- gse[[2]]
label_df <- as.data.frame(t(rbind(gsm, meta))) |> separate_wider_delim(cols = meta, delim = ", ", names = c("sample", "genotype", "timepoint", "replicate")) |> separate_wider_delim(cols = replicate, delim = " ", names = c("rep", "replicate")) |> separate_wider_delim(cols = genotype, delim = " ", names = c("genotype", "gen"))

meta_dryseq <- label_df |>
  mutate(genotype_new = recode(genotype, control = "C", cKO = "KO")) |> 
  mutate(timepoint_new = str_remove(timepoint, "ZT")) |>
  dplyr::select(gsm, genotype_new, timepoint_new, replicate) |> column_to_rownames("gsm")


```


```{r}
### Metadata matrix creation

metadata_all <- read.table("../../data/Bignon_QC_raw_data/metadata_matrix.tsv", sep='\t', header = TRUE)

metadata <- metadata_all |> 
  dplyr::mutate(genotype = if_else(grepl("KO", sample_title, fixed = TRUE), true = 'KO', false = 'C')) |> # genotype 
  dplyr::mutate(time = str_extract(metadata_all$sample_title, 'ZT(\\d*)', group = 1)) |> # time 
  dplyr::mutate(replicate = str_extract(metadata_all$sample_title, 'rep (\\d*)', group = 1)) |> # replicate
  dplyr::select(experiment_accession, sample_alias, genotype, time, replicate) |>
  arrange(experiment_accession)

```



### Filter for protein-coding genes!
```{r}
raw_count <- raw_count_unfiltered |> rownames_to_column("gene_id") |> dplyr::filter(gene_id %in% annotations[annotations$gene_biotype == "protein_coding", ]$gene_id) |> column_to_rownames("gene_id")

```


### Normalisation

```{r}
# 1. Convert to CPM & filter

# Compute CPM
raw_count_cpm <- as.data.frame(raw_count) |>
  mutate(across(.cols = everything(), .fns = ~ . * 1000000 / sum(.), .names = "{col}"))

# Filter for at least CPM >= 1 in at least two samples
raw_count_cpm_filtered <- raw_count_cpm_filtered <- raw_count_cpm[rowSums(raw_count_cpm >= 1) >= 2, ]
raw_count_cpm_filtered

Bignon_processed_data_sorted <- Bignon_processed_data[order(rownames(Bignon_processed_data), decreasing=FALSE),]
sum(rownames(raw_count_cpm_filtered) %in% rownames(Bignon_processed_data_sorted))
nrow(raw_count_cpm_filtered)
nrow(Bignon_processed_data_sorted) # Bignon's number of genes detected
nrow(raw_count)


```




```{r}
# 2. TMM normalisation

# Filter raw count based on cpm
raw_count_filtered <- raw_count[rownames(raw_count) %in% rownames(raw_count_cpm_filtered),]
nrow(raw_count_filtered)

# Normalise library size of filtered raw count
raw_count_filtered_norm <- DGEList(counts=raw_count_filtered) |> normLibSizes()
str(raw_count_filtered_norm)
raw_count_filtered_norm$samples

# log2 cpm with normalised library size                     (**need to check this prior.count=3 again!!**)
raw_filtered_norm_logcpm <- cpm(raw_count_filtered_norm, log=TRUE)
raw_filtered_norm_logcpm_prior <- cpm(raw_count_filtered_norm, log=TRUE) #, prior.count=3)

```

```{r}
cov <- sd(raw_count_filtered_norm$samples$lib.size) / mean(raw_count_filtered_norm$samples$lib.size) * 100
cov
```

```{r}
### RUV

x <- str_c(metadata$genotype,metadata$time,sep="_")
differences <- makeGroups(x)
controls <- rownames(raw_filtered_norm_logcpm_prior)

# k=2, prior
RUVs_output_prior <- RUVs(raw_filtered_norm_logcpm_prior, controls, k=2, differences, isLog=TRUE)
raw_filtered_norm_logcpm_prior_RUVcount_k2 <- RUVs_output_prior$normalizedCounts
as.data.frame(raw_filtered_norm_logcpm_prior)
as.data.frame(raw_filtered_norm_logcpm_prior_RUVcount_k2)
W1 <- RUVs_output_prior$W[,1]
W2 <- RUVs_output_prior$W[,2]

```


```{r}
pdf("../../results/Plots/Bignon_RLE.pdf", width = 9, height = 2.5)

par(mfrow = c(1,3))

## filtered, unnormalised
plotRLE(as.matrix(raw_count_filtered), outline=FALSE, exprs_logged = FALSE,
        ylim=c(-0.9, 0.9), main = 'Unnormalised', xlab = "Samples", ylab = "Relative log", cex.main = 1.2)

## filtered, TMM normalised
plotRLE(raw_filtered_norm_logcpm_prior, outline=FALSE, exprs_logged = TRUE,
        ylim=c(-0.9, 0.9), main = 'TMM normalised', xlab = "Samples", ylab = "Relative log", cex.main = 1.2)

## filtered, TMM & RUV normalised
plotRLE(RUVs_output_prior$normalizedCounts, outline=FALSE, exprs_logged = TRUE,
        ylim=c(-0.9, 0.9), main = 'TMM & RUV normalised', xlab = "Samples", ylab = "Relative log", cex.main = 1.2)

dev.off()
```




```{r}
# PCA plot: CPM filtered raw count
pca_test_raw <- prcomp(t(raw_count_filtered), center=TRUE, scale=TRUE)
pca_df_raw <- as.data.frame(pca_test_raw$x)
pca_df_raw <- left_join(rownames_to_column(pca_df_raw, "experiment_accession"), metadata, by = "experiment_accession") |> dplyr::mutate(genotype = ifelse(genotype == "C", "WT", "Bmal1KO")) |> dplyr::mutate(genotype_fac = factor(genotype, levels = c("WT", "Bmal1KO")))
pca_df_raw$lightdark <- "light"
pca_df_raw$lightdark[pca_df_raw$time %in% c("0", "16", "20")] <- "dark"

ggplot(pca_df_raw, aes(PC1, PC2, color = interaction(lightdark, genotype_fac, sep=':'), label = time)) +
  geom_point(size = 2) +
  geom_text(hjust=-0.2, vjust=-0.2, color = "black", size = 2.5) +
  theme_bw() +
  theme(
    panel.grid = element_blank(),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.6),
    axis.text = element_text(colour = "black"),
    axis.title = element_text(colour = "black", size = 12),
    plot.title = element_text(colour = "black", size = 12, hjust = 0),
    legend.key = element_blank()
  ) + 
  labs(title = "",
       x = paste0("PC1 (", round(summary(pca_test_raw)$importance[2,1] * 100, 1), "%)"),
       y = paste0("PC2 (", round(summary(pca_test_raw)$importance[2,2] * 100, 1), "%)"),
       color = 'Light condition:Genotype') 

ggsave(
  filename = "../../results/Plots/Bignon_volcano_PCA_raw.svg",
  plot = get_last_plot(),
  width = 8, 
  height = 5, 
  units = "in"
)
```


```{r}

## PCA plot: TMM_log_cpm normalised

pca_test <- prcomp(t(raw_filtered_norm_logcpm_prior), center=TRUE, scale=TRUE)
pca_df_TMM <- as.data.frame(pca_test$x)
pca_df_TMM <- left_join(rownames_to_column(pca_df_TMM, "experiment_accession"), metadata, by = "experiment_accession") |> dplyr::mutate(genotype = ifelse(genotype == "C", "WT", "Bmal1KO")) |> dplyr::mutate(genotype_fac = factor(genotype, levels = c("WT", "Bmal1KO")))
pca_df_TMM$lightdark <- "light"
pca_df_TMM$lightdark[pca_df_raw$time %in% c("0", "16", "20")] <- "dark"

ggplot(pca_df_TMM, aes(PC1, PC2, color = interaction(lightdark, genotype_fac, sep=':'), label = time)) +
  geom_point(size = 2) +
  geom_text(hjust=-0.2, vjust=-0.2, color = "black", size = 2.5) +
  theme_bw() +
  theme(
    panel.grid = element_blank(),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.6),
    axis.text = element_text(colour = "black"),
    axis.title = element_text(colour = "black", size = 12),
    plot.title = element_text(colour = "black", size = 12, hjust = 0),
    legend.key = element_blank()
  ) + 
  labs(title = "",
       x = paste0("PC1 (", round(summary(pca_test_raw)$importance[2,1] * 100, 1), "%)"),
       y = paste0("PC2 (", round(summary(pca_test_raw)$importance[2,2] * 100, 1), "%)"),
       color = 'Light condition:Genotype') 

ggsave(
  filename = "../../results/Plots/Bignon_volcano_PCA_TMM.svg",
  plot = get_last_plot(),
  width = 8, 
  height = 5, 
  units = "in"
)
```



```{r}
## PCA plot: RUV normalised, k=2, prior

pca_test <- prcomp(t(raw_filtered_norm_logcpm_prior_RUVcount_k2), center=TRUE, scale=TRUE)
pca_df_RUV <- as.data.frame(pca_test$x)
pca_df_RUV <- left_join(rownames_to_column(pca_df_RUV, "experiment_accession"), metadata, by = "experiment_accession") |> dplyr::mutate(genotype = ifelse(genotype == "C", "WT", "Bmal1KO")) |> dplyr::mutate(genotype_fac = factor(genotype, levels = c("WT", "Bmal1KO")))
pca_df_RUV$lightdark <- "light"
pca_df_RUV$lightdark[pca_df_raw$time %in% c("0", "16", "20")] <- "dark"

ggplot(pca_df_RUV, aes(PC1, PC2, color = interaction(lightdark, genotype_fac, sep=':'), label = time)) +
  geom_point(size = 2) +
  geom_text(hjust=-0.2, vjust=-0.2, color = "black", size = 2.5) +
  theme_bw() +
  theme(
    panel.grid = element_blank(),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.6),
    axis.text = element_text(colour = "black"),
    axis.title = element_text(colour = "black", size = 12),
    plot.title = element_text(colour = "black", size = 12, hjust = 0),
    legend.key = element_blank()
  ) + 
  labs(title = "",
       x = paste0("PC1 (", round(summary(pca_test_raw)$importance[2,1] * 100, 1), "%)"),
       y = paste0("PC2 (", round(summary(pca_test_raw)$importance[2,2] * 100, 1), "%)"),
       color = 'Light condition:Genotype') 

ggsave(
  filename = "../../results/Plots/Bignon_volcano_PCA_RUV.svg",
  plot = get_last_plot(),
  width = 8, 
  height = 5, 
  units = "in"
)
```

### Annotation
```{r}
annotation_selection <- annotations |> 
    dplyr::select(gene_id, gene_name, description, entrezid) |>
    dplyr::filter(gene_id %in% rownames(raw_count_filtered))

length(annotation_selection$entrezid)
length(unique(annotation_selection$entrezid))
sum(is.na(annotation_selection$entrezid))

# 
# raw_count_filtered_annot <- as.data.frame(raw_count_filtered) |> 
#     rownames_to_column("gene_id") |>
#     left_join(annotation_selection, "gene_id") |>
#     relocate(c(gene_name, description, entrezid), .after=gene_id)


annotations[annotations$gene_biotype == "protein_coding", ]$gene_id

```



## Rhythmic analysis

### a. dryR

```{r}
# RUV dryR

count_data_drylm_RUV <- as.data.frame(RUVs_output_prior$normalizedCounts)

# Check for missing values or empty rows
table(rowSums(count_data_drylm_RUV) == 0) 
table(is.na(count_data_drylm_RUV))

group_drylm_RUV <- metadata$genotype

time_drylm_RUV <- as.numeric(metadata$time)

replicate_drylm_RUV <- c(1:5)
replicate_drylm_RUV <- rep(replicate_drylm, times=12)


drylm_result_RUV = drylm(count_data_drylm_RUV,group_drylm_RUV,time_drylm_RUV, period = 24)

# explore the results
drylm_result_RUV[["results"]]     # data frame summarizing results
drylm_result_RUV[["parameters"]]  # coefficients: phase, amplitude and mean for each group
drylm_result_RUV[["ncounts"]]     # normalized counts
drylm_result_RUV[["counts"]]      # raw counts (NULL)
drylm_result_RUV[["cook"]]        # cook's distance for outlier detection (NULL)
drylm_result_RUV[["BICW_rhythm"]] # BICW for each rhythmic model
drylm_result_RUV[["BICW_mean"]]   # BICW for each mean model

# generate a pdf with a global summary of all models
plot_models_rhythm(drylm_result_RUV, "./")

# plot a feature of interest
dry_plot(drylm_result_RUV, "ENSMUSG00000000001")

## add in ambiguous classification (group 0)
rhythm_result_raw <- drylm_result_RUV$parameters |> dplyr::select(chosen_model, relamp_C) |> rownames_to_column('gene_id')
rhythm_result <- mutate(rhythm_result_raw,chosen_model = replace(chosen_model, drylm_result_RUV$parameters[,'chosen_model_BICW']<0.65, 0))

table(column_to_rownames(rhythm_result, 'gene_id'))

# Group for result with model 0-5
rhythm_result_raw_grouped <- rhythm_result_raw |> dplyr::mutate(grouped_chosen_model = if_else(chosen_model %in% c(2,4,5), true='rhythmic', false='non-rhythmic'))
rhythm_result_grouped <- rhythm_result |> dplyr::mutate(grouped_chosen_model = if_else(chosen_model %in% c(2,4,5), true='rhythmic', false='non-rhythmic')) |> dplyr::mutate(grouped_chosen_model = if_else(chosen_model %in% c(0), true='ambiguous', false=grouped_chosen_model)) 
rhythm_result_grouped_AMPfiltered <- rhythm_result_raw_grouped |> dplyr::mutate(grouped_chosen_model = if_else(chosen_model %in% c(2,4,5) & relamp_C > 0.29, true='rhythmic', false='non-rhythmic'))

table(rhythm_result_grouped_AMPfiltered$grouped_chosen_model)
table(rhythm_result_grouped$grouped_chosen_model)
table(rhythm_result_raw_grouped$grouped_chosen_model)
```

non-rhythmic     rhythmic 
       13661          133 

   ambiguous non-rhythmic     rhythmic 
        3957         4826         5011 

non-rhythmic     rhythmic 
        6776         7018
-----
non-rhythmic     rhythmic 
       13293          296 

   ambiguous non-rhythmic     rhythmic 
        3900         4702         4987 

non-rhythmic     rhythmic 
        6618         6971 


### b. Limorhyde

```{r}
# Limorhyde

period = 24
qvalRhyCutoff = 0.15

### should i use unRUV matrix???? yes
y = raw_filtered_norm_logcpm_prior #_RUVcount_k2

metadata_lmr <- metadata |> dplyr::select(experiment_accession, genotype, time) |> 
  dplyr::mutate(time = as.integer(time), cond = factor(genotype)) |>
  dplyr::mutate(sample_title = metadata_all$sample_title) |> 
  dplyr::relocate(sample_title, .after=experiment_accession)


metadata_lmr = cbind(metadata_lmr, limorhyde(metadata_lmr$time, 'time_'), W1, W2)
class(metadata_lmr)
setDT(metadata_lmr)

```


```{r}
condNow <- "C"

design = model.matrix(~ time_cos + time_sin, data = metadata_lmr[cond == condNow])
fit = lmFit(y[, metadata_lmr$cond == condNow], design)
fit = eBayes(fit, trend = TRUE)
rhyNow = data.table(topTable(fit, coef = 2:3, number = Inf), keep.rownames = TRUE)
setnames(rhyNow, 'rn', 'gene_id')
rhyNow[, cond := condNow]

rhyLimmaSummary = rhyNow[, .(P.Value = min(P.Value)), by = gene_id]
rhyLimmaSummary[, adj.P.Val := p.adjust(P.Value, method = 'BH')]
setorderv(rhyLimmaSummary, 'adj.P.Val')

rhyLimmaSummary[rhyLimmaSummary$adj.P.Val < 0.05, ]

```

y=RUV_count 
gene_id
<chr>
P.Value
<dbl>
adj.P.Val
<dbl>
ENSMUSG00000028150	1.591005e-18	1.844721e-14		
ENSMUSG00000042487	2.354462e-18	1.844721e-14		
ENSMUSG00000020889	5.454826e-18	2.849238e-14		
ENSMUSG00000055116	1.824733e-17	6.015901e-14		
ENSMUSG00000059824	1.919560e-17	6.015901e-14		
ENSMUSG00000021775	1.567585e-16	3.792929e-13		
ENSMUSG00000026077	1.694353e-16	3.792929e-13		
ENSMUSG00000037003	3.961385e-16	7.759363e-13		
ENSMUSG00000017428	5.295154e-16	9.219451e-13		
ENSMUSG00000029009	1.041123e-15	1.631440e-12	


y=pre-RUV
gene_id
<chr>
P.Value
<dbl>
adj.P.Val
<dbl>
ENSMUSG00000028150	1.915487e-18	3.001568e-14		
ENSMUSG00000055116	1.668041e-17	1.306910e-13		
ENSMUSG00000059824	4.549391e-17	1.903846e-13		
ENSMUSG00000021775	4.916462e-17	1.903846e-13		
ENSMUSG00000042487	6.820904e-17	1.903846e-13		
ENSMUSG00000020889	7.289773e-17	1.903846e-13		
ENSMUSG00000026077	2.391758e-16	5.354120e-13		
ENSMUSG00000028957	1.621035e-15	3.175202e-12		
ENSMUSG00000038550	2.093560e-15	3.645121e-12		
ENSMUSG00000027111	6.262701e-15	9.813653e-12	



min p vs joint 2-df F-test => no use lol

JTK more stringent than dryR & limorhyde


#### Proper AMP calculation for JTK result thresholding
```{r}

# Extract coefficients from your limma fit
coefs = coef(fit)[, c("(Intercept)", "time_cos", "time_sin")]


mean = coefs[, "(Intercept)"]

# Compute (absolute) amplitude
amplitude_lmr_abs = 2^(mean + amplitude_lmr) - 2^mean

# cbind(amplitude_lmr, amplitude_lmr_rel, coefs[, "(Intercept)"])
str(coef(fit))

hist(amplitude_lmr, ylim = c(0,100))


rhyLimmaSummary <- rhyLimmaSummary |> left_join(rownames_to_column(as.data.frame(amplitude_lmr), "gene_id"))
lmr_rhythmic <- rhyLimmaSummary[rhyLimmaSummary$adj.P.Val < 0.05 & rhyLimmaSummary$amplitude_lmr > 0.29, ]$gene_id 
```

I do have some genes with rAMP > 1: (my A calcutaed is the peak-mean ratio, hence possible) very high peak to trough ratio!?

### c. JTK (WT)

```{r}
# log-TMM & RUV normalised matrix, control only
RUVs_output_prior$normalizedCounts[,1:30]

JTK_matrix_RUV <- as.data.frame(RUVs_output_prior$normalizedCounts[,1:30])
JTK_matrix_RUV <- rownames_to_column(JTK_matrix_RUV, 'Probe')


write.table(JTK_matrix_RUV, file = 'JTK_Cycle/Bignon_QC_JTK_RUV_data.txt', sep = '\t', row.names = FALSE, col.names = TRUE, quote=FALSE)

write.table(JTK_matrix_RUV['Probe'], file = 'JTK_Cycle/Bignon_QC_JTK_RUV_annot.txt', sep = '/t', row.names = FALSE, col.names = TRUE, quote=FALSE)

```

--- running JTK cycle ---

```{r}
JTK_output_RUV <- read.table(file = 'JTK_Cycle/Bignon_JTK/JTK.Bignon_QC_JTK_RUV_run.txt', header = TRUE ) |> dplyr::rename('gene_id' = 'Probe')

# Lable as rythmic & non-rhythmic using limorhyde's amplitude

JTK_output_RUV$amp_lmr <- amplitude_lmr[JTK_output_RUV$gene_id]
JTK_output_RUV$amp_lmr_abs <- amplitude_lmr_abs[JTK_output_RUV$gene_id]
JTK_output_RUV <- dplyr::relocate(JTK_output_RUV, amp_lmr, amp_lmr_abs, .after = AMP)


JTK_output_RUV$Rhythmicity_lmrAMP <- 'Ambiguous'
JTK_output_RUV$Rhythmicity_lmrAMP[JTK_output_RUV$BH.Q > 0.1 | JTK_output_RUV$amp_lmr < 0.29] <- 'Non-rhythmic'
JTK_output_RUV$Rhythmicity_lmrAMP[JTK_output_RUV$BH.Q < 0.05 & JTK_output_RUV$amp_lmr > 0.29] <- 'Rhythmic'
JTK_output_RUV <- dplyr::relocate(JTK_output_RUV, Rhythmicity_lmrAMP, .before = BH.Q)
table(JTK_output_RUV$Rhythmicity_lmrAMP)


# AMP unfiltered
JTK_output_RUV$Rhythmicity <- 'Non-rhythmic'
JTK_output_RUV$Rhythmicity[JTK_output_RUV$BH.Q < 0.05] <- 'Rhythmic'
JTK_output_RUV <- dplyr::relocate(JTK_output_RUV, Rhythmicity, .before = BH.Q)
table(JTK_output_RUV$Rhythmicity)

```
JTK AMP (0.3):
Non-rhythmic     Rhythmic 
       15030          640 

Limorhyde AMP (0.3): 
Non-rhythmic     Rhythmic 
       14853          817 
       
Limorhyde relative AMP (0.2):     
Non-rhythmic     Rhythmic 
       15126          544 

(abs vs rel: 362 in common)

There seems to be more 'rhythmic genes' with amplitude calculated from amp_lmr


limorhyde 0.05 (vs 0.1) & 0.58 filter:
   Ambiguous Non-rhythmic     Rhythmic 
        5127        10372          171
        
protein-coding filter:  (:X)
   Ambiguous Non-rhythmic     Rhythmic 
        6829         8717          124 

-----
   Ambiguous Non-rhythmic     Rhythmic 
          82        12776          731


```{r}
## Visualise overlap (my QC vs Bignon QC)
JTK_output_bignon <- read.table(file = 'JTK_Cycle/Bignon_JTK/JTK.Bignon_JTK_RUV_run.txt', header = TRUE ) |> dplyr::rename('gene_id' = 'Probe')
JTK_output_bignon$Rhythmicity <- 'Non-rhythmic'
JTK_output_bignon$Rhythmicity[JTK_output_bignon$BH.Q < 0.05 & JTK_output_bignon$AMP > 0.3] <- 'Rhythmic'
JTK_output_bignon <- dplyr::relocate(JTK_output_bignon, Rhythmicity, .before = BH.Q)
table(JTK_output_bignon$Rhythmicity)

JTK_RUV_rhythmic_list_myQC <- JTK_output_RUV$gene_id[JTK_output_RUV$Rhythmicity == 'Rhythmic'] 
JTK_RUV_rhythmic_list_bignon <- JTK_output_bignon$gene_id[JTK_output_bignon$Rhythmicity == 'Rhythmic'] 

length(JTK_RUV_rhythmic_list_myQC)
length(JTK_RUV_rhythmic_list_bignon)

vlist <- list(JTK_RUV_rhythmic_list_myQC, JTK_RUV_rhythmic_list_bignon)
names(vlist) <- c("JTK_RUV_myQC", "JTK_RUV_bignon")

venn.diagram(
  x = vlist,
  category.names = c("JTK_RUV_myQC", "JTK_RUV_bignon"),
  filename = "VennDiagram/JTK_RUV_myQC_vs_RUV_bignon_AMP_lmr.png",
  output = TRUE,
  main = "JTK my QC vs Bignon's QC", # Title
  fill = c("lightyellow", "lightblue"), # Circle fill colors
  cat.col = "black", # Category label colors
  cex = 1.2 # Size of text
)

```

AMP > 0.3:
(Bignon's processed count)
Non-rhythmic     Rhythmic 
       14906          640 

(my own QC)
Non-rhythmic     Rhythmic 
       15030          640


```{r}
## Visualise overlap (my QC VS Bignon no RUV QC)
JTK_output_bignon <- read.table(file = 'JTK_Cycle/Bignon_JTK/JTK.Bignon_JTK_run.txt', header = TRUE ) |> dplyr::rename('gene_id' = 'Probe')
JTK_output_bignon$Rhythmicity <- 'Non-rhythmic'
JTK_output_bignon$Rhythmicity[JTK_output_bignon$BH.Q < 0.05 & JTK_output_bignon$AMP > 0.3] <- 'Rhythmic'
JTK_output_bignon <- dplyr::relocate(JTK_output_bignon, Rhythmicity, .before = BH.Q)
table(JTK_output_bignon$Rhythmicity)

JTK_RUV_rhythmic_list_myQC <- JTK_output_RUV$gene_id[JTK_output_RUV$Rhythmicity == 'Rhythmic'] 
JTK_RUV_rhythmic_list_bignon <- JTK_output_bignon$gene_id[JTK_output_bignon$Rhythmicity == 'Rhythmic'] 

length(JTK_RUV_rhythmic_list_myQC)
length(JTK_RUV_rhythmic_list_bignon)

vlist <- list(JTK_RUV_rhythmic_list_myQC, JTK_RUV_rhythmic_list_bignon)
names(vlist) <- c("JTK_RUV_myQC", "JTK_noRUV_bignon")

venn.diagram(
  x = vlist,
  category.names = c("JTK_RUV_myQC", "JTK_RUV_bignon"),
  filename = "VennDiagram/JTK_RUV_myQC_vs_noRUV_bignon.png",
  output = TRUE,
  main = "JTK my QC vs Bignon's QC", # Title
  fill = c("lightyellow", "lightblue"), # Circle fill colors
  cat.col = "black", # Category label colors
  cex = 1.2 # Size of text
)
```



#### Venn diagram comparison between dryR & JTK (filtered or unfiltered)
```{r}
## a. rhythmic (ambiguous & AMP filtered)
dryR_rhythmic_list <- rhythm_result_grouped_AMPfiltered$gene_id[rhythm_result_grouped_AMPfiltered$grouped_chosen_model == 'rhythmic']
JTK_rhythmic_list <- JTK_output_RUV$gene_id[JTK_output_RUV$Rhythmicity_lmrAMP == 'Rhythmic'] 


## b. rhythmic (unfiltered)
dryR_rhythmic_list <- rhythm_result_raw_grouped$gene_id[rhythm_result_raw_grouped$grouped_chosen_model == 'rhythmic']
JTK_rhythmic_list <- JTK_output_RUV$gene_id[JTK_output_RUV$Rhythmicity == 'Rhythmic'] 

length(dryR_rhythmic_list) 
length(JTK_rhythmic_list)


vlist <- list(dryR_rhythmic_list, JTK_rhythmic_list)
names(vlist) <- c("dryR", "JTK_CYCLE")
myCol <- brewer.pal(3, "Pastel2")

venn.plot <- venn.diagram(
  x = vlist,
  filename = NULL,
  output = FALSE,
  #main = "dryR vs JTK_Cycle (Rhythmic genes, ambiguous unfiltered)", # Title

  # Output features
  imagetype="svg" ,
  height = 480 , 
  width = 480 , 
  resolution = 300,
  compression = "lzw",
  
  # Circles
  lwd = 2,
  lty = 'blank',
  fill = myCol[1:2],
  
  # Numbers
  cex = 1,
  fontface = "bold",
  fontfamily = "sans",
  
  # Set names
  cat.cex = 1,
  cat.fontface = "bold",
  cat.default.pos = "outer",
  cat.pos = c(-30, -10),
  cat.dist = c(0.03, 0.03),
  cat.fontfamily = "sans",
  cat.col = myCol[1:2],
  rotation.degree = 1,
  
  # The --3 label
  # scaled = FALSE,
  ext.pos = c(180.5, 180.5),
  ext.dist = c(-0.17, 0.015),
  ext.length = c(0.7, 0.7),
  ext.lwd = 2
)


pdf("VennDiagram/dryR_vs_JTK_rhythmic_filtered.pdf", width = 4, height = 4)
grid::grid.newpage()
grid::grid.draw(venn.plot)
dev.off()
```



#### Venn diagram comparison between JTK (filtered) & limorhyde (filtered)
```{r}
## rhythmic (filtered)
JTK_rhythmic_list <- JTK_output_RUV$gene_id[JTK_output_RUV$Rhythmicity_lmrAMP == 'Rhythmic'] 
lmr_rhythmic_list <- lmr_rhythmic

length(JTK_rhythmic_list) 
length(lmr_rhythmic_list)


vlist <- list(lmr_rhythmic_list, JTK_rhythmic_list)
names(vlist) <- c("LimoRhyde", "JTK_CYCLE")
myCol <- brewer.pal(3, "Pastel2")

venn.plot <- venn.diagram(
  x = vlist,
  filename = NULL,
  output = FALSE,
  #main = "Limorhyde vs JTK_Cycle (Rhythmic genes, ambiguous unfiltered)", # Title

  # Output features
  imagetype="svg" ,
  height = 480 , 
  width = 480 , 
  resolution = 300,
  compression = "lzw",
  
  # Circles
  lwd = 2,
  lty = 'blank',
  fill = myCol[1:2],
  
  # Numbers
  cex = 1,
  fontface = "bold",
  fontfamily = "sans",
  
  # Set names
  cat.cex = 1,
  cat.fontface = "bold",
  cat.default.pos = "outer",
  cat.pos = c(-30, 30),
  cat.dist = c(0.03, 0.03),
  cat.fontfamily = "sans",
  cat.col = myCol[1:2],
  rotation.degree = 1,
  
  # The --3 label
  # scaled = FALSE,
  ext.pos = c(180.5, 180.5),
  ext.dist = c(-0.13, -0.13),
  ext.length = c(0.7, 0.7),
  ext.lwd = 2
)


pdf("VennDiagram/lmr_vs_JTK_rhythmic_filtered.pdf", width = 4, height = 4)
grid::grid.newpage()
grid::grid.draw(venn.plot)
dev.off()
```

#### Venn diagram comparison between dryR (ambiguous filtered) & JTK (amp filtered) & limorhyde (amp filtered)
```{r}

## rhythmic (ambiguous & AMP filtered)
dryR_rhythmic_list <- rhythm_result_grouped_AMPfiltered$gene_id[rhythm_result_grouped_AMPfiltered$grouped_chosen_model == 'rhythmic']
#rhythm_result_raw_grouped$gene_id[rhythm_result_raw_grouped$grouped_chosen_model == 'rhythmic']
JTK_rhythmic_list <- JTK_output_RUV$gene_id[JTK_output_RUV$Rhythmicity == 'Rhythmic'] 
lmr_rhythmic_list <- lmr_rhythmic

length(dryR_rhythmic_list) 
length(JTK_rhythmic_list)


vlist <- list(dryR_rhythmic_list, JTK_rhythmic_list, lmr_rhythmic_list)
names(vlist) <- c("DryR", "JTK_Cycle", "Limorhyde")
myCol <- brewer.pal(3, "Pastel2")

venn.plot <- venn.diagram(
  x = vlist,
  filename = NULL,
  output = FALSE,
  #main = "dryR vs JTK_Cycle vs Limorhyde (Rhythmic genes, ambiguous unfiltered)", # Title

  # Output features
  imagetype="svg" ,
  height = 480 , 
  width = 480 , 
  resolution = 300,
  compression = "lzw",
  
  # Circles
  lwd = 2,
  lty = 'blank',
  fill = myCol,
  
  # Numbers
  cex = 1,
  fontface = "bold",
  fontfamily = "sans",
  
  # Set names
  cat.cex = 1,
  cat.fontface = "bold",
  cat.default.pos = "outer",
  cat.pos = c(-30, -10, -20),
  cat.dist = c(0.03, 0.03, 0.03),
  cat.fontfamily = "sans",
  cat.col = myCol,
  rotation.degree = 1,
  
  # The --3 label
  # scaled = FALSE,
  ext.pos = c(180.5, 180.5, 0),
  ext.dist = c(-0.17, 0.015, 0),
  ext.length = c(0.7, 0.7,0),
  ext.lwd = 2
)


pdf("VennDiagram/dryR_vs_JTK_lmr_rhythmic_filtered.pdf", width = 4, height = 4)
grid::grid.newpage()
grid::grid.draw(venn.plot)
dev.off()
```

```{r}
library(eulerr)
library(RColorBrewer)

## your lists
vlist <- list(
  Limorhyde = lmr_rhythmic_list,
  JTK_Cycle = JTK_rhythmic_list,
  DryR = dryR_rhythmic_list
)

fit <- euler(vlist)

cols <- brewer.pal(3, "Set2")

pdf("VennDiagram/dryR_vs_JTK_rhythmic_unfiltered_scaled.pdf", width = 5.2, height = 4.6)

plot(
  fit,
  fills = list(fill = myCol, alpha = 0.45),
  edges = list(col = NA),          # remove dark outlines
  labels = FALSE,                  # add labels manually
  quantities = list(
    cex = 1.25,
    font = 2,
    col = "black"
  )
)

# Manual set labels to mimic your previous plot style
text(-1.9,  1.55, "JTK_Cycle",  col = myCol[1], cex = 1.2, font = 2, family = "sans")
text(-0.75, 1.80, "DryR",       col = myCol[2], cex = 1.2, font = 2, family = "sans")
text( 0.65, 1.72, "Limorhyde",  col = myCol[3], cex = 1.2, font = 2, family = "sans")

dev.off()
```



### JTK (mutant)

```{r}
# log-TMM & RUV normalised matrix, control only
RUVs_output_prior$normalizedCounts[,31:60]

JTK_matrix_RUV_mut <- as.data.frame(RUVs_output_prior$normalizedCounts[,31:60]) |>
  rownames_to_column('Probe')


write.table(JTK_matrix_RUV_mut, file = 'JTK_Cycle/Bignon_JTK/Bignon_QC_JTK_RUV_mut_data.txt', sep = '\t', row.names = FALSE, col.names = TRUE, quote=FALSE)

write.table(JTK_matrix_RUV_mut['Probe'], file = 'JTK_Cycle/Bignon_JTK/Bignon_QC_JTK_RUV_mut_annot.txt', sep = '/t', row.names = FALSE, col.names = TRUE, quote=FALSE)

str(RUVs_output_prior$normalizedCounts)
metadata
```

--- running JTK cycle ---

```{r}
JTK_output_RUV_mut <- read.table(file = 'JTK_Cycle/Bignon_JTK/JTK.Bignon_QC_JTK_RUV_mut_run.txt', header = TRUE ) |> dplyr::rename('gene_id' = 'Probe')

# Lable as rythmic & non-rhythmic using limorhyde's amplitude

JTK_output_RUV_mut$amp_lmr <- amplitude_lmr[JTK_output_RUV_mut$gene_id]
JTK_output_RUV_mut$amp_lmr_abs <- amplitude_lmr_abs[JTK_output_RUV_mut$gene_id]
JTK_output_RUV_mut <- dplyr::relocate(JTK_output_RUV_mut, amp_lmr, amp_lmr_abs, .after = AMP)

JTK_output_RUV_mut$Rhythmicity_lmrAMP <- 'Ambiguous'
JTK_output_RUV_mut$Rhythmicity_lmrAMP[JTK_output_RUV_mut$BH.Q > 0.1 | JTK_output_RUV_mut$amp_lmr < 0.29] <- 'Non-rhythmic'
JTK_output_RUV_mut$Rhythmicity_lmrAMP[JTK_output_RUV_mut$BH.Q < 0.05 & JTK_output_RUV_mut$amp_lmr > 0.29] <- 'Rhythmic'
JTK_output_RUV_mut <- dplyr::relocate(JTK_output_RUV_mut, Rhythmicity_lmrAMP, .before = BH.Q)
table(JTK_output_RUV_mut$Rhythmicity_lmrAMP)


```
   Ambiguous Non-rhythmic     Rhythmic 
          56        13104          429 

```{r}
## Visualise overlap (WT vs mutant rhythmic genes)

JTK_RUV_rhythmic_list_wt <- JTK_output_RUV$gene_id[JTK_output_RUV$Rhythmicity_lmrAMP == 'Rhythmic'] 
JTK_RUV_rhythmic_list_mut <- JTK_output_RUV_mut$gene_id[JTK_output_RUV_mut$Rhythmicity_lmrAMP == 'Rhythmic'] 

length(JTK_RUV_rhythmic_list_wt)
length(JTK_RUV_rhythmic_list_mut)

vlist <- list(JTK_RUV_rhythmic_list_wt, JTK_RUV_rhythmic_list_mut)
names(vlist) <- c("JTK_RUV_wt", "JTK_RUV_mut")
myCol <- brewer.pal(3, "Pastel2")

venn.plot <- venn.diagram(
  x = vlist,
  category.names = c("WT", "Bmal1 KO"),
  filename = NULL,
  output = FALSE,
  # main = "Bignon: WT vs Mutant rhythmic genes", # Title
  
  # Output features
  imagetype="svg" ,
  height = 480 , 
  width = 480 , 
  resolution = 300,
  compression = "lzw",
  
  # Circles
  lwd = 2,
  lty = 'blank',
  fill = myCol[1:2],
  
  # Numbers
  cex = 1.5,
  fontface = "bold",
  fontfamily = "sans",
  
  # Set names
  cat.cex = 1.5,
  cat.fontface = "bold",
  cat.default.pos = "outer",
  cat.pos = c(-30, 25),
  cat.dist = c(0.03, 0.035),
  cat.fontfamily = "sans",
  cat.col = myCol[1:2],
  rotation.degree = 1,
  
  # The --3 label
  # scaled = FALSE,
  ext.pos = 181.5,
  ext.dist = -0.035,
  ext.length = 0.84,
  ext.lwd = 2
)


pdf("Bignon_JTK_RUV_WTvsMut.pdf", width = 4, height = 4)
grid::grid.newpage()
grid::grid.draw(venn.plot)
dev.off()
```

I would say it's systemic effect


### Limorhyde x Limma DE analysis

```{r}
# with interaction

design = model.matrix(~ cond + time_cos + time_sin + cond:time_cos + cond:time_sin + W1 + W2, data = metadata_lmr)

qr(design)$rank  # check if W1/2 are colinear with cond
cor(metadata_lmr[, c("W1", "W2", "time_cos", "time_sin")])  # check if W1/2 is eating away time information


fit = lmFit(y, design)
fit = eBayes(fit, trend = TRUE)
plotSA <- plotSA(fit)
deLimma = data.table(topTable(fit, coef = 2, number = Inf), keep.rownames = TRUE)
setnames(deLimma, 'rn', 'gene_id')

# record mean level of WT and KO!
mean_wt <- fit$coefficients[,1]
mean_ko <- fit$coefficients[,1] + fit$coefficients[,2]

# deLimma = deLimma[!(gene_id %in% drLimma[adj.P.Val <= qvalDrCutoff]$gene_id)] ## don't ignore differentially rhythmic genes
deLimma[, adj.P.Val := p.adjust(P.Value, method = 'BH')]
setorderv(deLimma, 'adj.P.Val')

deLimma

deLimma$diffExpressed <- 'NO'
deLimma$diffExpressed[deLimma$logFC>0.58 & deLimma$adj.P.Val<0.05] <- 'UP'
deLimma$diffExpressed[deLimma$logFC<(-0.58) & deLimma$adj.P.Val<0.05] <- 'DOWN'
table(deLimma$diffExpressed)

pdf("VennDiagram/Bignon_plotSA.pdf", width = 6, height = 4)
plotSA(fit)
dev.off()
```

### Quality checking for this many df
[1] 3
                  W1          W2      time_cos      time_sin
W1        1.00000000  0.03083128  3.350203e-02 -1.714264e-02
W2        0.03083128  1.00000000 -3.540168e-02  3.013925e-02
time_cos  0.03350203 -0.03540168  1.000000e+00  5.835313e-17
time_sin -0.01714264  0.03013925  5.835313e-17  1.000000e+00


 DOWN    NO    UP 
  201 15317   152 
---
 DOWN    NO    UP 
  259 13094   236

```{r}
# without interaction

design = model.matrix(~ cond + time_cos + time_sin + W1 + W2, data = metadata_lmr)

fit_noint = lmFit(y, design)
fit_noint = eBayes(fit_noint, trend = TRUE)
deLimma_noint = data.table(topTable(fit_noint, coef = 2, number = Inf), keep.rownames = TRUE)
setnames(deLimma_noint, 'rn', 'gene_id')

# deLimma = deLimma[!(gene_id %in% drLimma[adj.P.Val <= qvalDrCutoff]$gene_id)] ## don't ignore differentially rhythmic genes
deLimma_noint[, adj.P.Val := p.adjust(P.Value, method = 'BH')]
setorderv(deLimma_noint, 'adj.P.Val')

deLimma_noint[1:5, ]

deLimma_noint$diffExpressed <- 'NO'
deLimma_noint$diffExpressed[deLimma_noint$logFC>0.8 & deLimma_noint$adj.P.Val<0.05] <- 'UP'
deLimma_noint$diffExpressed[deLimma_noint$logFC<(-0.8) & deLimma_noint$adj.P.Val<0.05] <- 'DOWN'
table(deLimma_noint$diffExpressed)

```


```{r}
deLimma$gene_id[deLimma$diffExpressed != "NO"] %in% deLimma_noint$gene_id[deLimma_noint$diffExpressed != "NO"]


```

why is there not much of a difference???????? does it suggest differential rhythmicity doesn't affect much of the DE here



## Rhythmicity vs DE

```{r}
JTK_output_RUV

deLimma

JTK_deLimma <- left_join(deLimma, JTK_output_RUV, by = "gene_id") |> 
  dplyr::mutate(Rhythmicity_fac = factor(Rhythmicity_lmrAMP)) |>
  # dplyr::mutate(Rhythmicity_abs_fac = factor(Rhythmicity_lmrAMP_abs)) |>
  dplyr::mutate(diffExpressed_fac = factor(diffExpressed)) |> 
  dplyr::relocate(Rhythmicity_fac, .after = "gene_id") |> 
  dplyr::relocate(diffExpressed_fac, .after = "gene_id") |> 
  left_join(annotation_selection[, 1:2], by = "gene_id") |> 
  dplyr::relocate(gene_name, .after = gene_id)

table(JTK_deLimma[, c('diffExpressed', 'Rhythmicity_lmrAMP')])
# table(JTK_deLimma[, c('diffExpressed', 'Rhythmicity_lmrAMP_abs')])
```
             Rhythmicity
diffExpressed Non-rhythmic Rhythmic
         DOWN          146       55
         NO          14795      522
         UP             89       63
         
         Rhythmicity_lmrAMP
diffExpressed Non-rhythmic Rhythmic
         DOWN          138       63
         NO          14637      681
         UP             78       73
         
         
             Rhythmicity_lmrAMP_rel
diffExpressed Non-rhythmic Rhythmic
         DOWN          165       36
         NO          14859      459
         UP            102       49
        


Applying AMP threshold seems to make a bigger proportion of rhythmic genes to be DE!?


logFC > 1, rAMP > 0.58 (p < 0.05, p > 0.1):
             Rhythmicity_lmrAMP
diffExpressed Ambiguous Non-rhythmic Rhythmic
         DOWN        37           66       24
         NO        5054        10270      117
         UP          36           36       30

             Rhythmicity_lmrAMP
diffExpressed Ambiguous Non-rhythmic Rhythmic
         DOWN        32           42       19
         NO        4826         8504       82
         UP          33           28       23

 ---
             Rhythmicity_lmrAMP
diffExpressed Ambiguous Non-rhythmic Rhythmic
         DOWN         7          176       76
         NO          67        12470      557
         UP           8          130       98


```{r}
table <- table(JTK_deLimma[, c('diffExpressed', 'Rhythmicity_lmrAMP')])

DEratio <- (table[1, ] + table[3,]) / colSums(table)
DEratio
```

DEratio:
Non-rhythmic     Rhythmic 
  0.01462363   0.16245487


logFC > 1, rAMP > 0.58 (p < 0.05, p > 0.1):
Ambiguous Non-rhythmic     Rhythmic 
 0.014238346  0.009834169  0.315789474 
 
   Ambiguous Non-rhythmic     Rhythmic 
 0.013289716  0.008164217  0.338709677
 
 ---
    Ambiguous Non-rhythmic     Rhythmic 
  0.18292683   0.02395116   0.23803010 


```{r}
# Check for NA values (seems all to be in LAG)
table(which(is.na(JTK_deLimma), arr.ind=TRUE)[,2])

sum(is.na(JTK_deLimma$LAG))
sum(is.na(JTK_deLimma$gene_name))
```




```{r}
volcano_data_rhythmic <- JTK_deLimma |> dplyr::filter(Rhythmicity_lmrAMP=='Rhythmic')
volcano_data_nonrhythmic <- JTK_deLimma |> dplyr::filter(Rhythmicity_lmrAMP=='Non-rhythmic')

ggplot(data=volcano_data_rhythmic, aes(x=logFC, y=-log10(adj.P.Val), col = diffExpressed_fac)) +
  geom_vline(xintercept = c(-0.58, 0.58), col='gray', linetype='dashed') +
  geom_hline(yintercept=c(-log10(0.05)), col='gray', linetype='dashed') +
  geom_point(size=1) +
  scale_color_manual(values = c("#00AFBB", "grey", "#FC4E07", "black"), 
                     labels = c("Downregulated", "Not significant", "Upregulated", "black")) +
  ggtitle("Volcano plot of rhythmic genes in wild type") +
  labs(colour = "Differential expression")+ 
  xlim(-6, 6) +
  ylim(-1, 55) +
  labs(
    x = expression(log[2]~fold~change),
    y = expression(-log[10]~adjusted~p~value),
    colour = NULL) +
  geom_text_repel(data = dplyr::filter(volcano_data_rhythmic, abs(logFC) > 2 | -log10(adj.P.Val) > 30), aes(, label = gene_name), hjust=0, vjust=0, color = "black", size = 2.5) + 
  theme_bw(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.6),
    axis.text = element_text(colour = "black"),
    axis.title = element_text(colour = "black", size = 12),
    plot.title = element_text(colour = "black", size = 12, hjust = 0),
    legend.key = element_blank()
  )

ggsave(
  filename = "../../results/Plots/Bignon_volcano_plot_rhythmic.svg",
  plot = get_last_plot(),
  width = 8, 
  height = 5, 
  units = "in"
)


ggplot(data=volcano_data_nonrhythmic, aes(x=logFC, y=-log10(adj.P.Val
), col = diffExpressed_fac)) +
  geom_vline(xintercept = c(-0.58, 0.58), col='gray', linetype='dashed') +
  geom_hline(yintercept=c(-log10(0.05)), col='gray', linetype='dashed') +
  geom_point(size=1) +
  scale_color_manual(values = c("#00AFBB", "grey", "#FC4E07"), 
                     labels = c("Downregulated", "Not significant", "Upregulated")) +
  ggtitle("Volcano plot of non-rhythmic genes in wild type") +
  labs(colour = "Differential expression")+ 
  xlim(-6, 6) +
  ylim(-1, 55) +
  labs(
    x = expression(log[2]~fold~change),
    y = expression(-log[10]~adjusted~p~value),
    colour = NULL) +
  geom_text_repel(data = dplyr::filter(volcano_data_nonrhythmic, abs(logFC) > 2.5 | -log10(adj.P.Val) > 40), aes(, label = gene_name), hjust=0, vjust=0, color = "black", size = 2.5) +
  theme_bw(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.6),
    axis.text = element_text(colour = "black"),
    axis.title = element_text(colour = "black", size = 12),
    plot.title = element_text(colour = "black", size = 12, hjust = 0),
    legend.key = element_blank()
  )

ggsave(
  filename = "../../results/Plots/Bignon_volcano_plot_nonrhythmic.svg",
  plot = get_last_plot(),
  width = 8, 
  height = 5, 
  units = "in"
)
```


### Annotation of rhythmic & DE genes

#### a. DE & rhythmic
```{r}
DE_rhythmic_genes <- JTK_deLimma[JTK_deLimma$BH.Q < 0.05 & 
                                   JTK_deLimma$amp_lmr > 0.29 & 
                                   abs(JTK_deLimma$logFC) > 0.58 & 
                                   JTK_deLimma$adj.P.Val < 0.05]

DE_rhythmic_genes_annot <- right_join(annotations[match(DE_rhythmic_genes$gene_id, annotations$gene_id),], DE_rhythmic_genes) |>
  arrange(desc(abs(logFC))) |> as.matrix()

write.table(DE_rhythmic_genes_annot, file = "../../results/Bignon_results/Bignon_DE_rhythmic_genes_annot.tsv", sep = "\t", row.names = FALSE)
```
```{r}
sig_symbols <- DE_rhythmic_genes$gene_name


sig_map <- AnnotationDbi::select(
  org.Mm.eg.db,
  keys = sig_symbols,
  columns = "ENTREZID",
  keytype = "SYMBOL"
)
sig_ids <- unique(na.omit(sig_map$ENTREZID))


bg_id <- rownames(raw_count_filtered)
bg_map <- AnnotationDbi::select(
  org.Mm.eg.db,
  keys = bg_id,
  columns = "ENTREZID",
  keytype = "ENSEMBL"
)
bg_ids <- unique(na.omit(bg_map$ENTREZID))


ego <- enrichGO(
  gene = sig_ids,
  universe = bg_ids,
  OrgDb = org.Mm.eg.db,
  keyType = "ENTREZID",
  ont = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff = 0.1,
  qvalueCutoff = 0.1,
  readable = TRUE
)


go_res <- as.data.frame(ego)
head(go_res)
pdf("../../results/Plots/Bignon_GO_DE_rhythmic.pdf", width = 6, height = 4)
dotplot(ego, showCategory = 10)
dev.off()

dotplot(ego, showCategory = 10)
```

#### a2. DE & rhythmic (high AMP)
```{r}
DE_rhythmic_genes_AMP <- JTK_deLimma[JTK_deLimma$BH.Q < 0.05 & 
                                   JTK_deLimma$amp_lmr > 0.58 & 
                                   abs(JTK_deLimma$logFC) > 0.58 & 
                                   JTK_deLimma$adj.P.Val < 0.05]

DE_rhythmic_genes_AMP_annot <- right_join(annotations[match(DE_rhythmic_genes_AMP$gene_id, annotations$gene_id),], DE_rhythmic_genes_AMP) |>
  arrange(desc(abs(logFC))) |> as.matrix()

write.table(DE_rhythmic_genes_AMP, file = "../../results/Bignon_results/Bignon_DE_rhythmic_genes_AMP_annot.tsv", sep = "\t", row.names = FALSE)
```
```{r}
sig_symbols <- DE_rhythmic_genes_AMP$gene_name
sig_map <- AnnotationDbi::select(
  org.Mm.eg.db,
  keys = sig_symbols,
  columns = "ENTREZID",
  keytype = "SYMBOL"
)
sig_ids <- unique(na.omit(sig_map$ENTREZID))


bg_id <- rownames(raw_count_filtered)
bg_map <- AnnotationDbi::select(
  org.Mm.eg.db,
  keys = bg_id,
  columns = "ENTREZID",
  keytype = "ENSEMBL"
)
bg_ids <- unique(na.omit(bg_map$ENTREZID))


ego <- enrichGO(
  gene = sig_ids,
  universe = bg_ids,
  OrgDb = org.Mm.eg.db,
  keyType = "ENTREZID",
  ont = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff = 0.05,
  qvalueCutoff = 0.05,
  readable = TRUE
)


go_res <- as.data.frame(ego)
head(go_res)
pdf("../../results/Plots/Bignon_GO_DE_rhythmic_AMP.pdf", width = 6, height = 4)
dotplot(ego, showCategory = 10)
dev.off()

dotplot(ego, showCategory = 10)
```

#### b. DE only
```{r}
DE_genes <- JTK_deLimma[abs(JTK_deLimma$logFC) > 0.58 & 
                                   JTK_deLimma$adj.P.Val < 0.05]

DE_genes_annot <- right_join(annotations[match(DE_genes$gene_id, annotations$gene_id),], DE_genes) |>
  arrange(desc(abs(logFC))) |> as.matrix()

write.table(DE_genes_annot, file = "../../results/Bignon_results/Bignon_DE_genes_annot.tsv", sep = "\t", row.names = FALSE)
```
```{r}
sig_symbols <- DE_genes$gene_name


sig_map <- AnnotationDbi::select(
  org.Mm.eg.db,
  keys = sig_symbols,
  columns = "ENTREZID",
  keytype = "SYMBOL"
)
sig_ids <- unique(na.omit(sig_map$ENTREZID))


bg_id <- rownames(raw_count_filtered)
bg_map <- AnnotationDbi::select(
  org.Mm.eg.db,
  keys = bg_id,
  columns = "ENTREZID",
  keytype = "ENSEMBL"
)
bg_ids <- unique(na.omit(bg_map$ENTREZID))


ego <- enrichGO(
  gene = sig_ids,
  universe = bg_ids,
  OrgDb = org.Mm.eg.db,
  keyType = "ENTREZID",
  ont = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff = 0.05,
  qvalueCutoff = 0.05,
  readable = TRUE
)

go_res <- as.data.frame(ego)
head(go_res)

pdf("../../results/Plots/Bignon_GO_DE.pdf", width = 6, height = 4)
dotplot(ego, showCategory = 10)
dev.off()

dotplot(ego, showCategory = 10)

```

#### c. Rhythmic only
```{r}
rhythmic_genes <- JTK_deLimma[JTK_deLimma$BH.Q < 0.05 & 
                                   JTK_deLimma$amp_lmr > 0.29]

rhythmic_genes_annot <- right_join(annotations[match(rhythmic_genes$gene_id, annotations$gene_id),], rhythmic_genes) |>
  arrange(desc(abs(logFC))) |> as.matrix()

write.table(rhythmic_genes_annot, file = "../../results/Bignon_results/Bignon_rhythmic_genes_annot.tsv", sep = "\t", row.names = FALSE)
```
```{r}
sig_symbols <- rhythmic_genes$gene_name


sig_map <- AnnotationDbi::select(
  org.Mm.eg.db,
  keys = sig_symbols,
  columns = "ENTREZID",
  keytype = "SYMBOL"
)
sig_ids <- unique(na.omit(sig_map$ENTREZID))


bg_id <- rownames(raw_count_filtered)
bg_map <- AnnotationDbi::select(
  org.Mm.eg.db,
  keys = bg_id,
  columns = "ENTREZID",
  keytype = "ENSEMBL"
)
bg_ids <- unique(na.omit(bg_map$ENTREZID))


ego <- enrichGO(
  gene = sig_ids,
  universe = bg_ids,
  OrgDb = org.Mm.eg.db,
  keyType = "ENTREZID",
  ont = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff = 0.05,
  qvalueCutoff = 0.05,
  readable = TRUE
)

go_res <- as.data.frame(ego)
head(go_res)

pdf("../../results/Plots/Bignon_GO_rhythmic.pdf", width = 6, height = 4)
dotplot(ego, showCategory = 10)
dev.off()

```


### *Comparison of variance & means*

```{r}
JTK_deLimma_mean <- JTK_deLimma |>
  left_join(rownames_to_column(as.data.frame(mean_wt), "gene_id")) |>
  left_join(rownames_to_column(as.data.frame(mean_ko), "gene_id")) |>
  mutate(abs_diff = 2^mean_ko - 2^mean_wt) |> 
  dplyr::relocate(mean_wt, mean_ko, abs_diff, .after = gene_name)
mean_wt[geneIdsNow]
mean_ko[geneIdsNow]

```


```{r}
var_rhythmic_wt <- var(dplyr::filter(JTK_deLimma_mean, Rhythmicity == "Rhythmic")$mean_wt)
var_rhythmic_ko <- var(dplyr::filter(JTK_deLimma_mean, Rhythmicity == "Rhythmic")$mean_ko)
var_nonrhythmic_wt <- var(dplyr::filter(JTK_deLimma_mean, Rhythmicity == "Non-rhythmic")$mean_wt)
var_nonrhythmic_ko <- var(dplyr::filter(JTK_deLimma_mean, Rhythmicity == "Non-rhythmic")$mean_ko)
var_rhythmic_diff <- var(dplyr::filter(JTK_deLimma_mean, Rhythmicity == "Rhythmic")$abs_diff)
var_nonrhythmic_diff <- var(dplyr::filter(JTK_deLimma_mean, Rhythmicity == "Non-rhythmic")$abs_diff)

var_rhythmic_wt
var_rhythmic_ko
var_nonrhythmic_wt
var_nonrhythmic_ko

var(mean_wt)
var(mean_ko)

as.data.frame(raw_filtered_norm_logcpm_prior_RUVcount_k2)
fit$sigma
```

```{r}
### Variance analysis attempt!!

y_var_C = raw_filtered_norm_logcpm_prior_RUVcount_k2[, filter(metadata_lmr, cond == "C")$experiment_accession]
y_var_KO = raw_filtered_norm_logcpm_prior_RUVcount_k2[, filter(metadata_lmr, cond == "KO")$experiment_accession]

design_var_C = model.matrix(~ time_cos + time_sin, data = metadata_lmr[experiment_accession %in% filter(metadata_lmr, cond == "C")$experiment_accession])
design_var_KO = model.matrix(~ time_cos + time_sin, data = metadata_lmr[experiment_accession %in% filter(metadata_lmr, cond == "KO")$experiment_accession])



fit_C = lmFit(y_var_C, design_var_C)
fit_C = eBayes(fit_C, trend = TRUE)
varLimma_C = data.table(topTable(fit_C, coef = 2, number = Inf), keep.rownames = TRUE)
setnames(varLimma_C, 'rn', 'gene_id')

fit_KO = lmFit(y_var_KO, design_var_KO)
fit_KO = eBayes(fit_KO, trend = TRUE)
varLimma_KO = data.table(topTable(fit_KO, coef = 2, number = Inf), keep.rownames = TRUE)
setnames(varLimma_KO, 'rn', 'gene_id')


fit_C$sigma
fit_KO$sigma

c(fit_C$sigma, fit_KO$sigma)

```

```{r}
sigma <- as.data.frame()
```


```{r}
hist(fit_C$sigma)
```
```{r}
hist(fit_KO$sigma)
```


Not really change in variance??

```{r}
var_rhythmic_diff
var_nonrhythmic_diff # but this doesn't mean anything...

```


```{r}
mean(abs(dplyr::filter(JTK_deLimma_mean, Rhythmicity == "Rhythmic")$abs_diff))
mean(abs(dplyr::filter(JTK_deLimma_mean, Rhythmicity == "Non-rhythmic")$abs_diff))
```

Average of absolute change in mean level is higher in rhythmic > non-rhythmic










### Abs vs rel AMP threshold

```{r}
# plotting logFC vs amp
JTK_deLimma
JTK_deLimma$DE_col <- ifelse(JTK_deLimma$diffExpressed_fac == "NO", "Not DE", "DE")
ggplot(JTK_deLimma, aes(x = amp_lmr, y = logFC)) +
  geom_vline(xintercept = c(0.29), col='gray', linetype='dashed') +
  geom_hline(yintercept=c(-0.58, 0.58), col='gray', linetype='dashed') +
  geom_hline(yintercept=c(0), col='black', lwd = 0.2) +
  geom_point(alpha = 0.5) +
  theme_minimal() +
  xlab("log2 relative amplitude") +
  ylab("log2 Fold Change") +
  ggtitle("DE vs rhythmic amplitude") +
  geom_text_repel(
    data = dplyr::filter(JTK_deLimma, abs(logFC) > 1.2 & amp_lmr > 0.8 | amp_lmr > 1.5), 
    aes(, label = gene_name), 
    hjust=0, vjust=0, 
    color = "red", 
    size = 2.5) + 
  theme(
    panel.grid = element_blank(),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.6),
    axis.text = element_text(colour = "black"),
    axis.title = element_text(colour = "black", size = 12),
    plot.title = element_text(colour = "black", size = 12, hjust = 0),
    legend.key = element_blank()
  )

ggsave(
  filename = "../../results/Plots/Bignon_DE_amp.svg",
  plot = get_last_plot(),
  width = 7, 
  height = 5, 
  units = "in"
)

cor(JTK_deLimma$logFC, JTK_deLimma$amp_lmr)

```
Doesn't really have a correlation between FC and rhythmicity => do no have bias for more DE with more rhythmicity!


```{r}
JTK_deLimma$amp_bin <- cut(JTK_deLimma$AMP, breaks = 4)

JTK_deLimma$amp_bin <- cut(
  JTK_deLimma$AMP,
  breaks = quantile(JTK_deLimma$AMP, probs = seq(0,1,0.25)),
  include.lowest = TRUE
)

ggplot(JTK_deLimma, aes(x = amp_bin, y = logFC)) +
  geom_boxplot() +
  theme_classic() +
  labs(x = "Amplitude bins", y = "log2 Fold Change")
```


```{r}
ggplot(JTK_deLimma[JTK_deLimma$Rhythmicity_fac == "Rhythmic", ], aes(x = amp_lmr, y = logFC, color = Rhythmicity_fac)) +
  geom_point(alpha = 0.6) +
  scale_color_manual(values = c(
    "Non-rhythmic" = "transparent",
    "Rhythmic" = "royalblue"
  )) +
  theme_minimal(base_size = 14) +
  xlab("Rhythmic absolute amplitude") +
  ylab("log2 Fold Change") +
  xlim(0, 3) +
  ggtitle("logFC vs rhythmic amplitude by rhythmicity") +
  theme(
    legend.title = element_text(face = "bold"),
    legend.position = "right"
  ) +
  geom_text(data = dplyr::filter(JTK_deLimma, (amp_lmr > 1.1 | abs(logFC) > 2.5) & amp_lmr > 0.3), aes(, label = gene_name), 
            hjust=0, vjust=0, color = "black", size = 2)


ggplot(JTK_deLimma_mean, aes(x = amp_lmr_abs, y = abs_diff)) +
  geom_point(alpha = 0.6, color = "coral") +
  # scale_color_manual(values = c(
  #   "Non-rhythmic" = "transparent",
  #   "Rhythmic" = "royalblue"
  # )) +
  theme_minimal(base_size = 14) +
  xlab("Absolute amplitude") +
  ylab("Absolute difference") +
  ggtitle("Absolute difference vs absolute amplitude") +
  theme(
    legend.title = element_text(face = "bold"),
    legend.position = "right"
  ) +
  geom_text(data = dplyr::filter(JTK_deLimma_mean, amp_lmr_abs > 250), aes(, label = gene_name), 
            hjust=0, vjust=0, color = "black", size = 3)
```


### 3D plot (abs &)
```{r}
### Relative AMP
AMP_threshold_list <- seq(0, 2, by=0.05)
logFC_threshold_list <- seq(0, 5, by=0.1)


threshold_result <- data.frame(
  AMP = NA,
  logFC = NA,
  ratio = NA,
  rhythmic_DE = NA,
  DE = NA,
  rhythmic = NA,
  log_rhythmic_DE = NA
)


# 3d threshold plot data
for (logFC_threshold in logFC_threshold_list){
  for (AMP_threshold in AMP_threshold_list){
    N_rhythmic_DE <- nrow(JTK_deLimma[abs(JTK_deLimma$logFC) > logFC_threshold & JTK_deLimma$adj.P.Val < 0.05 & JTK_deLimma$BH.Q < 0.05 & JTK_deLimma$amp_lmr > AMP_threshold])
    N_rhythmic <- nrow(JTK_deLimma[JTK_deLimma$BH.Q < 0.05 & JTK_deLimma$amp_lmr > AMP_threshold])
    # print(c(AMP_threshold, N_rhythmic))
    N_DE <- nrow(JTK_deLimma[abs(JTK_deLimma$logFC) > logFC_threshold & JTK_deLimma$adj.P.Val < 0.05])
    
    if (N_rhythmic == 0){
      DE_ratio = NA
    } 
    else{
      DE_ratio <- N_rhythmic_DE/N_rhythmic
    }
    threshold_result <- rbind(threshold_result, c(AMP = AMP_threshold, logFC = logFC_threshold, ratio = DE_ratio, rhythmic_DE = N_rhythmic_DE, DE = N_DE, rhythmic = N_rhythmic, log10_rhythmic_DE = log10(N_rhythmic_DE)))
  }
}

threshold_result <- drop_na(threshold_result)
threshold_result_finite <- threshold_result[is.finite(threshold_result$log_rhythmic_DE), ]

threshold_result[threshold_result$AMP == 0.55 & threshold_result$logFC == 1.0, ]

```



```{r}
### Absolute AMP threshold
AMP_abs_threshold_list <- seq(0, 300, by=10)
difference_threshold_list <- seq(0, 1100, by=20)


threshold_result_abs <- data.frame(
  AMP_abs = NA,
  difference = NA,
  ratio = NA,
  rhythmic_DE = NA,
  DE = NA,
  rhythmic = NA
)


# 3d threshold plot data
for (difference_threshold in difference_threshold_list){
  for (AMP_abs_threshold in AMP_abs_threshold_list){
    N_rhythmic_DE <- nrow(JTK_deLimma_mean[abs(JTK_deLimma_mean$abs_diff) > difference_threshold & 
                                             JTK_deLimma_mean$adj.P.Val < 0.05 & 
                                             JTK_deLimma_mean$BH.Q < 0.05 & 
                                             JTK_deLimma_mean$amp_lmr_abs > AMP_abs_threshold])
    N_rhythmic <- nrow(JTK_deLimma_mean[JTK_deLimma_mean$BH.Q < 0.05 & 
                                          JTK_deLimma_mean$amp_lmr_abs > AMP_abs_threshold])
    # print(c(AMP_threshold, N_rhythmic))
    N_DE <- nrow(JTK_deLimma_mean[abs(JTK_deLimma_mean$abs_diff) > difference_threshold & 
                                    JTK_deLimma_mean$adj.P.Val < 0.05])
    
    if (N_rhythmic == 0){
      DE_ratio = NA
    } 
    else{
      DE_ratio <- N_rhythmic_DE/N_rhythmic
    }
    threshold_result_abs <- rbind(threshold_result_abs, c(AMP_abs = AMP_abs_threshold, difference = difference_threshold, ratio = DE_ratio, rhythmic_DE = N_rhythmic_DE, DE = N_DE, rhythmic = N_rhythmic))
  }
}

threshold_result_abs <- drop_na(threshold_result_abs)



```



```{r}
plot_df <- threshold_result |>
  filter(
    abs(logFC - 0.3) < 1e-6 |
    abs(logFC - 0.5) < 1e-6 |
    abs(logFC - 0.7) < 1e-6 |
    abs(logFC - 1) < 1e-6
  ) |>
  mutate(logFC_label = paste0("|logFC| ≥ ", round(logFC, 2)))

plot_df

ggplot(plot_df, aes(x = AMP, y = ratio, color = log_rhythmic_DE)) +
  geom_point(size = 1) +
  facet_wrap(~ logFC_label, ncol = 2, nrow = 2) +
  labs(
    x = "Relative amplitude threshold",
    y = "Proportion of rhythmic genes that are DE",
    color = "# rhythmic DE genes"
  ) +
  ylim(0, 0.9) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.6),
    axis.text = element_text(colour = "black"),
    axis.title = element_text(colour = "black", size = 12),
    axis.title.x = element_text(margin = margin(t = 10)),
    axis.title.y = element_text(margin = margin(r = 10)),
    plot.title = element_text(colour = "black", size = 12, hjust = 0),
    legend.key = element_blank(),
    legend.title.align = 0.5
  ) +
  scale_color_viridis_c(
    breaks = c(0, 1, 2, 3),
    labels = c("1", "10", "100", "1000"),
    values = scales::rescale(c(0, 1, 2, 3)),
    limits = c(0, 3)
  )

ggsave(
  filename = "../../results/Plots/Bignon_AMP_threshold_fixedFC.svg",
  plot = get_last_plot(),
  width = 8, 
  height = 5, 
  units = "in"
)
```


*But I wonder what's the point of having very high proportion when the denominator is just like 5...*


```{r}
logFC_plot <- threshold_result[near(threshold_result$AMP, 0.5),]

sum(near(threshold_result$AMP, 0.5))

ggplot(logFC_plot) +
  geom_point(mapping = aes(x = logFC, y = log(ratio))) +
  xlim(0, 2)
```

What does this linear relationship mean?? does it suggest the DE is quite random? not really following a trend / source of differential expression?


```{r}

fig <- plot_ly(threshold_result_finite, x = ~logFC, y = ~AMP, z = ~ratio,
                mode = "markers",
                marker = list(
                  size = 3,
                  color = ~log_rhythmic_DE,
                  colorscale = "Rainbow",
                  cmin = 0,
                  cmax = 3.2,
                  colorbar = list(
                    title = "# DE-rhythmic genes",
                    tickmode = "array",
                    tickvals = c(0, 1, 2, 3),
                    ticktext = c(1, 10, 100, 1000),
                    x = 0.8,
                    y = 0.5,
                    len = 0.5,    
                    thickness = 15
                  )
                ),
               hovertemplate = paste(" logFC: %{x}<br>",
                                     "AMP: %{y}<br>",
                                     "ratio: %{z:.2f}<extra></extra>"))
#fig <- fig %>% add_markers()
fig <- fig %>% layout(title = list(text="(Relative) DE ratio for various logFC and relative AMP thresholds", y = 0.95, x = 0.5, xanchor = 'center', yanchor = "top"),
  scene = list(xaxis = list(title = 'log2 Fold Change <br> Magnitude',
                                                range = c(0, 3.5)),
                                   yaxis = list(title = 'Amplitude (relative)',
                                                range = c(0, 1.5)),
                                   zaxis = list(title = 'Ratio of DE genes : Rhythmic genes',
                                                range = c(0, 0.9)),
               aspectmode = "cube")
  )

htmlwidgets::saveWidget(fig, "../../results/Plots/Bignon_AMP_FC_threshold_3D.html", selfcontained = TRUE)
fig
```


```{r}

fig <- plot_ly(threshold_result_abs, x = ~difference, y = ~AMP_abs, z = ~ratio, color = ~AMP_abs, size = I(30))
fig <- fig %>% add_markers()
fig <- fig %>% layout(title = list(text="(Absolute) DE ratio for various differences and absolute AMP thresholds", y = 0.95, x = 0.5, xanchor = 'center', yanchor = "top"),
  scene = list(xaxis = list(title = 'Absolute Difference Magnitude',
                                                range = c(0, 1100)),
                                   yaxis = list(title = 'Amplitude (absolute)',
                                                range = c(0, 300)),
                                   zaxis = list(title = 'Ratio of DE genes : Rhythmic genes',
                                                range = c(0, 1))))

fig

annotations[annotations$gene_name == "H2bc3",]
annotation_selection[annotation_selection$gene_name == "H2bc3",]
"H2b" %in% "H2bdf"
annotation_selection[grepl("H2a", annotation_selection$gene_name),]
```

### Validaton of reasonable DE analysis

```{r}
## Plot most & least DE
most_DE <- head(deLimma$gene_id, 3)
least_DE <- tail(deLimma$gene_id, 3)

metadata_dt <- as.data.table(metadata)

geneIdsNow = most_DE
geneSymbolsNow = annotations[match(geneIdsNow, annotations$gene_id),]$gene_name

df = data.table(t(y[geneIdsNow, ]))
setnames(df, geneSymbolsNow)
df[, experiment_accession := colnames(y[geneIdsNow, ])]

df = merge(df, metadata_dt[, .(experiment_accession, genotype, time)], by = 'experiment_accession')
df = melt(df, measure.vars = geneSymbolsNow, variable.name = 'symbol',
          value.name = 'expr')
df[, symbol := factor(symbol, geneSymbolsNow)]
df[, time := as.integer(time)]
df[, genotype := factor(genotype,
                        labels = c("Bmal1_WT", "Bmal1_KO"),
                        levels = c("C", "KO"))]

ggplot(df) +
  facet_grid(symbol ~ genotype, scales = 'free_y') +
  geom_point(aes(x = time, y = expr, shape = genotype, color = symbol), size = 2) +
  labs(x = 'Zeitgeber time (h)', y = 'Expression (log)') +
  scale_shape(solid = FALSE, guide = 'none') +
  scale_color_brewer(type = 'qual', palette = 'Dark2', guide = 'none') +
  scale_x_continuous(limits = c(0, 24), breaks = seq(0, 24, 4))

ggsave(
  filename = "../../results/Plots/Bignon_mostDE.svg",
  plot = get_last_plot()
)


geneIdsNow = least_DE
geneSymbolsNow = annotations[match(geneIdsNow, annotations$gene_id),]$gene_name

df = data.table(t(y[geneIdsNow, ]))
setnames(df, geneSymbolsNow)
df[, experiment_accession := colnames(y[geneIdsNow, ])]

df = merge(df, metadata_dt[, .(experiment_accession, genotype, time)], by = 'experiment_accession')
df = melt(df, measure.vars = geneSymbolsNow, variable.name = 'symbol',
          value.name = 'expr')
df[, symbol := factor(symbol, geneSymbolsNow)]
df[, time := as.integer(time)]
df[, genotype := factor(genotype,
                        labels = c("Bmal1_WT", "Bmal1_KO"),
                        levels = c("C", "KO"))]

ggplot(df) +
  facet_grid(symbol ~ genotype, scales = 'free_y') +
  geom_point(aes(x = time, y = expr, shape = genotype, color = symbol), size = 2) +
  labs(x = 'Zeitgeber time (h)', y = 'Expression (log)') +
  scale_shape(solid = FALSE, guide = 'none') +
  scale_color_brewer(type = 'qual', palette = 'Dark2', guide = 'none') +
  scale_x_continuous(limits = c(0, 24), breaks = seq(0, 24, 4))

ggsave(
  filename = "../../results/Plots/Bignon_leastDE.svg",
  plot = get_last_plot()
)
```

```{r}
### Plot known rhythmic vs non-rhythmic

geneSymbolsNow = c("Per1", "Per2", "Dbp", "Clock")
geneIdsNow = annotations[match(geneSymbolsNow, annotations$gene_name),]$gene_id

df = data.table(t(y[geneIdsNow, ]))
setnames(df, geneSymbolsNow)
df[, experiment_accession := colnames(y[geneIdsNow, ])]

df = merge(df, metadata_dt[, .(experiment_accession, genotype, time)], by = 'experiment_accession') 
df = df[genotype == "C", ]
df = melt(df, measure.vars = geneSymbolsNow, variable.name = 'symbol',
          value.name = 'expr')
df[, symbol := factor(symbol, geneSymbolsNow)]
df[, time := as.integer(time)]
df[, genotype := factor(genotype,
                        labels = c("Bmal1_WT"),
                        levels = c("C"))]

ggplot(df) +
  facet_grid(symbol ~ genotype, scales = 'free_y') +
  geom_point(aes(x = time, y = expr, shape = genotype, color = symbol), size = 2) +
  labs(x = 'Zeitgeber time (h)', y = 'Expression (log)') +
  scale_shape(solid = FALSE, guide = 'none') +
  scale_color_brewer(type = 'qual', palette = 'Dark2', guide = 'none') +
  scale_x_continuous(limits = c(0, 24), breaks = seq(0, 24, 4))

ggsave(
  filename = "../../results/Plots/Bignon_known_rhythmic.svg",
  plot = get_last_plot()
)

JTK_output_RUV[JTK_output_RUV$gene_id %in% geneIdsNow,]



geneSymbolsNow = c("Ppia", "Tbp", "Rplp0", "Actb")
geneIdsNow = annotations[match(geneSymbolsNow, annotations$gene_name),]$gene_id

df = data.table(t(y[geneIdsNow, ]))
setnames(df, geneSymbolsNow)
df[, experiment_accession := colnames(y[geneIdsNow, ])]

df = merge(df, metadata_dt[, .(experiment_accession, genotype, time)], by = 'experiment_accession') 
df = df[genotype == "C", ]
df = melt(df, measure.vars = geneSymbolsNow, variable.name = 'symbol',
          value.name = 'expr')
df[, symbol := factor(symbol, geneSymbolsNow)]
df[, time := as.integer(time)]
df[, genotype := factor(genotype,
                        labels = c("Bmal1_WT"),
                        levels = c("C"))]


ggplot(df) +
  facet_grid(symbol ~ genotype, scales = 'free_y') +
  geom_point(aes(x = time, y = expr, shape = genotype, color = symbol), size = 2) +
  labs(x = 'Zeitgeber time (h)', y = 'Expression (log)') +
  scale_shape(solid = FALSE, guide = 'none') +
  scale_color_brewer(type = 'qual', palette = 'Dark2', guide = 'none') +
  scale_x_continuous(limits = c(0, 24), breaks = seq(0, 24, 4))

ggsave(
  filename = "../../results/Plots/Bignon_known_arrhythmic.svg",
  plot = get_last_plot()
)

JTK_output_RUV[JTK_output_RUV$gene_id %in% geneIdsNow,]
```




```{r}
### Plot genes (using limorhyde's version now, to be modifie)

metadata_dt <- as.data.table(metadata)

geneSymbolsNow = c("Per1", "Per2", "Per3", "Cry1", "Cry2", "Arntl")
#, "Clock", "Npas2", "Nr1d1","Dbp", "Actb", "Kcnj15"
geneIdsNow = annotations[match(geneSymbolsNow, annotations$gene_name),]$gene_id


df = data.table(t(y[geneIdsNow, ]))
setnames(df, geneSymbolsNow)
df[, experiment_accession := colnames(y[geneIdsNow, ])]

df = merge(df, metadata_dt[, .(experiment_accession, genotype, time)], by = 'experiment_accession')
df = melt(df, measure.vars = geneSymbolsNow, variable.name = 'symbol',
          value.name = 'expr')
df[, symbol := factor(symbol, geneSymbolsNow)]
df[, time := as.integer(time)]
df[, genotype := factor(genotype,
                        labels = c("Bmal1_WT", "Bmal1_KO"),
                        levels = c("C", "KO"))]

ggplot(df) +
  facet_grid(symbol ~ genotype, scales = 'free_y') +
  geom_point(aes(x = time, y = expr, shape = genotype, color = symbol), size = 2) +
  labs(x = 'Zeitgeber time (h)', y = 'Expression (log)') +
  scale_shape(solid = FALSE, guide = 'none') +
  scale_color_brewer(type = 'qual', palette = 'Dark2', guide = 'none') +
  scale_x_continuous(limits = c(0, 24), breaks = seq(0, 24, 4))

ggsave(
  filename = "../../results/Plots/Bignon_oscillations1.svg",
  plot = get_last_plot()
)

## This shows the log2 value!, hence the AMP ~ relative AMP
```

```{r}

geneSymbolsNow = c("Clock", "Npas2", "Nr1d1","Dbp", "Actb", "Rplp0")
geneIdsNow = annotations[match(geneSymbolsNow, annotations$gene_name),]$gene_id


df = data.table(t(y[geneIdsNow, ]))
setnames(df, geneSymbolsNow)
df[, experiment_accession := colnames(y[geneIdsNow, ])]

df = merge(df, metadata_dt[, .(experiment_accession, genotype, time)], by = 'experiment_accession')
df = melt(df, measure.vars = geneSymbolsNow, variable.name = 'symbol',
          value.name = 'expr')
df[, symbol := factor(symbol, geneSymbolsNow)]
df[, time := as.integer(time)]
df[, genotype := factor(genotype,
                        labels = c("Bmal1_WT", "Bmal1_KO"),
                        levels = c("C", "KO"))]

ggplot(df) +
  facet_grid(symbol ~ genotype, scales = 'free_y') +
  geom_point(aes(x = time, y = expr, shape = genotype, color = symbol), size = 2) +
  labs(x = 'Zeitgeber time (h)', y = 'Expression (log)') +
  scale_shape(solid = FALSE, guide = 'none') +
  scale_color_brewer(type = 'qual', palette = 'Dark2', guide = 'none') +
  scale_x_continuous(limits = c(0, 24), breaks = seq(0, 24, 4))

ggsave(
  filename = "../../results/Plots/Bignon_oscillations2.svg",
  plot = get_last_plot()
)
```



```{r}

ggplot(df) +
  facet_grid(symbol ~ genotype, scales = 'free_y') +
  geom_point(aes(x = time, y = 2^(expr), shape = genotype, color = symbol), size = 2) +
  labs(x = 'Zeitgeber time (h)', y = 'Expression (linear)') +
  scale_shape(solid = FALSE, guide = 'none') +
  scale_color_brewer(type = 'qual', palette = 'Dark2', guide = 'none') +
  scale_x_continuous(limits = c(0, 24), breaks = seq(0, 24, 4))

```

