---
title: "R Notebook"
output: html_notebook
---


```{r}
mamba_folder <- "/Users/elysialuo/miniforge3"
env_name <- "circadian"
.libPaths(paste(mamba_folder, "envs", env_name, "lib/R/library", sep = "/"))



install.packages(c("MASS", "janitor", "naniar", "patchwork"))
install.packages(c("scales", "tidyverse"))

install.packages("devtools")
devtools::install_github("naef-lab/dryR")
install.packages('simphony')
install.packages('stringr')
install.packages('ggpubr')

library(tidyverse)
library("dryR")
library(stringr)
library(ggplot2)
library(ggpubr)

library(MASS)
library(janitor)
library(naniar)
library(patchwork)
library(scales)
library(cluster)

```

```{r}
processed_data <- read_csv("../../data/processed/Bignon_RNA-seq_processed.csv")
processed_data
```

```{r}

# prepare arguments
countData = simData[["countData"]]
group     = simData[["group"]]
time      = simData[["time"]]

# prefilter
keep <- rowSums(countData) > 0
countData <- countData[keep,]

# run the analysis for count data (e.g. RNA-Seq data)
dryList   = dryseq(countData,group,time)

# explore the results
dryList[["results"]]     # data frame summarizing results
dryList[["parameters"]]  # coefficients: phase, amplitude and mean for each group
dryList[["ncounts"]]     # normalized counts
dryList[["counts"]]      # raw counts
dryList[["cook"]]        # cook's distance for outlier detection
dryList[["BICW_rhythm"]] # BICW for each rhythmic model
dryList[["BICW_mean"]]   # BICW for each mean model

# generate a pdf with a global summary of all models
plot_models_rhythm(dryList, "./")

# plot a feature of interest
dry_plot(dryList, "feature_113")

```

### Generate processed count matrix

```{r}
processed_data
count_data_drylm <- processed_data |>
    column_to_rownames("Gene_ID") |>
    select(-c(Gene_Name, description, Gene_type))

# Check for missing values or empty rows
table(rowSums(count_data_drylm) == 0) 
table(is.na(count_data_drylm))

group_drylm <- as.data.frame(t(count_data_drylm)) |>
    rownames_to_column("Label") |>
    select(Label) |> t()

time_drylm <- str_extract(group_drylm, 'ZT(\\d*)_', group = 1) |> strtoi (base=10)

# test <- 'ZT0_KO3'
group_drylm_new <- str_extract(group_drylm, 'ZT\\d*_(\\D*)', group = 1)
group_drylm_new

replicate_drylm <- c(1:5)
replicate_drylm <- rep(replicate_drylm, times=12)


```

```{r}

drylm_result = drylm(count_data_drylm,group_drylm_new,time_drylm, period = 24)

# explore the results
drylm_result[["results"]]     # data frame summarizing results
drylm_result[["parameters"]]  # coefficients: phase, amplitude and mean for each group
drylm_result[["ncounts"]]     # normalized counts
drylm_result[["counts"]]      # raw counts (NULL)
drylm_result[["cook"]]        # cook's distance for outlier detection (NULL)
drylm_result[["BICW_rhythm"]] # BICW for each rhythmic model
drylm_result[["BICW_mean"]]   # BICW for each mean model

# generate a pdf with a global summary of all models
plot_models_rhythm(drylm_result, "./")

# plot a feature of interest
dry_plot(drylm_result, "ENSMUSG00000000001")


```

This is what the Bignon paper did, they used drylm on logCPM normalised (?) data.

### Generate raw count matrix

```{r}
install.packages("BiocManager")
BiocManager::install("GEOquery")
library(GEOquery)
library(stringr)
library(tidyr)

gse <- getGEO(filename = "../../Count data/GSE216252_series_matrix.txt.gz", GSEMatrix = TRUE,
    getGPL = FALSE)

meta <- gse[[1]]
gsm <- gse[[2]]
label_df <- as.data.frame(t(rbind(gsm, meta))) |> separate_wider_delim(cols = meta, delim = ", ", names = c("sample", "genotype", "timepoint", "replicate")) |> separate_wider_delim(cols = replicate, delim = " ", names = c("rep", "replicate")) |> separate_wider_delim(cols = genotype, delim = " ", names = c("genotype", "gen"))

meta_dryseq <- label_df |>
  mutate(genotype_new = recode(genotype, control = "C", cKO = "KO")) |> 
  mutate(timepoint_new = str_remove(timepoint, "ZT")) |>
  dplyr::select(gsm, genotype_new, timepoint_new, replicate) |> column_to_rownames("gsm")


```

```{r}
### run whole chunk
setwd("../../Count data/GSE216252_RAW")

file_list <- list.files(".")
count_matrix <- read.delim(file_list[1]) |> dplyr::select(Gene_ID)

for (file in file_list){
  gsm <- str_extract(file, '([[:alnum:]]*)_\\S*', group = 1)
  count_matrix <- read.delim(file) |> 
      rename_with(~gsm, 2) |> 
    full_join(count_matrix)
}

```

```{r}
summary(count_matrix)
sum(is.na(count_matrix))
count_matrix

```


```{r}
# sort gsm (i.e.) in count_matrix
count_data_dryseq <- count_matrix[1:(nrow(count_matrix)-5),] |> remove_rownames() |>
  column_to_rownames("Gene_ID")  # remove last five lines of garbage
count_data_dryseq <- count_data_dryseq[,order(colnames(count_data_dryseq))] |> as.matrix()

tail(count_data_dryseq)
  
is.unsorted(meta_dryseq$gsm) # check if the metadata is sorted

group_dryseq <- meta_dryseq$genotype_new 
time_dryseq <- strtoi(meta_dryseq$timepoint_new, base=10)
```

# Data distribution check

```{r}

library_sizes <- colSums(count_data_dryseq)
library_sizes

boxplot(count_data_dryseq[,1:20])

cor_mat <- cor(count_data_dryseq, method="spearman")
heatmap(cor_mat)

```

1. rather uniform sequence depth
2. some very highly expressed genes -> normalisation needed

```{r}
# prefilter
keep <- rowSums(count_data_dryseq) > 0
count_data_dryseq_prefilter <- count_data_dryseq[keep,]
nrow(count_data_dryseq_prefilter)

dryseq_result = dryseq(count_data_dryseq_prefilter, group_dryseq, time_dryseq)

# explore the results
dryseq_result[["results"]]     # data frame summarizing results
dryseq_result[["parameters"]]  # coefficients: phase, amplitude and mean for each group
dryseq_result[["ncounts"]]     # normalized counts
dryseq_result[["counts"]]      # raw counts
dryseq_result[["cook"]]        # cook's distance for outlier detection
dryseq_result[["BICW_rhythm"]] # BICW for each rhythmic model
dryseq_result[["BICW_mean"]]   # BICW for each mean model

# generate a pdf with a global summary of all models
plot_models_rhythm(dryseq_result, "./")

# plot a feature of interest
dry_plot(dryseq_result, "ENSMUSG00000000440")
dry_plot(dryseq_result, "ENSMUSG00000000441")
dry_plot(dryseq_result, "ENSMUSG00000000486")

dryseq_parameters <- dryseq_result[["parameters"]]
table(dryseq_parameters$chosen_model)

```
    1     2     3     4     5 
26307  1369   786  4222   240

Bruhhh it seems that the RNA-seq data is too skewed (with a few extremely big values), and it breaks the negative binomial (NB) algorithm in dryseq...

HUH why did it suddenly work again when I changed to LMB mac (R version issue? or did I not filter away the rowsum=0???)


```{r}
count_data_dryseq_sorted <- as.data.frame(count_data_dryseq[order(rowSums(count_data_dryseq), decreasing=TRUE),])

BiocManager::install("AnnotationHub")
BiocManager::install("ensembldb")
library(AnnotationHub)
library(AnnotationDbi)
library(ensembldb)

ah <- AnnotationHub()

# search for id of desired dataset
dbResult <- query(ah, c("EnsDb", "Mus musculus", "102")) 
dbResult[1]

# load mouse transcript dataset
MouseEnsDb <- ah[[dbResult$ah_id]]
MouseEnsDb

# output data of interest in form of data.frame
annotations <- genes(MouseEnsDb, return.type = "data.frame")
annotations

annotation_selection <- annotations |> 
    dplyr::select(gene_id, gene_name, description, entrezid) |>
    dplyr::filter(gene_id %in% rownames(count_data_dryseq_sorted))

length(annotation_selection$entrezid)
length(unique(annotation_selection$entrezid))
sum(is.na(annotation_selection$entrezid))



count_data_dryseq_sorted_annot <- count_data_dryseq_sorted |> 
    rownames_to_column("gene_id") |>
    left_join(annotation_selection, "gene_id") |>
    relocate(c(gene_name, description, entrezid), .after=gene_id)
```

- The extremely high read are all related to kidney functioning / respiration
bruhhh okay let's go with drylm first


### Drylm result analysis
```{r}
BiocManager::install("limma")
library(limma)

drylm_parameters <- drylm_result[["parameters"]]
#tail(drylm_result[["parameters"]])

table(drylm_parameters$chosen_model)
nrow(drylm_parameters)
nrow(count_data_drylm)
nrow(count_data_dryseq)
sum(rowSums(count_data_dryseq) > 100)

```
   1    2    3    4    5 
7021 1624 1067 5252  582 


Why the number of transcripts in in drylm dataset is so low?? what did they remove??
* 23-11-2025: filtering

```{r}
# plot a feature of interest
# model 3: gain-of rhythm
dry_plot(drylm_result, "ENSMUSG00000022698")
dry_plot(drylm_result, "ENSMUSG00000054893")

# model 5: altered rhythm
dry_plot(drylm_result, "ENSMUSG00000059824")
dry_plot(drylm_result, "ENSMUSG00000028957")

# model 2: loss of rhythm
dry_plot(drylm_result, "ENSMUSG00000026077")
```
```{r}
ggpaired(drylm_parameters[1:100,], cond1 = "mean_C", cond2 = "mean_KO")
#plot(drylm_parameters$mean_C, drylm_parameters$mean_KO)
```

Oh it seems like the reason they choose limma over DEseq is because limma accepts normalised values but DEseq only accepts raw count!
* 27-11-25: actually no, it's the statistics

```{r}
drylm_parameters_limma <- dplyr::select(drylm_parameters, c(mean_C, mean_KO))
group_limma <- factor(c("C", "KO"))
design <- model.matrix(~ group_limma)

fit <- lmFit(drylm_parameters_limma, design)
# fit <- eBayes(fit)
# topTable(fit, coef = "groupTreatment", number = Inf)

count_data_drylm

```

__You cannot perform a significance test on two means unless you know the variability behind those means.__ HUHHHHHH

```{r}
meta_drylm <- rbind(gse[[2]], group_drylm, group_drylm_new, time_drylm, replicate_drylm)
meta_drylm <- as.data.frame(t(meta_drylm)) |> dplyr::rename(gene_id = V1, label = Label, group = group_drylm_new, time = time_drylm, replicate = replicate_drylm)
meta_drylm

# calculate mean value over timepoints

t_count_data_drylm <- as.data.frame(t(count_data_drylm)) |> group_by(group = meta_drylm$group, replicate = meta_drylm$replicate) 


```

nope it doesn't work, the replicates over each timepoint are not paired.


### Comparing the rhythmic output from two methods

drylm:
   1    2    3    4    5 
7021 1624 1067 5252  582 

dryseq:
    1     2     3     4     5 
26307  1369   786  4222   240

1 nonrhythmic
2 loss of rhythm
3 gain
4 unaltered rhythm
5 altered 

```{r}

common_2 <- drylm_parameters |> 
    rownames_to_column("gene_id") |> 
    dplyr::filter(chosen_model == 2) |>
    dplyr::filter(gene_id %in% rownames(dplyr::filter(dryseq_parameters, chosen_model == 2)))
 
nrow(common_2)

common_3 <- drylm_parameters |> 
    rownames_to_column("gene_id") |> 
    dplyr::filter(chosen_model == 3) |>
    dplyr::filter(gene_id %in% rownames(dplyr::filter(dryseq_parameters, chosen_model == 3)))
 
nrow(common_3)

common_4 <- drylm_parameters |> 
    rownames_to_column("gene_id") |> 
    dplyr::filter(chosen_model == 4) |>
    dplyr::filter(gene_id %in% rownames(dplyr::filter(dryseq_parameters, chosen_model == 4)))
 
nrow(common_4)

common_5 <- drylm_parameters |> 
    rownames_to_column("gene_id") |> 
    dplyr::filter(chosen_model == 5) |>
    dplyr::filter(gene_id %in% rownames(dplyr::filter(dryseq_parameters, chosen_model == 5)))
 
nrow(common_5)

dryseq_result


```

erm two methods not completely the same but ok-ish overlap


### Following Bignon 2023's method

```{r}
# 1. Convert to CPM

# Compute CPM
count_data_dryseq_cpm <- as.data.frame(count_data_dryseq) |>
  mutate(across(.cols = everything(), .fns = ~ . * 1000000 / sum(.), .names = "{col}"))

# Filter for at least CPM >= 1 in one sample
count_data_dryseq_cpm_filtered <- dplyr::filter(count_data_dryseq_cpm, apply(count_data_dryseq_cpm, 1, max) >= 1)
count_data_dryseq_cpm_filtered
count_data_drylm_sorted <- count_data_drylm[order(rownames(count_data_drylm), decreasing=FALSE),]
sum(rownames(count_data_dryseq_cpm_filtered)==rownames(count_data_drylm_sorted))
nrow(count_data_dryseq_cpm_filtered)
nrow(count_data_dryseq)
nrow(count_data_drylm)

```
-> filtered row no. match that in Bignon 2023!



```{r}
# 2. TMM normalisation

BiocManager::install("edgeR")
library(edgeR)

# Filter raw count based on cpm
count_data_dryseq_filtered <- count_data_dryseq[rownames(count_data_dryseq) %in% rownames(count_data_dryseq_cpm_filtered),]
nrow(count_data_dryseq_filtered)

# Normalise library size of filtered raw count
count_data_dryseq_filtered_norm <- DGEList(counts=count_data_dryseq_filtered) |> normLibSizes()
str(count_data_dryseq_filtered_norm)
count_data_dryseq_filtered_norm$samples

# log2 cpm with normalised library size
dryseq_filtered_norm_logcpm <- cpm(count_data_dryseq_filtered_norm, log=TRUE)


```



```{r}
# PCA plot: CPM filtered raw count
pca_test_raw <- prcomp(t(count_data_dryseq_filtered), center=TRUE, scale=TRUE)
pca_df_raw <- as.data.frame(pca_test_raw$x)
pca_df_raw <- cbind(pca_df_raw, meta_dryseq[rownames(pca_df_raw),])

ggplot(pca_df_raw, aes(PC1, PC2, color = genotype_new, shape = timepoint_new)) +
  geom_point(size = 2) +
  theme_classic() +
  labs(title = "PCA: timepoint & genotype",
       x = paste0("PC1 (", round(summary(pca_test_raw)$importance[2,1] * 100, 1), "%)"),
       y = paste0("PC2 (", round(summary(pca_test_raw)$importance[2,2] * 100, 1), "%)"))


```


```{r}

## PCA plot: TMM_log_cpm normalised

pca_test <- prcomp(t(dryseq_filtered_norm_logcpm), center=TRUE, scale=TRUE)
summary(pca_test)

pca_df_norm <- as.data.frame(pca_test$x)
pca_df_norm <- cbind(pca_df_norm, meta_dryseq[rownames(pca_df_norm),])

ggplot(pca_df_norm, aes(PC1, PC2, color = genotype_new, shape = timepoint_new)) +
  geom_point(size = 2) +
  theme_classic() +
  labs(title = "PCA: timepoint & genotype",
       x = paste0("PC1 (", round(summary(pca_test)$importance[2,1] * 100, 1), "%)"),
       y = paste0("PC2 (", round(summary(pca_test)$importance[2,2] * 100, 1), "%)"))

```


### RUV attempt
```{r}
BiocManager::install("RUVSeq")
library(RUVSeq)

```
```{r}
## filtered, unnormalised
boxplot(count_data_dryseq_filtered[,1:10])
plotRLE(count_data_dryseq_filtered[,1:10], outlint=FALSE)

## filtered, log2 cpm normalised
boxplot(dryseq_filtered_norm_logcpm[,1:10])
```
seems pretty normalised after log2 cpm only?! doesn't justify the need for RUV! or does PCA help justify???

```{r}
x <- str_c(meta_dryseq[,1],meta_dryseq[,2],sep="_")
differences <- makeGroups(x)
controls <- rownames(dryseq_filtered_norm_logcpm)

# k=2
RUVs_output <- RUVs(dryseq_filtered_norm_logcpm, controls, k=2, differences, isLog=TRUE)
dryseq_filtered_norm_logcpm_RUVcount_k2 <- RUVs_output$normalizedCounts
as.data.frame(dryseq_filtered_norm_logcpm)
as.data.frame(dryseq_filtered_norm_logcpm_RUVcount_k2)

# k=1
RUVs_output_k1 <- RUVs(dryseq_filtered_norm_logcpm, controls, k=1, differences, isLog=TRUE)
dryseq_filtered_norm_logcpm_RUVcount_k1 <- RUVs_output_k1$normalizedCounts

# k=5
RUVs_output_k5 <- RUVs(dryseq_filtered_norm_logcpm, controls, k=5, differences, isLog=TRUE)
dryseq_filtered_norm_logcpm_RUVcount_k5 <- RUVs_output_k5$normalizedCounts

# k=10
RUVs_output_k10 <- RUVs(dryseq_filtered_norm_logcpm, controls, k=10, differences, isLog=TRUE)
dryseq_filtered_norm_logcpm_RUVcount_k10 <- RUVs_output_k10$normalizedCounts
```

How valid is mixing C and KO in covariance analysis?

```{r}
boxplot(RUVs_output$normalizedCounts[,1:10])
```
```{r}

# PCA plot, k=1
pca_test <- prcomp(t(dryseq_filtered_norm_logcpm_RUVcount_k1), center=TRUE, scale=TRUE)
summary(pca_test)
pca_df_norm <- as.data.frame(pca_test$x)
pca_df_norm <- cbind(pca_df_norm, meta_dryseq[rownames(pca_df_norm),])
ggplot(pca_df_norm, aes(PC1, PC2, color = genotype_new, shape = timepoint_new)) +
  geom_point(size = 2) +
  theme_classic() +
  labs(title = "PCA: timepoint & genotype (k=1)",
       x = paste0("PC1 (", round(summary(pca_test)$importance[2,1] * 100, 1), "%)"),
       y = paste0("PC2 (", round(summary(pca_test)$importance[2,2] * 100, 1), "%)"))

# PCA plot, k=2
pca_test <- prcomp(t(dryseq_filtered_norm_logcpm_RUVcount_k2), center=TRUE, scale=TRUE)
summary(pca_test)
pca_df_norm <- as.data.frame(pca_test$x)
pca_df_norm <- cbind(pca_df_norm, meta_dryseq[rownames(pca_df_norm),])
ggplot(pca_df_norm, aes(PC1, PC2, color = genotype_new, shape = timepoint_new)) +
  geom_point(size = 2) +
  theme_classic() +
  labs(title = "PCA: timepoint & genotype (k=2)",
       x = paste0("PC1 (", round(summary(pca_test)$importance[2,1] * 100, 1), "%)"),
       y = paste0("PC2 (", round(summary(pca_test)$importance[2,2] * 100, 1), "%)"))

# PCA plot, k=5
pca_test <- prcomp(t(dryseq_filtered_norm_logcpm_RUVcount_k5), center=TRUE, scale=TRUE)
summary(pca_test)
pca_df_norm <- as.data.frame(pca_test$x)
pca_df_norm <- cbind(pca_df_norm, meta_dryseq[rownames(pca_df_norm),])
ggplot(pca_df_norm, aes(PC1, PC2, color = genotype_new, shape = timepoint_new)) +
  geom_point(size = 2) +
  theme_classic() +
  labs(title = "PCA: timepoint & genotype (k=5)",
       x = paste0("PC1 (", round(summary(pca_test)$importance[2,1] * 100, 1), "%)"),
       y = paste0("PC2 (", round(summary(pca_test)$importance[2,2] * 100, 1), "%)"))

# PCA plot, k=10
pca_test <- prcomp(t(dryseq_filtered_norm_logcpm_RUVcount_k10), center=TRUE, scale=TRUE)
summary(pca_test)
pca_df_norm <- as.data.frame(pca_test$x)
pca_df_norm <- cbind(pca_df_norm, meta_dryseq[rownames(pca_df_norm),])
ggplot(pca_df_norm, aes(PC1, PC2, color = genotype_new, shape = timepoint_new)) +
  geom_point(size = 2) +
  theme_classic() +
  labs(title = "PCA: timepoint & genotype (k=10)",
       x = paste0("PC1 (", round(summary(pca_test)$importance[2,1] * 100, 1), "%)"),
       y = paste0("PC2 (", round(summary(pca_test)$importance[2,2] * 100, 1), "%)"))
```
looks very similar to the paper's plot!! but idk why the RUV normalised number are similar but slightly off (+- 0.01??)

```{r}
plotRLE(count_data_dryseq_filtered[,1:20], outline=FALSE)
plotRLE(dryseq_filtered_norm_logcpm[,1:20], outline=FALSE)
plotRLE(dryseq_filtered_norm_logcpm_RUVcount_k1[,1:20], outline=FALSE)
plotRLE(dryseq_filtered_norm_logcpm_RUVcount_k2[,1:20], outline=FALSE)
plotRLE(dryseq_filtered_norm_logcpm_RUVcount_k5[,1:20], outline=FALSE)
plotRLE(dryseq_filtered_norm_logcpm_RUVcount_k10[,1:20], outline=FALSE)


```

Is there a need for RUV correction!?!??!
