---
title: "R Notebook"
output:
  html_document:
    df_print: paged
---


```{r}
mamba_folder <- "/Users/elysialuo/miniforge3"
env_name <- "circadian"
.libPaths(paste(mamba_folder, "envs", env_name, "lib/R/library", sep = "/"))



install.packages(c("MASS", "janitor", "naniar", "patchwork"))
install.packages(c("scales", "tidyverse"))

install.packages("devtools")
devtools::install_github("naef-lab/dryR")
install.packages('simphony')
install.packages('stringr')
install.packages('ggpubr')
install.packages('limma')

library(tidyverse)
library("dryR")
library(stringr)
library(ggplot2)
library(ggpubr)

library(MASS)
library(janitor)
library(naniar)
library(patchwork)
library(scales)
library(cluster)

library(RColorBrewer)
library(ggrepel)
```

```{r}
processed_data <- read_csv("../../data/processed/Bignon_RNA-seq_processed.csv")
processed_data
```
### DryR usage example
```{r}

# prepare arguments
countData = simData[["countData"]]
group     = simData[["group"]]
time      = simData[["time"]]

# prefilter
keep <- rowSums(countData) > 0
countData <- countData[keep,]

# run the analysis for count data (e.g. RNA-Seq data)
dryList   = dryseq(countData,group,time)

# explore the results
dryList[["results"]]     # data frame summarizing results
dryList[["parameters"]]  # coefficients: phase, amplitude and mean for each group
dryList[["ncounts"]]     # normalized counts
dryList[["counts"]]      # raw counts
dryList[["cook"]]        # cook's distance for outlier detection
dryList[["BICW_rhythm"]] # BICW for each rhythmic model
dryList[["BICW_mean"]]   # BICW for each mean model

# generate a pdf with a global summary of all models
plot_models_rhythm(dryList, "./")

# plot a feature of interest
dry_plot(dryList, "feature_113")

```

### Generate processed count matrix

```{r}
processed_data
count_data_drylm <- processed_data |>
    column_to_rownames("Gene_ID") |>
    dplyr::select(-c(Gene_Name, description, Gene_type))

# Check for missing values or empty rows
table(rowSums(count_data_drylm) == 0) 
table(is.na(count_data_drylm))

group_drylm <- as.data.frame(t(count_data_drylm)) |>
    rownames_to_column("Label") |>
    dplyr::select(Label) |> t()

time_drylm <- str_extract(group_drylm, 'ZT(\\d*)_', group = 1) |> strtoi (base=10)

# test <- 'ZT0_KO3'
group_drylm <- str_extract(group_drylm, 'ZT\\d*_(\\D*)', group = 1)
group_drylm

replicate_drylm <- c(1:5)
replicate_drylm <- rep(replicate_drylm, times=12)


```

```{r}

drylm_result = drylm(count_data_drylm,group_drylm,time_drylm, period = 24)

# explore the results
drylm_result[["results"]]     # data frame summarizing results
drylm_result[["parameters"]]  # coefficients: phase, amplitude and mean for each group
drylm_result[["ncounts"]]     # normalized counts
drylm_result[["counts"]]      # raw counts (NULL)
drylm_result[["cook"]]        # cook's distance for outlier detection (NULL)
drylm_result[["BICW_rhythm"]] # BICW for each rhythmic model
drylm_result[["BICW_mean"]]   # BICW for each mean model

# generate a pdf with a global summary of all models
plot_models_rhythm(drylm_result, "./")

# plot a feature of interest
dry_plot(drylm_result, "ENSMUSG00000000001")


```

This is what the Bignon paper did, they used *drylm* on logCPM normalised (?) data.

### Generate raw count matrix

```{r}
install.packages("BiocManager")
BiocManager::install("GEOquery")
library(GEOquery)
library(stringr)
library(tidyr)
```

```{r}
# Read in metadata

gse <- getGEO(filename = "../../Count data/GSE216252_series_matrix.txt.gz", GSEMatrix = TRUE,
    getGPL = FALSE)

meta <- gse[[1]]
gsm <- gse[[2]]
label_df <- as.data.frame(t(rbind(gsm, meta))) |> separate_wider_delim(cols = meta, delim = ", ", names = c("sample", "genotype", "timepoint", "replicate")) |> separate_wider_delim(cols = replicate, delim = " ", names = c("rep", "replicate")) |> separate_wider_delim(cols = genotype, delim = " ", names = c("genotype", "gen"))

meta_dryseq <- label_df |>
  mutate(genotype_new = recode(genotype, control = "C", cKO = "KO")) |> 
  mutate(timepoint_new = str_remove(timepoint, "ZT")) |>
  dplyr::select(gsm, genotype_new, timepoint_new, replicate) |> column_to_rownames("gsm")


```


```{r}
# Read in raw count
### run whole chunk
setwd("../../Count data/GSE216252_RAW")

file_list <- list.files(".")
count_matrix <- read.delim(file_list[1]) |> dplyr::select(Gene_ID)

for (file in file_list){
  gsm <- str_extract(file, '([[:alnum:]]*)_\\S*', group = 1)
  count_matrix <- read.delim(file) |> 
      rename_with(~gsm, 2) |> 
    full_join(count_matrix)
}

```

```{r}
summary(count_matrix)
sum(is.na(count_matrix))
count_matrix

```

### Clean up data matrix
```{r}
# sort gsm (i.e.) in count_matrix
count_data_dryseq <- count_matrix[1:(nrow(count_matrix)-5),] |> remove_rownames() |>
  column_to_rownames("Gene_ID")  # remove last five lines of garbage
count_data_dryseq <- count_data_dryseq[,order(colnames(count_data_dryseq))] |> as.matrix()

tail(count_data_dryseq)
  
is.unsorted(meta_dryseq$gsm) # check if the metadata is sorted

group_dryseq <- meta_dryseq$genotype_new 
time_dryseq <- strtoi(meta_dryseq$timepoint_new, base=10)
```

# Data distribution check

```{r}

library_sizes <- colSums(count_data_dryseq)
library_sizes

boxplot(count_data_dryseq[,1:20])

cor_mat <- cor(count_data_dryseq, method="spearman")
heatmap(cor_mat)

```

1. rather uniform sequence depth
2. some very highly expressed genes -> normalisation needed

### dryseq pipeline on unnormalised raw count
```{r}
# prefilter
keep <- rowSums(count_data_dryseq) > 0
count_data_dryseq_prefilter <- count_data_dryseq[keep,]
nrow(count_data_dryseq_prefilter)

dryseq_result = dryseq(count_data_dryseq_prefilter, group_dryseq, time_dryseq)

# explore the results
dryseq_result[["results"]]     # data frame summarizing results
dryseq_result[["parameters"]]  # coefficients: phase, amplitude and mean for each group
dryseq_result[["ncounts"]]     # normalized counts
dryseq_result[["counts"]]      # raw counts
dryseq_result[["cook"]]        # cook's distance for outlier detection
dryseq_result[["BICW_rhythm"]] # BICW for each rhythmic model
dryseq_result[["BICW_mean"]]   # BICW for each mean model

# generate a pdf with a global summary of all models
plot_models_rhythm(dryseq_result, "./")

# plot a feature of interest
dry_plot(dryseq_result, "ENSMUSG00000000440")
dry_plot(dryseq_result, "ENSMUSG00000000441")
dry_plot(dryseq_result, "ENSMUSG00000000486")

dryseq_parameters <- dryseq_result[["parameters"]]
table(dryseq_parameters$chosen_model)

```
    1     2     3     4     5 
26307  1369   786  4222   240

Bruhhh it seems that the RNA-seq data is too skewed (with a few extremely big values), and it breaks the negative binomial (NB) algorithm in dryseq...

HUH why did it suddenly work again when I changed to LMB mac (R version issue? or did I not filter away the rowsum=0???)


```{r}
# Sort & annotate
count_data_dryseq_sorted <- as.data.frame(count_data_dryseq[order(rowSums(count_data_dryseq), decreasing=TRUE),])

BiocManager::install("AnnotationHub")
BiocManager::install("ensembldb")
library(AnnotationHub)
library(AnnotationDbi)
library(ensembldb)

ah <- AnnotationHub()

# search for id of desired dataset
dbResult <- query(ah, c("EnsDb", "Mus musculus", "102")) 
dbResult[1]

# load mouse transcript dataset
MouseEnsDb <- ah[[dbResult$ah_id]]
MouseEnsDb

# output data of interest in form of data.frame
annotations <- genes(MouseEnsDb, return.type = "data.frame")
annotations

annotation_selection <- annotations |> 
    dplyr::select(gene_id, gene_name, description, entrezid) |>
    dplyr::filter(gene_id %in% rownames(count_data_dryseq_sorted))

length(annotation_selection$entrezid)
length(unique(annotation_selection$entrezid))
sum(is.na(annotation_selection$entrezid))



count_data_dryseq_sorted_annot <- count_data_dryseq_sorted |> 
    rownames_to_column("gene_id") |>
    left_join(annotation_selection, "gene_id") |>
    relocate(c(gene_name, description, entrezid), .after=gene_id)
```

- The extremely high read are all related to kidney functioning / respiration
bruhhh okay let's go with drylm first


### Drylm result analysis
```{r}
BiocManager::install("limma")
library(limma)

drylm_parameters <- drylm_result[["parameters"]]
#tail(drylm_result[["parameters"]])

table(drylm_parameters$chosen_model)
nrow(drylm_parameters)
nrow(count_data_drylm)
nrow(count_data_dryseq)
sum(rowSums(count_data_dryseq) > 100)

```
   1    2    3    4    5 
7021 1624 1067 5252  582 


Why the number of transcripts in in drylm dataset is so low?? what did they remove??
* 23-11-2025: filtering

```{r}
# plot a feature of interest
# model 3: gain-of rhythm
dry_plot(drylm_result, "ENSMUSG00000022698")
dry_plot(drylm_result, "ENSMUSG00000054893")

# model 5: altered rhythm
dry_plot(drylm_result, "ENSMUSG00000059824")
dry_plot(drylm_result, "ENSMUSG00000028957")

# model 2: loss of rhythm
dry_plot(drylm_result, "ENSMUSG00000026077")
```
```{r}
ggpaired(drylm_parameters[1:100,], cond1 = "mean_C", cond2 = "mean_KO")
#plot(drylm_parameters$mean_C, drylm_parameters$mean_KO)
```

Oh it seems like the reason they choose limma over DEseq is because limma accepts normalised values but DEseq only accepts raw count!
* 27-11-25: actually no, it's the statistics

```{r}
drylm_parameters_limma <- dplyr::select(drylm_parameters, c(mean_C, mean_KO))
group_limma <- factor(c("C", "KO"))
design <- model.matrix(~ group_limma)

fit <- lmFit(drylm_parameters_limma, design)
# fit <- eBayes(fit)
# topTable(fit, coef = "groupTreatment", number = Inf)

count_data_drylm

```

__You cannot perform a significance test on two means unless you know the variability behind those means.__ HUHHHHHH

```{r}
meta_drylm <- rbind(gse[[2]], group_drylm, group_drylm_new, time_drylm, replicate_drylm)
meta_drylm <- as.data.frame(t(meta_drylm)) |> dplyr::rename(gene_id = V1, label = Label, group = group_drylm_new, time = time_drylm, replicate = replicate_drylm)
meta_drylm

# calculate mean value over timepoints

t_count_data_drylm <- as.data.frame(t(count_data_drylm)) |> group_by(group = meta_drylm$group, replicate = meta_drylm$replicate) 


```

nope it doesn't work, the replicates over each timepoint are not paired.


### Comparing the rhythmic output from two methods

drylm:
   1    2    3    4    5 
7021 1624 1067 5252  582 

dryseq:
    1     2     3     4     5 
26307  1369   786  4222   240

1 nonrhythmic
2 loss of rhythm
3 gain
4 unaltered rhythm
5 altered 

```{r}

common_2 <- drylm_parameters |> 
    rownames_to_column("gene_id") |> 
    dplyr::filter(chosen_model == 2) |>
    dplyr::filter(gene_id %in% rownames(dplyr::filter(dryseq_parameters, chosen_model == 2)))
 
nrow(common_2)

common_3 <- drylm_parameters |> 
    rownames_to_column("gene_id") |> 
    dplyr::filter(chosen_model == 3) |>
    dplyr::filter(gene_id %in% rownames(dplyr::filter(dryseq_parameters, chosen_model == 3)))
 
nrow(common_3)

common_4 <- drylm_parameters |> 
    rownames_to_column("gene_id") |> 
    dplyr::filter(chosen_model == 4) |>
    dplyr::filter(gene_id %in% rownames(dplyr::filter(dryseq_parameters, chosen_model == 4)))
 
nrow(common_4)

common_5 <- drylm_parameters |> 
    rownames_to_column("gene_id") |> 
    dplyr::filter(chosen_model == 5) |>
    dplyr::filter(gene_id %in% rownames(dplyr::filter(dryseq_parameters, chosen_model == 5)))
 
nrow(common_5)

dryseq_result


```

erm two methods not completely the same but ok-ish overlap


## Following Bignon 2023's method!!!

```{r}
# 1. Convert to CPM

# Compute CPM
count_data_dryseq_cpm <- as.data.frame(count_data_dryseq) |>
  mutate(across(.cols = everything(), .fns = ~ . * 1000000 / sum(.), .names = "{col}"))

# Filter for at least CPM >= 1 in one sample
count_data_dryseq_cpm_filtered <- dplyr::filter(count_data_dryseq_cpm, apply(count_data_dryseq_cpm, 1, max) >= 1)
count_data_dryseq_cpm_filtered
count_data_drylm_sorted <- count_data_drylm[order(rownames(count_data_drylm), decreasing=FALSE),]
sum(rownames(count_data_dryseq_cpm_filtered)==rownames(count_data_drylm_sorted))
nrow(count_data_dryseq_cpm_filtered)
nrow(count_data_dryseq)
nrow(count_data_drylm_sorted)

```
-> filtered row no. match that in Bignon 2023!



```{r}
# 2. TMM normalisation

BiocManager::install("edgeR")
library(edgeR)

# Filter raw count based on cpm
count_data_dryseq_filtered <- count_data_dryseq[rownames(count_data_dryseq) %in% rownames(count_data_dryseq_cpm_filtered),]
nrow(count_data_dryseq_filtered)

# Normalise library size of filtered raw count
count_data_dryseq_filtered_norm <- DGEList(counts=count_data_dryseq_filtered) |> normLibSizes()
str(count_data_dryseq_filtered_norm)
count_data_dryseq_filtered_norm$samples

# log2 cpm with normalised library size
dryseq_filtered_norm_logcpm <- cpm(count_data_dryseq_filtered_norm, log=TRUE)
dryseq_filtered_norm_logcpm_prior <- cpm(count_data_dryseq_filtered_norm, log=TRUE, prior.count=3)

```

bruhhhh Bignon et al used prior.count=3 without specifying it in the paper......
baddd

```{r}
# PCA plot: CPM filtered raw count
pca_test_raw <- prcomp(t(count_data_dryseq_filtered), center=TRUE, scale=TRUE)
pca_df_raw <- as.data.frame(pca_test_raw$x)
pca_df_raw <- cbind(pca_df_raw, meta_dryseq[rownames(pca_df_raw),])

ggplot(pca_df_raw, aes(PC1, PC2, color = genotype_new, shape = timepoint_new)) +
  geom_point(size = 2) +
  theme_classic() +
  labs(title = "PCA: timepoint & genotype",
       x = paste0("PC1 (", round(summary(pca_test_raw)$importance[2,1] * 100, 1), "%)"),
       y = paste0("PC2 (", round(summary(pca_test_raw)$importance[2,2] * 100, 1), "%)"))


```


```{r}

## PCA plot: TMM_log_cpm normalised

pca_test <- prcomp(t(dryseq_filtered_norm_logcpm), center=TRUE, scale=TRUE)
summary(pca_test)

pca_df_norm <- as.data.frame(pca_test$x)
pca_df_norm <- cbind(pca_df_norm, meta_dryseq[rownames(pca_df_norm),])

ggplot(pca_df_norm, aes(PC1, PC2, color = genotype_new, shape = timepoint_new)) +
  geom_point(size = 2) +
  theme_classic() +
  labs(title = "PCA: timepoint & genotype",
       x = paste0("PC1 (", round(summary(pca_test)$importance[2,1] * 100, 1), "%)"),
       y = paste0("PC2 (", round(summary(pca_test)$importance[2,2] * 100, 1), "%)"))

```


### RUV attempt
```{r}
BiocManager::install("RUVSeq")
library(RUVSeq)
```

```{r}
## filtered, unnormalised
boxplot(count_data_dryseq_filtered[,1:40])
plotRLE(count_data_dryseq_filtered[,1:40], outlint=FALSE)

## filtered, log2 cpm normalised
boxplot(dryseq_filtered_norm_logcpm[,1:40])
```
seems pretty normalised after log2 cpm only?! doesn't justify the need for RUV! or does PCA help justify???

```{r}
x <- str_c(meta_dryseq[,1],meta_dryseq[,2],sep="_")
differences <- makeGroups(x)
controls <- rownames(dryseq_filtered_norm_logcpm)

# k=2
RUVs_output <- RUVs(dryseq_filtered_norm_logcpm, controls, k=2, differences, isLog=TRUE)
dryseq_filtered_norm_logcpm_RUVcount_k2 <- RUVs_output$normalizedCounts
as.data.frame(dryseq_filtered_norm_logcpm)
as.data.frame(dryseq_filtered_norm_logcpm_RUVcount_k2)

# k=2, prior
RUVs_output_prior <- RUVs(dryseq_filtered_norm_logcpm_prior, controls, k=2, differences, isLog=TRUE)
dryseq_filtered_norm_logcpm_prior_RUVcount_k2 <- RUVs_output_prior$normalizedCounts
as.data.frame(dryseq_filtered_norm_logcpm_prior)
as.data.frame(dryseq_filtered_norm_logcpm_prior_RUVcount_k2)

# k=1
RUVs_output_k1 <- RUVs(dryseq_filtered_norm_logcpm, controls, k=1, differences, isLog=TRUE)
dryseq_filtered_norm_logcpm_RUVcount_k1 <- RUVs_output_k1$normalizedCounts

# k=5
RUVs_output_k5 <- RUVs(dryseq_filtered_norm_logcpm, controls, k=5, differences, isLog=TRUE)
dryseq_filtered_norm_logcpm_RUVcount_k5 <- RUVs_output_k5$normalizedCounts

# k=10
RUVs_output_k10 <- RUVs(dryseq_filtered_norm_logcpm, controls, k=10, differences, isLog=TRUE)
dryseq_filtered_norm_logcpm_RUVcount_k10 <- RUVs_output_k10$normalizedCounts
```

How valid is mixing C and KO in covariance analysis?

```{r}

boxplot(RUVs_output_prior$normalizedCounts[,1:40])
```

```{r}
# PCA plots comparison

# PCA plot, k=1
pca_test <- prcomp(t(dryseq_filtered_norm_logcpm_RUVcount_k1), center=TRUE, scale=TRUE)
summary(pca_test)
pca_df_norm <- as.data.frame(pca_test$x)
pca_df_norm <- cbind(pca_df_norm, meta_dryseq[rownames(pca_df_norm),])
ggplot(pca_df_norm, aes(PC1, PC2, color = genotype_new, shape = timepoint_new)) +
  geom_point(size = 2) +
  theme_classic() +
  labs(title = "PCA: timepoint & genotype (k=1)",
       x = paste0("PC1 (", round(summary(pca_test)$importance[2,1] * 100, 1), "%)"),
       y = paste0("PC2 (", round(summary(pca_test)$importance[2,2] * 100, 1), "%)"))

# PCA plot, k=2
pca_test <- prcomp(t(dryseq_filtered_norm_logcpm_prior_RUVcount_k2), center=TRUE, scale=TRUE)
summary(pca_test)
pca_df_norm <- as.data.frame(pca_test$x)
pca_df_norm <- cbind(pca_df_norm, meta_dryseq[rownames(pca_df_norm),])
ggplot(pca_df_norm, aes(PC1, PC2, color = genotype_new, shape = timepoint_new)) +
  geom_point(size = 2) +
  theme_classic() +
  labs(title = "PCA: timepoint & genotype (k=2)",
       x = paste0("PC1 (", round(summary(pca_test)$importance[2,1] * 100, 1), "%)"),
       y = paste0("PC2 (", round(summary(pca_test)$importance[2,2] * 100, 1), "%)"))

# PCA plot, k=2, prior
pca_test <- prcomp(t(dryseq_filtered_norm_logcpm_RUVcount_k2), center=TRUE, scale=TRUE)
summary(pca_test)
pca_df_norm <- as.data.frame(pca_test$x)
pca_df_norm <- cbind(pca_df_norm, meta_dryseq[rownames(pca_df_norm),])
ggplot(pca_df_norm, aes(PC1, PC2, color = genotype_new, shape = timepoint_new)) +
  geom_point(size = 2) +
  theme_classic() +
  labs(title = "PCA: timepoint & genotype (k=2, prior)",
       x = paste0("PC1 (", round(summary(pca_test)$importance[2,1] * 100, 1), "%)"),
       y = paste0("PC2 (", round(summary(pca_test)$importance[2,2] * 100, 1), "%)"))

# PCA plot, k=5
pca_test <- prcomp(t(dryseq_filtered_norm_logcpm_RUVcount_k5), center=TRUE, scale=TRUE)
summary(pca_test)
pca_df_norm <- as.data.frame(pca_test$x)
pca_df_norm <- cbind(pca_df_norm, meta_dryseq[rownames(pca_df_norm),])
ggplot(pca_df_norm, aes(PC1, PC2, color = genotype_new, shape = timepoint_new)) +
  geom_point(size = 2) +
  theme_classic() +
  labs(title = "PCA: timepoint & genotype (k=5)",
       x = paste0("PC1 (", round(summary(pca_test)$importance[2,1] * 100, 1), "%)"),
       y = paste0("PC2 (", round(summary(pca_test)$importance[2,2] * 100, 1), "%)"))

# PCA plot, k=10
pca_test <- prcomp(t(dryseq_filtered_norm_logcpm_RUVcount_k10), center=TRUE, scale=TRUE)
summary(pca_test)
pca_df_norm <- as.data.frame(pca_test$x)
pca_df_norm <- cbind(pca_df_norm, meta_dryseq[rownames(pca_df_norm),])
ggplot(pca_df_norm, aes(PC1, PC2, color = genotype_new, shape = timepoint_new)) +
  geom_point(size = 2) +
  theme_classic() +
  labs(title = "PCA: timepoint & genotype (k=10)",
       x = paste0("PC1 (", round(summary(pca_test)$importance[2,1] * 100, 1), "%)"),
       y = paste0("PC2 (", round(summary(pca_test)$importance[2,2] * 100, 1), "%)"))
```
looks very similar to the paper's plot!! but idk why the RUV normalised number are similar but slightly off (+- 0.01??)
=> Jan: it's due to prior.count=3, but only slightly changes the value!

```{r}
plotRLE(count_data_dryseq_filtered[,1:20], outline=FALSE)
plotRLE(dryseq_filtered_norm_logcpm[,1:20], outline=FALSE)
plotRLE(dryseq_filtered_norm_logcpm_RUVcount_k1[,1:20], outline=FALSE)
plotRLE(dryseq_filtered_norm_logcpm_RUVcount_k2[,1:20], outline=FALSE)
plotRLE(dryseq_filtered_norm_logcpm_RUVcount_k5[,1:20], outline=FALSE)
plotRLE(dryseq_filtered_norm_logcpm_RUVcount_k10[,1:20], outline=FALSE)

```

Is there a need for RUV correction!?!??!
- it indeed separates light vs dark phase data but how much useful info did it remove?




### Limma adventure!!

```{r}

count_data_drylm_sorted

as.data.frame(dryseq_filtered_norm_logcpm)

as.data.frame(RUVs_output$normalizedCounts)

as.data.frame(RUVs_output_prior$normalizedCounts)


```

Bignon used prior.count=3 during TMM normalisation...

```{r}
### Detour to understand splines
library(splines)
as.numeric(meta_dryseq$timepoint_new)
X <- ns(as.numeric(meta_dryseq$timepoint_new), df=5)
dim(X)
```


## a. RUV version

```{r}
meta_dryseq
genotype <- factor(meta_dryseq$genotype_new)

y <- as.data.frame(dryseq_filtered_norm_logcpm)
y_normalised <- RUVs_output_prior$normalizedCounts
W1 <- RUVs_output_prior$W[,1]
W2 <- RUVs_output_prior$W[,2]
sin <- c(0, 0.866, 0.866, 0, -0.866, -0.866) |> rep(2, each=5)
cos <- c(1, 0.5, -0.5, -1, -0.5, 0.5) |> rep(2, each=5)

design <- model.matrix(~ genotype + sin + cos + W1 + W2) 
fit <- lmFit(y, design)
fit <- eBayes(fit, trend=TRUE)
fit_RUVnorm <- lmFit(y_normalised, design)
fit_RUVnorm <- eBayes(fit_RUVnorm, trend=TRUE)

fit$p.value
results <- decideTests(fit, adjust.method = "none")
results_adj <- decideTests(fit, adjust.method = "BH")
results_RUVnorm <- decideTests(fit_RUVnorm, adjust.method = "BH")
summary(results)
summary(results_adj)
summary(results_RUVnorm)


mean_result_level<- as.data.frame(fit$coefficients[,1:2]) |> mutate(mean_WT = `(Intercept)`, mean_KO = `(Intercept)` + genotypeKO, FC = genotypeKO/`(Intercept)`, log2FC = log(genotypeKO/`(Intercept)`, base=2)) |> dplyr::select(-c(`(Intercept)`, genotypeKO))

## is this the correct way to calculate fold_change??????????????

mean_result_sig_filtered <- topTable(fit, coef='genotypeKO', p.value=0.05, adjust.method = "BH", number=Inf)

mean_result <- topTable(fit, coef='genotypeKO', adjust.method = "BH", number=Inf)

```
decideTests does BH p-vlue adjustment by default

1. ~ genotype
      (Intercept) genotypeKO
Down          1445       2575
NotSig         488       9993
Up           13613       2978

2. ~ genotype + W1 + W2
       (Intercept) genotypeKO    W1    W2
Down           665       3204  4823  1560
NotSig        1661       8620  5946 12498
Up           13220       3722  4777  1488

3. ~ genotype + sin + cos 
       (Intercept) genotypeKO   sin   cos
Down          1458       2796  1853  1132
NotSig         464       9547 11671 13416
Up           13624       3203  2022   998

4. ~ genotype + sin + cos + W1 + W2
       (Intercept) genotypeKO   sin   cos    W1    W2
Down           708       3495  2769  1851  4997  1900
NotSig        1587       8007  9986 11780  5540 11986
Up           13251       4044  2791  1915  5009  1660

5. #4 with RUV normalised y as input (not suggested by paper)
       (Intercept) genotypeKO   sin   cos  WW_1  WW_2
Down           617       3498  2776  1846     0     0
NotSig        1603       8001  9970 11786 15546 15546
Up           13326       4047  2800  1914     0     0
(therefore difference not due to this...)

Bignon:
Down    3505
NotSig  7970
Up      4070

quite similar actually... small difference due to FDR??

```{r}
### Checking if limma-trend's usage is reasonable
count_data_dryseq_sorted
colSums(count_data_dryseq_sorted)
```
Limma-trend is simpler than voom, good when sample size fairly consistent. Here the colSum is very close -> limma-trend good to use for the simplicity and speed!!

### Compare with rhythmic models

```{r}
# Re-run dryR on data normalised by myself

count_data_drylm_new <- as.data.frame(RUVs_output_prior$normalizedCounts)

# Check for missing values or empty rows
table(rowSums(count_data_drylm_new) == 0) 
table(is.na(count_data_drylm_new))

group_drylm_new <- meta_dryseq[,'genotype_new']

time_drylm_new <- as.numeric(meta_dryseq[,'timepoint_new'])

replicate_drylm_new <- c(1:5)
replicate_drylm_new <- rep(replicate_drylm, times=12)


drylm_result_new = drylm(count_data_drylm_new,group_drylm_new,time_drylm_new, period = 24)

# explore the results
drylm_result_new[["results"]]     # data frame summarizing results
drylm_result_new[["parameters"]]  # coefficients: phase, amplitude and mean for each group
drylm_result_new[["ncounts"]]     # normalized counts
drylm_result_new[["counts"]]      # raw counts (NULL)
drylm_result_new[["cook"]]        # cook's distance for outlier detection (NULL)
drylm_result_new[["BICW_rhythm"]] # BICW for each rhythmic model
drylm_result_new[["BICW_mean"]]   # BICW for each mean model

# generate a pdf with a global summary of all models
plot_models_rhythm(drylm_result_new, "./")

# plot a feature of interest
dry_plot(drylm_result_new, "ENSMUSG00000000001")


```


```{r}
# mean change classification
mean_result <- as.data.frame(results_adj[,'genotypeKO']) |> rownames_to_column('gene_id')

# rhythm model classification
rhythm_result_raw <- drylm_result_new$parameters |> dplyr::select(chosen_model) |> rownames_to_column('gene_id')

## add in ambiguous classification (group 0)
rhythm_result <- mutate(rhythm_result_raw,chosen_model = replace(chosen_model, drylm_result_new$parameters[,'chosen_model_BICW']<0.65, 0))

```

```{r}
# check model number with Bignon's count
table(column_to_rownames(rhythm_result, 'gene_id'))
```

Bignon:
chosen_model
   0    1    2    3    4    5 
4461 5274  970  535 3936  370 

```{r}
mean_rhythm_result <- full_join(rhythm_result, mean_result, by='gene_id') |> column_to_rownames('gene_id')

table(mean_rhythm_result)
table(column_to_rownames(mean_result, 'gene_id'))
```

1. Without specifying 'ambiguous' data

            genotypeKO
chosen_model   -1    0    1
           1 1288 4139 1594
           2  445  612  567
           3  293  502  272
           4 1229 2615 1408
           5  240  139  203
           
          all 3495 8007 4044 
           
omg i got it!! but the number is off and there's no group 0...
okay i found out why: they categorised some as 'ambiguous' for BICW >0.65(model) and >0.95(mean)...


2. Specifying 'ambiguous' data

            genotypeKO
chosen_model   -1    0    1
           0 1040 2230 1191
           1  942 3179 1153
           2  281  328  361
           3  168  231  136
           4  908 1959 1069
           5  156   80  134
 
          all 3495 8007 4044 

*==> okay this looks more like the result in the paper but how valid is it to draw conclusion ignoring the 'ambiguous' data????????? and why the criteria BICW<0.65????*


1. RUV
            genotypeKO
chosen_model   -1    0    1
           0 1040 2230 1191
           1  942 3179 1153
           2  281  328  361
           3  168  231  136
           4  908 1959 1069
           5  156   80  134
genotypeKO
  -1    0    1 
3495 8007 4044


## b. W/O RUV version

```{r}
meta_dryseq
genotype <- factor(meta_dryseq$genotype_new)

y <- as.data.frame(dryseq_filtered_norm_logcpm)
sin <- c(0, 0.866, 0.866, 0, -0.866, -0.866) |> rep(2, each=5)
cos <- c(1, 0.5, -0.5, -1, -0.5, 0.5) |> rep(2, each=5)

design <- model.matrix(~ genotype + sin + cos) 
fit <- lmFit(y, design)
fit <- eBayes(fit, trend=TRUE)

results_noRUV_adj <- decideTests(fit, adjust.method = "BH")

summary(results_noRUV_adj)

```
       (Intercept) genotypeKO   sin   cos
Down          1458       2796  1853  1132
NotSig         464       9547 11671 13416
Up           13624       3203  2022   998


```{r}
# Re-run dryR (w/o RUV)

count_data_drylm_new_noRUV <- as.data.frame(dryseq_filtered_norm_logcpm)

# Check for missing values or empty rows
table(rowSums(count_data_drylm_new) == 0) 
table(is.na(count_data_drylm_new))

group_drylm_new <- meta_dryseq[,'genotype_new']

time_drylm_new <- as.numeric(meta_dryseq[,'timepoint_new'])

replicate_drylm_new <- c(1:5)
replicate_drylm_new <- rep(replicate_drylm, times=12)


drylm_result_new_noRUV = drylm(count_data_drylm_new_noRUV,group_drylm_new,time_drylm_new, period = 24)

# explore the results
drylm_result_new_noRUV[["results"]]     # data frame summarizing results
drylm_result_new_noRUV[["parameters"]]  # coefficients: phase, amplitude and mean for each group
drylm_result_new_noRUV[["ncounts"]]     # normalized counts
drylm_result_new_noRUV[["counts"]]      # raw counts (NULL)
drylm_result_new_noRUV[["cook"]]        # cook's distance for outlier detection (NULL)
drylm_result_new_noRUV[["BICW_rhythm"]] # BICW for each rhythmic model
drylm_result_new_noRUV[["BICW_mean"]]   # BICW for each mean model

# generate a pdf with a global summary of all models
plot_models_rhythm(drylm_result_new_noRUV, "./")

# plot a feature of interest
dry_plot(drylm_result_new_noRUV, "ENSMUSG00000000001")
```

```{r}
# mean change classification
mean_result_noRUV <- as.data.frame(results_noRUV_adj[,'genotypeKO']) |> rownames_to_column('gene_id')

# rhythm model classification
rhythm_result_noRUV <- drylm_result_new_noRUV$parameters |> dplyr::select(chosen_model) |> rownames_to_column('gene_id')
## add in ambiguous classification (group 0)
rhythm_result_noRUV <- mutate(rhythm_result,chosen_model = replace(chosen_model, drylm_result_new_noRUV$parameters[,'chosen_model_BICW']<0.65, 0))

```

```{r}
# check model number with Bignon's count
table(column_to_rownames(rhythm_result_noRUV, 'gene_id'))
```

1. RUV:
chosen_model
   0    1    2    3    4    5 
4461 5274  970  535 3936  370 

2. no RUV
chosen_model
   0    1    2    3    4    5 
3918 *7477*  792  337 2869  153 
==> no RUV gives fewer rhythmic genes

```{r}
mean_rhythm_result_noRUV <- full_join(rhythm_result_noRUV, mean_result_noRUV, by='gene_id') |> column_to_rownames('gene_id')

table(mean_rhythm_result_noRUV)
table(column_to_rownames(mean_result, 'gene_id'))
```


1. RUV
            genotypeKO
chosen_model   -1    0    1
           0 1040 2230 1191
           1  942 3179 1153
           2  281  328  361
           3  168  231  136
           4  908 1959 1069
           5  156   80  134
genotypeKO
  -1    0    1 
3495 8007 4044


2. no RUV
            genotypeKO
chosen_model   -1    0    1
           0  823 2213  882
           1 1055 5220 1202
           2  228  270  294
           3   82  174   81
           4  545 1641  683
           5   63   29   61
genotypeKO
  -1    0    1 
2796 9547 3203

But similar conclusion can be drawn? Supports 




# Combining group 2,4,5 vs 1,3 vs 0 (as suggested by JO)

Only knowing the number of genes in each category is not informative - need to consider how much they changed too.

```{r}
# Group for result with model 0-5
rhythm_result_grouped <- rhythm_result |> dplyr::mutate(grouped_chosen_model = if_else(chosen_model %in% c(2,4,5), true='rhythmic', false='non-rhythmic')) |> dplyr::mutate(grouped_chosen_model = if_else(chosen_model %in% c(0), true='ambiguous', false=grouped_chosen_model)) 

# Group for result with model 1-5 (pre ambiguity filtering)
rhythm_result_raw_grouped <-rhythm_result_raw |> dplyr::mutate(grouped_chosen_model = if_else(chosen_model %in% c(2,4,5), true='rhythmic', false='non-rhythmic')) |> dplyr::mutate(grouped_chosen_model = if_else(chosen_model %in% c(0), true='ambiguous', false=grouped_chosen_model)) 


mean_rhythm_result_grouped_level_sig_filtered <- mean_result_sig_filtered |> dplyr::select(logFC, P.Value, adj.P.Val) |> rownames_to_column('gene_id') |> left_join(rhythm_result_grouped, by='gene_id')

```

```{r}

```


```{r}
ggplot(data = mean_rhythm_result_grouped_level_sig_filtered,
       mapping = aes(x = grouped_chosen_model, y = logFC, fill = grouped_chosen_model)) + 
  geom_point() + 
  geom_violin(alpha = 0.8, linewidth = 0) +
  geom_boxplot()

ggplot(data = mean_rhythm_result_grouped_level_sig_filtered,
       mapping = aes(x = factor(chosen_model), y = logFC, fill = factor(chosen_model))) + 
  geom_boxplot()
```

```{r}

volcano_data <- mean_result |> dplyr::select(logFC, P.Value, adj.P.Val) |> rownames_to_column('gene_id') |> left_join(rhythm_result_grouped, by='gene_id')

volcano_data$diffExpressed <- 'NO'
volcano_data$diffExpressed[volcano_data$logFC>0.8 & volcano_data$adj.P.Val<0.05] <- 'UP'
volcano_data$diffExpressed[volcano_data$logFC<(-0.8) & volcano_data$adj.P.Val<0.05] <- 'DOWN'

table(volcano_data$diffExpressed)
table(volcano_data[, c('diffExpressed', 'grouped_chosen_model')])
table(volcano_data[, c('diffExpressed', 'chosen_model')])

```
*logFC > 1*
DOWN    NO    UP 
  118 15319   109 
             grouped_chosen_model
diffExpressed ambiguous non-rhythmic rhythmic
         DOWN        40           41       37
         NO        4391         5749     5179
         UP          30           19       60
             chosen_model
diffExpressed    0    1    2    3    4    5
         DOWN   40   37   20    4    9    8
         NO   4391 5221  909  528 3922  348
         UP     30   16   41    3    5   14


*logFC > 0.8*
 DOWN    NO    UP 
  212 15168   166 
             grouped_chosen_model
diffExpressed ambiguous non-rhythmic rhythmic
         DOWN        64           80       68
         NO        4349         5696     5123
         UP          48           33       85
             chosen_model
diffExpressed    0    1    2    3    4    5
         DOWN   64   71   31    9   23   14
         NO   4349 5177  887  519 3900  336
         UP     48   26   52    7   13   20


Bignon's paper didn't set logFC limit??!?! ()


```{r}
volcano_data_rhythmic <- volcano_data |> filter(grouped_chosen_model=='rhythmic')
volcano_data_nonrhythmic <- volcano_data |> filter(grouped_chosen_model=='non-rhythmic')

ggplot(data=volcano_data_rhythmic, aes(x=logFC, y=-log10(adj.P.Val
), col = diffExpressed)) +
  geom_vline(xintercept = c(-0.8, 0.8), col='gray', linetype='dashed') +
  geom_hline(yintercept=c(-log10(0.05)), col='gray', linetype='dashed') +
  geom_point(size=1) +
  scale_color_manual(values = c("#00AFBB", "grey", "#FC4E07"), 
                     labels = c("Downregulated", "Not significant", "Upregulated")) +
  ggtitle("Rhythmic genes") +
  xlim(-5, 5) +
  ylim(-1, 50)

ggplot(data=volcano_data_nonrhythmic, aes(x=logFC, y=-log10(adj.P.Val
), col = diffExpressed)) +
  geom_vline(xintercept = c(-0.8, 0.8), col='gray', linetype='dashed') +
  geom_hline(yintercept=c(-log10(0.05)), col='gray', linetype='dashed') +
  geom_point(size=1) +
  scale_color_manual(values = c("#00AFBB", "grey", "#FC4E07"), 
                     labels = c("Downregulated", "Not significant", "Upregulated")) +
  ggtitle("Non-rhythmic genes") +
  xlim(-5, 5) +
  ylim(-1, 50)

```






# *JTK & RAIN comparison!*

```{r}
# log-TMM normalised matrix, control only

JTK_matrix <- as.data.frame(dryseq_filtered_norm_logcpm[,1:30])
colnames(JTK_matrix) <- meta_drylm$label[1:30]
JTK_matrix <- rownames_to_column(JTK_matrix, 'Probe')

```


```{r}
write.table(JTK_matrix, file = 'JTK_Cycle/Bignon_JTK_data.txt', sep = '\t', row.names = FALSE, col.names = TRUE, quote=FALSE)

write.table(JTK_matrix['Probe'], file = 'JTK_Cycle/Bignon_JTK_annot.txt', sep = '/t', row.names = FALSE, col.names = TRUE, quote=FALSE)

```


--- running JTK cycle ---

```{r}
getwd()
JTK_output <- read.table(file = 'JTK_Cycle/JTK.Bignon_JTK_run.txt', header = TRUE ) |> rename('Probe' = 'gene_id')

# Lable as rythmic & non-rhythmic
JTK_output$Rhythmicity <- 'Non-rhythmic'
JTK_output$Rhythmicity[JTK_output$BH.Q < 0.05] <- 'Rhythmic'
JTK_output <- relocate(JTK_output, Rhythmicity, .before = BH.Q)
table(JTK_output$Rhythmicity)
JTK_output

```

1. JTK: 
Non-rhythmic      Rythmic 
       12778         2768 
       

lol ADJ.P means “p-value adjusted for the discrete null distribution of the JTK test” here
q value = actually BH FDR adjusted p value

How to classify rhythmic vs non-rhythmic here? just q.value<0.05? how is this comparable to the use of BICW 'ambiguous' threshold? 


```{r}
rhythm_result_grouped
table(rhythm_result_grouped$grouped_chosen_model)

rhythm_result_raw_grouped
table(rhythm_result_raw_grouped$grouped_chosen_model)

```

2. dryR 
(ambiguity filtered)
   ambiguous non-rhythmic     rhythmic 
        4461         5809         5276 

(unfiltered)
non-rhythmic     rhythmic 
        8088         7458 

### Venn diagram comparison
```{r}
# Plot Venn diagram
install.packages("VennDiagram")
library(VennDiagram)

## rhythmic (ambiguous filtered)
dryR_rhythmic_list <- rhythm_result_grouped$gene_id[rhythm_result_grouped$grouped_chosen_model == 'rhythmic']
JTK_rhythmic_list <- JTK_output$gene_id[JTK_output$Rhythmicity == 'Rhythmic'] 

length(dryR_rhythmic_list) 
length(JTK_rhythmic_list)


vlist <- list(dryR_rhythmic_list, JTK_rhythmic_list)
names(vlist) <- c("DryR", "JTK_Cycle")

venn.diagram(
  x = vlist,
  category.names = c("DryR", "JTK_Cycle"),
  filename = "VennDiagram/dryR_vs_JTK_rhythmic_filtered.png",
  output = TRUE,
  main = "dryR vs JTK_Cycle (Rhythmic genes, ambiguous filtered)", # Title
  fill = c("lightgreen", "lightcoral"), # Circle fill colors
  cat.col = "black", # Category label colors
  cex = 1.2 # Size of text
)


## rhythmic (unfiltered)
dryR_rhythmic_list <- rhythm_result_raw_grouped$gene_id[rhythm_result_raw_grouped$grouped_chosen_model == 'rhythmic']
JTK_rhythmic_list <- JTK_output$gene_id[JTK_output$Rhythmicity == 'Rhythmic'] 

length(dryR_rhythmic_list) 
length(JTK_rhythmic_list)


vlist <- list(dryR_rhythmic_list, JTK_rhythmic_list)
names(vlist) <- c("DryR", "JTK_Cycle")

venn.diagram(
  x = vlist,
  category.names = c("DryR", "JTK_Cycle"),
  filename = "VennDiagram/dryR_vs_JTK_rhythmic_unfiltered.png",
  output = TRUE,
  main = "dryR vs JTK_Cycle (Rhythmic genes, ambiguous unfiltered)", # Title
  fill = c("lightgreen", "lightcoral"), # Circle fill colors
  cat.col = "black", # Category label colors
  cex = 1.2 # Size of text
)


## non-rhythmic
dryR_rhythmic_list <- rhythm_result_grouped$gene_id[rhythm_result_grouped$grouped_chosen_model == 'non-rhythmic']
JTK_rhythmic_list <- JTK_output$gene_id[JTK_output$Rhythmicity == 'Non-rhythmic'] 

length(dryR_rhythmic_list) 
length(JTK_rhythmic_list)


vlist <- list(dryR_rhythmic_list, JTK_rhythmic_list)
names(vlist) <- c("DryR", "JTK_Cycle")

venn.diagram(
  x = vlist,
  category.names = c("DryR", "JTK_Cycle"),
  filename = "VennDiagram/dryR_vs_JTK_unrhythmic_filtered.png",
  output = TRUE,
  main = "dryR vs JTK_Cycle (Non-rhythmic genes, ambiguous filtered)", # Title
  fill = c("lightgreen", "lightcoral"), # Circle fill colors
  cat.col = "black", # Category label colors
  cex = 1.2 # Size of text
)


dryR_rhythmic_list <- rhythm_result_raw_grouped$gene_id[rhythm_result_raw_grouped$grouped_chosen_model == 'non-rhythmic']
JTK_rhythmic_list <- JTK_output$gene_id[JTK_output$Rhythmicity == 'Non-rhythmic'] 

length(dryR_rhythmic_list) 
length(JTK_rhythmic_list)


vlist <- list(dryR_rhythmic_list, JTK_rhythmic_list)
names(vlist) <- c("DryR", "JTK_Cycle")

venn.diagram(
  x = vlist,
  category.names = c("DryR", "JTK_Cycle"),
  filename = "VennDiagram/dryR_vs_JTK_unrhythmic_unfiltered.png",
  output = TRUE,
  main = "dryR vs JTK_Cycle (Non-rhythmic genes, ambiguous unfiltered)", # Title
  fill = c("lightgreen", "lightcoral"), # Circle fill colors
  cat.col = "black", # Category label colors
  cex = 1.2 # Size of text
)


```



### Volcano plots of JTK!
```{r}
JTK_output_Rhythmicity <- JTK_output |> dplyr::select(gene_id, Rhythmicity)

volcano_data <- mean_result |> dplyr::select(logFC, P.Value, adj.P.Val) |> rownames_to_column('gene_id') |> left_join(JTK_output_Rhythmicity, by='gene_id')

volcano_data$diffExpressed <- 'NO'
volcano_data$diffExpressed[volcano_data$logFC>0.8 & volcano_data$adj.P.Val<0.05] <- 'UP'
volcano_data$diffExpressed[volcano_data$logFC<(-0.8) & volcano_data$adj.P.Val<0.05] <- 'DOWN'

table(volcano_data$diffExpressed)
table(volcano_data[, c('diffExpressed', 'Rhythmicity')])
  
```
 DOWN    NO    UP 
  212 15168   166 
             Rhythmicity
diffExpressed Non-rhythmic Rhythmic
         DOWN          141       71
         NO          12549     2619
         UP             88       78


```{r}
volcano_data_rhythmic <- volcano_data |> filter(Rhythmicity=='Rhythmic')
volcano_data_nonrhythmic <- volcano_data |> filter(Rhythmicity=='Non-rhythmic')

ggplot(data=volcano_data_rhythmic, aes(x=logFC, y=-log10(adj.P.Val
), col = diffExpressed)) +
  geom_vline(xintercept = c(-0.8, 0.8), col='gray', linetype='dashed') +
  geom_hline(yintercept=c(-log10(0.05)), col='gray', linetype='dashed') +
  geom_point(size=1) +
  scale_color_manual(values = c("#00AFBB", "grey", "#FC4E07"), 
                     labels = c("Downregulated", "Not significant", "Upregulated")) +
  ggtitle("Rhythmic genes") +
  xlim(-5, 5) +
  ylim(-1, 50)
  

ggplot(data=volcano_data_nonrhythmic, aes(x=logFC, y=-log10(adj.P.Val
), col = diffExpressed)) +
  geom_vline(xintercept = c(-0.8, 0.8), col='gray', linetype='dashed') +
  geom_hline(yintercept=c(-log10(0.05)), col='gray', linetype='dashed') +
  geom_point(size=1) +
  scale_color_manual(values = c("#00AFBB", "grey", "#FC4E07"), 
                     labels = c("Downregulated", "Not significant", "Upregulated")) +
  ggtitle("Non-rhythmic genes") +
  xlim(-5, 5) +
  ylim(-1, 50)
```


