---
title: "Mehta Pipeline Analysis"
author: "Elysia Luo"
date: "2026-02-28"
output: html_document
---

```{r}
mamba_folder <- "/Users/elysialuo/miniforge3"
env_name <- "circadian"
.libPaths(paste(mamba_folder, "envs", env_name, "lib/R/library", sep = "/"))

library(tidyverse)
library(dryR)
library(stringr)
library(ggplot2)
library(ggpubr)

library(MASS)
library(janitor)
library(naniar)
library(patchwork)
library(scales)
library(cluster)

library(RColorBrewer)
library(ggrepel)

library(GEOquery)
library(stringr)
library(tidyr)

library(AnnotationHub)
library(AnnotationDbi)
library(ensembldb)

library(limma)
library(edgeR)
library(RUVSeq)

library(limorhyde)
library(annotate)
library(data.table)
library(org.Mm.eg.db)
library(foreach)

library(VennDiagram)

library('org.At.tair.db')
library(GEOquery)
library(stringr)
library(tidyr)
library(plotly)
library("RColorBrewer")
```

```{r}
getwd()
```


```{r}
ah <- AnnotationHub()

ah["AH121945"]

# search for id of desired dataset
dbResult <- query(ah, c("org.At.tair.db.sqlite","Arabidopsis")) 


# load mouse transcript dataset
AtEnsDb <- ah[[dbResult$ah_id]]
# class(AtEnsDb)
# keytypes(AtEnsDb)
# columns(AtEnsDb)
# head( keys(AtEnsDb, 'ENTREZID'))
# mapIds(AtEnsDb, keys=head( keys(AtEnsDb, 'ENTREZID')), column='GENENAME', keytype='ENTREZID')
# AtEnsDb[[1]]

# output data of interest in form of data.frame
annotations <- AnnotationDbi::select(AtEnsDb, keys = keys(AtEnsDb), columns = c("SYMBOL", "GENENAME", "ENTREZID"))
annotations
```

```{r}
# read in raw count matrix 
raw_count = read.csv("../../data/Mehta_raw_data/Mehta_raw_count.csv") |> column_to_rownames("Gene")
```


```{r}
# make metadata matrix

metadata <- as.data.frame(colnames(raw_count)) |>
  dplyr::mutate("title" = `colnames(raw_count)`) |>  
  separate_wider_delim(cols = `colnames(raw_count)`, delim = "_", names = c("time", "genotype", "replicate")) |>
  mutate(time = str_pad(str_remove(time, "ZT"), 2, pad="0")) |>
  arrange(genotype, time) |> 
  dplyr::relocate(title, .before = time)

metadata

raw_count <- raw_count[,metadata_all$title]   ### sort raw count matrix
```

    genotype
time lhycca1 prr7prr9 toc1 wt
  01       4        4    4  4
  03       4        4    4  4
  05       4        4    3  4
  07       4        4    4  4
  09       4        4    4  4
  11       4        3    4  4
  13       4        4    4  4
  15       4        4    4  4
  17       4        4    4  3
  19       4        4    4  4
  21       3        4    3  4
  23       4        4    4  4
  
  some only have 3 replicates


### Normalisation

```{r}
# 1. Convert to CPM

# Compute CPM
raw_count_cpm <- raw_count |>
  dplyr::mutate(across(.cols = everything(), .fns = ~ . * 1000000 / sum(.), .names = "{col}"))

# Filter for at least CPM >= 1 in one sample
raw_count_cpm_filtered <- dplyr::filter(raw_count_cpm, apply(raw_count_cpm, 1, max) >= 1)
raw_count_cpm_filtered

```

```{r}
# 2. TMM normalisation

# Filter raw count based on cpm
raw_count_filtered <- raw_count[rownames(raw_count) %in% rownames(raw_count_cpm_filtered),]
nrow(raw_count_filtered)

# Normalise library size of filtered raw count
raw_count_filtered_norm <- DGEList(counts=raw_count_filtered) |> normLibSizes()
str(raw_count_filtered_norm)
raw_count_filtered_norm$samples

# log2 cpm with normalised library size
raw_filtered_norm_logcpm <- cpm(raw_count_filtered_norm, log=TRUE)
raw_filtered_norm_logcpm_prior <- cpm(raw_count_filtered_norm, log=TRUE, prior.count=2)

```

```{r}
cov <- sd(raw_count_filtered_norm$samples$lib.size) / mean(raw_count_filtered_norm$samples$lib.size) * 100
cov
```

Ok-ish library size


```{r}
### RUV
x <- str_c(metadata$genotype, str_pad(metadata$time, 2, pad="0"), sep="_")

differences <- makeGroups(x)  # -1 assigned to missing sample (PerKO20 and PerKO8)
controls <- rownames(raw_filtered_norm_logcpm_prior)

# k=2, prior
RUVs_output_prior <- RUVs(raw_filtered_norm_logcpm_prior, controls, k=2, differences, isLog=TRUE)
raw_filtered_norm_logcpm_prior_RUVcount_k2 <- RUVs_output_prior$normalizedCounts
as.data.frame(raw_filtered_norm_logcpm_prior)
as.data.frame(raw_filtered_norm_logcpm_prior_RUVcount_k2)
W1 <- RUVs_output_prior$W[,1]
W2 <- RUVs_output_prior$W[,2]

```

```{r}
#par(mfrow = c(2,2))

## filtered, unnormalised
plotRLE(as.matrix(raw_count_filtered), outline=FALSE, exprs_logged = FALSE,
        ylim=c(-0.9, 0.9), main = 'Unnormalised (filtered)', xlab = "Samples", ylab = "Relative log")

## filtered, TMM normalised
plotRLE(raw_filtered_norm_logcpm_prior, outline=FALSE, exprs_logged = TRUE,
        ylim=c(-0.9, 0.9), main = 'TMM normalised (filtered)', xlab = "Samples", ylab = "Relative log")

## filtered, TMM & RUV normalised
plotRLE(RUVs_output_prior$normalizedCounts, outline=FALSE, exprs_logged = TRUE,
        ylim=c(-0.9, 0.9), main = 'TMM + RUV normalised (filtered)', xlab = "Samples", ylab = "Relative log")

```

```{r}
# PCA plot: CPM filtered raw count
pca_test_raw <- prcomp(t(raw_count_filtered), center=TRUE, scale=TRUE)
pca_df_raw <- as.data.frame(pca_test_raw$x)
pca_df_raw <- left_join(rownames_to_column(pca_df_raw, "title"), metadata, by = "title")
pca_df_raw$lightdark <- "dark"
pca_df_raw$lightdark[pca_df_raw$time %in% c("01", "03", "05", "07", "09", "11")] <- "light"

ggplot(pca_df_raw, aes(PC1, PC2, color = genotype, shape = lightdark, label = time)) +
  geom_point(size = 2) +
  #geom_text(hjust=0, vjust=0, color = "black", size = 2.5) +
  theme_classic() +
  labs(title = "PCA: timepoint & genotype (raw count)",
       x = paste0("PC1 (", round(summary(pca_test_raw)$importance[2,1] * 100, 1), "%)"),
       y = paste0("PC2 (", round(summary(pca_test_raw)$importance[2,2] * 100, 1), "%)"))


## PCA plot: TMM_log_cpm normalised
pca_test <- prcomp(t(raw_filtered_norm_logcpm_prior), center=TRUE, scale=TRUE)
pca_df_TMM <- as.data.frame(pca_test$x)
pca_df_TMM <- left_join(rownames_to_column(pca_df_TMM, "title"), metadata, by = "title")
pca_df_TMM$lightdark <- "dark"
pca_df_TMM$lightdark[pca_df_raw$time %in% c("01", "03", "05", "07", "09", "11")] <- "light"

ggplot(pca_df_TMM, aes(PC1, PC2, color = genotype, shape = lightdark, label = time)) +
  geom_point(size = 2) +
  #geom_text(hjust=0, vjust=0, color = "black", size = 2.5) +
  theme_classic() +
  labs(title = "PCA: timepoint & genotype (TMM normalised)",
       x = paste0("PC1 (", round(summary(pca_test)$importance[2,1] * 100, 1), "%)"),
       y = paste0("PC2 (", round(summary(pca_test)$importance[2,2] * 100, 1), "%)"))


## PCA plot: RUV normalised, k=2, prior

pca_test <- prcomp(t(raw_filtered_norm_logcpm_prior_RUVcount_k2), center=TRUE, scale=TRUE)
pca_df_RUV <- as.data.frame(pca_test$x)
pca_df_RUV <- left_join(rownames_to_column(pca_df_RUV, "title"), metadata, by = "title")
pca_df_RUV$lightdark <- "dark"
pca_df_RUV$lightdark[pca_df_raw$time %in% c("01", "03", "05", "07", "09", "11")] <- "light"

ggplot(pca_df_RUV, aes(PC1, PC2, color = genotype, shape = lightdark, label = time)) +
  geom_point(size = 2) +
  #geom_text(hjust=0, vjust=0, color = "black", size = 2.5) +
  theme_classic() +
  labs(title = "PCA: timepoint & genotype (TMM + RUV, k=2, prior)",
       x = paste0("PC1 (", round(summary(pca_test)$importance[2,1] * 100, 1), "%)"),
       y = paste0("PC2 (", round(summary(pca_test)$importance[2,2] * 100, 1), "%)"))

```

```{r}
# lhycca1 vs wt! (RUN THE WHOLE CHUNK)

# PCA plot: CPM filtered raw count
raw_count_filtered_lhycca1 <- raw_count_filtered[, metadata[metadata$genotype == "wt" | metadata$genotype == "lhycca1", ]$title]
raw_count_filtered_lhycca1_filtered <- dplyr::filter(raw_count_filtered_lhycca1, rowSums(raw_count_filtered_lhycca1) > 0)  # Filter away genes with row sum = 0

pca_test_raw <- prcomp(t(raw_count_filtered_lhycca1_filtered), center=TRUE, scale=TRUE)
pca_df_raw <- as.data.frame(pca_test_raw$x)
pca_df_raw <- left_join(rownames_to_column(pca_df_raw, "title"), metadata, by = "title")
pca_df_raw$lightdark <- "dark"
pca_df_raw$lightdark[pca_df_raw$time %in% c("01", "03", "05", "07", "09", "11")] <- "light"

ggplot(pca_df_raw, aes(PC1, PC2, color = genotype, shape = lightdark, label = time)) +
  geom_point(size = 2) +
  #geom_text(hjust=0, vjust=0, color = "black", size = 2.5) +
  theme_classic() +
  labs(title = "PCA:  timepoint & genotype (raw count, lhycca1 vs wt)",
       x = paste0("PC1 (", round(summary(pca_test_raw)$importance[2,1] * 100, 1), "%)"),
       y = paste0("PC2 (", round(summary(pca_test_raw)$importance[2,2] * 100, 1), "%)"))


## PCA plot: TMM_log_cpm normalised
raw_filtered_norm_logcpm_prior_lhycca1 <- raw_filtered_norm_logcpm_prior[, metadata[metadata$genotype == "wt" | metadata$genotype == "lhycca1", ]$title] |> as.data.frame()
raw_filtered_norm_logcpm_prior_lhycca1_filtered <- dplyr::filter(raw_filtered_norm_logcpm_prior_lhycca1, rowSums(raw_filtered_norm_logcpm_prior_lhycca1) > 0)  # Filter away genes with row sum = 0

pca_test <- prcomp(t(raw_filtered_norm_logcpm_prior_lhycca1_filtered), center=TRUE, scale=TRUE)
pca_df_TMM <- as.data.frame(pca_test$x)
pca_df_TMM <- left_join(rownames_to_column(pca_df_TMM, "title"), metadata, by = "title")
pca_df_TMM$lightdark <- "dark"
pca_df_TMM$lightdark[pca_df_raw$time %in% c("01", "03", "05", "07", "09", "11")] <- "light"

ggplot(pca_df_TMM, aes(PC1, PC2, color = genotype, shape = lightdark, label = time)) +
  geom_point(size = 2) +
  #geom_text(hjust=0, vjust=0, color = "black", size = 2.5) +
  theme_classic() +
  labs(title = "PCA: timepoint & genotype (TMM normalised, lhycca1 vs wt)",
       x = paste0("PC1 (", round(summary(pca_test)$importance[2,1] * 100, 1), "%)"),
       y = paste0("PC2 (", round(summary(pca_test)$importance[2,2] * 100, 1), "%)"))


## PCA plot: RUV normalised, k=2, prior
raw_filtered_norm_logcpm_prior_RUVcount_k2_lhycca1 <- raw_filtered_norm_logcpm_prior_RUVcount_k2[, metadata[metadata$genotype == "wt" | metadata$genotype == "lhycca1", ]$title] |> as.data.frame()
raw_filtered_norm_logcpm_prior_RUVcount_k2_lhycca1_filtered <- dplyr::filter(raw_filtered_norm_logcpm_prior_RUVcount_k2_lhycca1, rowSums(raw_filtered_norm_logcpm_prior_RUVcount_k2_lhycca1) > 0)  # Filter away genes with row sum = 0

pca_test <- prcomp(t(raw_filtered_norm_logcpm_prior_RUVcount_k2_lhycca1), center=TRUE, scale=TRUE)
pca_df_RUV <- as.data.frame(pca_test$x)
pca_df_RUV <- left_join(rownames_to_column(pca_df_RUV, "title"), metadata, by = "title")
pca_df_RUV$lightdark <- "dark"
pca_df_RUV$lightdark[pca_df_raw$time %in% c("01", "03", "05", "07", "09", "11")] <- "light"

ggplot(pca_df_RUV, aes(PC1, PC2, color = genotype, shape = lightdark, label = time)) +
  geom_point(size = 2) +
  #geom_text(hjust=0, vjust=0, color = "black", size = 2.5) +
  theme_classic() +
  labs(title = "PCA: timepoint & genotype (TMM + RUV, k=2, prior, lhycca1 vs wt)",
       x = paste0("PC1 (", round(summary(pca_test)$importance[2,1] * 100, 1), "%)"),
       y = paste0("PC2 (", round(summary(pca_test)$importance[2,2] * 100, 1), "%)"))

```

```{r}
# prr7prr9 vs wt! (RUN THE WHOLE CHUNK)

# PCA plot: CPM filtered raw count
raw_count_filtered_prr7prr9 <- raw_count_filtered[, metadata[metadata$genotype == "wt" | metadata$genotype == "prr7prr9", ]$title]
raw_count_filtered_prr7prr9_filtered <- dplyr::filter(raw_count_filtered_prr7prr9, rowSums(raw_count_filtered_prr7prr9) > 0)  # Filter away genes with row sum = 0

pca_test_raw <- prcomp(t(raw_count_filtered_prr7prr9_filtered), center=TRUE, scale=TRUE)
pca_df_raw <- as.data.frame(pca_test_raw$x)
pca_df_raw <- left_join(rownames_to_column(pca_df_raw, "title"), metadata, by = "title")
pca_df_raw$lightdark <- "dark"
pca_df_raw$lightdark[pca_df_raw$time %in% c("01", "03", "05", "07", "09", "11")] <- "light"

ggplot(pca_df_raw, aes(PC1, PC2, color = genotype, shape = lightdark, label = time)) +
  geom_point(size = 2) +
  #geom_text(hjust=0, vjust=0, color = "black", size = 2.5) +
  theme_classic() +
  labs(title = "PCA:  timepoint & genotype (raw count, prr7prr9 vs wt)",
       x = paste0("PC1 (", round(summary(pca_test_raw)$importance[2,1] * 100, 1), "%)"),
       y = paste0("PC2 (", round(summary(pca_test_raw)$importance[2,2] * 100, 1), "%)"))


## PCA plot: TMM_log_cpm normalised
raw_filtered_norm_logcpm_prior_prr7prr9 <- raw_filtered_norm_logcpm_prior[, metadata[metadata$genotype == "wt" | metadata$genotype == "prr7prr9", ]$title] |> as.data.frame()
raw_filtered_norm_logcpm_prior_prr7prr9_filtered <- dplyr::filter(raw_filtered_norm_logcpm_prior_prr7prr9, rowSums(raw_filtered_norm_logcpm_prior_prr7prr9) > 0)  # Filter away genes with row sum = 0

pca_test <- prcomp(t(raw_filtered_norm_logcpm_prior_prr7prr9_filtered), center=TRUE, scale=TRUE)
pca_df_TMM <- as.data.frame(pca_test$x)
pca_df_TMM <- left_join(rownames_to_column(pca_df_TMM, "title"), metadata, by = "title")
pca_df_TMM$lightdark <- "dark"
pca_df_TMM$lightdark[pca_df_raw$time %in% c("01", "03", "05", "07", "09", "11")] <- "light"

ggplot(pca_df_TMM, aes(PC1, PC2, color = genotype, shape = lightdark, label = time)) +
  geom_point(size = 2) +
  #geom_text(hjust=0, vjust=0, color = "black", size = 2.5) +
  theme_classic() +
  labs(title = "PCA: timepoint & genotype (TMM normalised, prr7prr9 vs wt)",
       x = paste0("PC1 (", round(summary(pca_test)$importance[2,1] * 100, 1), "%)"),
       y = paste0("PC2 (", round(summary(pca_test)$importance[2,2] * 100, 1), "%)"))


## PCA plot: RUV normalised, k=2, prior
raw_filtered_norm_logcpm_prior_RUVcount_k2_prr7prr9 <- raw_filtered_norm_logcpm_prior_RUVcount_k2[, metadata[metadata$genotype == "wt" | metadata$genotype == "prr7prr9", ]$title] |> as.data.frame()
raw_filtered_norm_logcpm_prior_RUVcount_k2_prr7prr9_filtered <- dplyr::filter(raw_filtered_norm_logcpm_prior_RUVcount_k2_prr7prr9, rowSums(raw_filtered_norm_logcpm_prior_RUVcount_k2_prr7prr9) > 0)  # Filter away genes with row sum = 0

pca_test <- prcomp(t(raw_filtered_norm_logcpm_prior_RUVcount_k2_prr7prr9), center=TRUE, scale=TRUE)
pca_df_RUV <- as.data.frame(pca_test$x)
pca_df_RUV <- left_join(rownames_to_column(pca_df_RUV, "title"), metadata, by = "title")
pca_df_RUV$lightdark <- "dark"
pca_df_RUV$lightdark[pca_df_raw$time %in% c("01", "03", "05", "07", "09", "11")] <- "light"

ggplot(pca_df_RUV, aes(PC1, PC2, color = genotype, shape = lightdark, label = time)) +
  geom_point(size = 2) +
  #geom_text(hjust=0, vjust=0, color = "black", size = 2.5) +
  theme_classic() +
  labs(title = "PCA: timepoint & genotype (TMM + RUV, k=2, prior, prr7prr9 vs wt)",
       x = paste0("PC1 (", round(summary(pca_test)$importance[2,1] * 100, 1), "%)"),
       y = paste0("PC2 (", round(summary(pca_test)$importance[2,2] * 100, 1), "%)"))

```

```{r}
# toc1 vs wt! (RUN THE WHOLE CHUNK)

# PCA plot: CPM filtered raw count
raw_count_filtered_toc1 <- raw_count_filtered[, metadata[metadata$genotype == "wt" | metadata$genotype == "toc1", ]$title]
raw_count_filtered_toc1_filtered <- dplyr::filter(raw_count_filtered_toc1, rowSums(raw_count_filtered_toc1) > 0)  # Filter away genes with row sum = 0

pca_test_raw <- prcomp(t(raw_count_filtered_toc1_filtered), center=TRUE, scale=TRUE)
pca_df_raw <- as.data.frame(pca_test_raw$x)
pca_df_raw <- left_join(rownames_to_column(pca_df_raw, "title"), metadata, by = "title")
pca_df_raw$lightdark <- "dark"
pca_df_raw$lightdark[pca_df_raw$time %in% c("01", "03", "05", "07", "09", "11")] <- "light"

ggplot(pca_df_raw, aes(PC1, PC2, color = genotype, shape = lightdark, label = time)) +
  geom_point(size = 2) +
  #geom_text(hjust=0, vjust=0, color = "black", size = 2.5) +
  theme_classic() +
  labs(title = "PCA:  timepoint & genotype (raw count, toc1 vs wt)",
       x = paste0("PC1 (", round(summary(pca_test_raw)$importance[2,1] * 100, 1), "%)"),
       y = paste0("PC2 (", round(summary(pca_test_raw)$importance[2,2] * 100, 1), "%)"))


## PCA plot: TMM_log_cpm normalised
raw_filtered_norm_logcpm_prior_toc1 <- raw_filtered_norm_logcpm_prior[, metadata[metadata$genotype == "wt" | metadata$genotype == "toc1", ]$title] |> as.data.frame()
raw_filtered_norm_logcpm_prior_toc1_filtered <- dplyr::filter(raw_filtered_norm_logcpm_prior_toc1, rowSums(raw_filtered_norm_logcpm_prior_toc1) > 0)  # Filter away genes with row sum = 0

pca_test <- prcomp(t(raw_filtered_norm_logcpm_prior_toc1_filtered), center=TRUE, scale=TRUE)
pca_df_TMM <- as.data.frame(pca_test$x)
pca_df_TMM <- left_join(rownames_to_column(pca_df_TMM, "title"), metadata, by = "title")
pca_df_TMM$lightdark <- "dark"
pca_df_TMM$lightdark[pca_df_raw$time %in% c("01", "03", "05", "07", "09", "11")] <- "light"

ggplot(pca_df_TMM, aes(PC1, PC2, color = genotype, shape = lightdark, label = time)) +
  geom_point(size = 2) +
  #geom_text(hjust=0, vjust=0, color = "black", size = 2.5) +
  theme_classic() +
  labs(title = "PCA: timepoint & genotype (TMM normalised, toc1 vs wt)",
       x = paste0("PC1 (", round(summary(pca_test)$importance[2,1] * 100, 1), "%)"),
       y = paste0("PC2 (", round(summary(pca_test)$importance[2,2] * 100, 1), "%)"))


## PCA plot: RUV normalised, k=2, prior
raw_filtered_norm_logcpm_prior_RUVcount_k2_toc1 <- raw_filtered_norm_logcpm_prior_RUVcount_k2[, metadata[metadata$genotype == "wt" | metadata$genotype == "toc1", ]$title] |> as.data.frame()
raw_filtered_norm_logcpm_prior_RUVcount_k2_toc1_filtered <- dplyr::filter(raw_filtered_norm_logcpm_prior_RUVcount_k2_toc1, rowSums(raw_filtered_norm_logcpm_prior_RUVcount_k2_toc1) > 0)  # Filter away genes with row sum = 0

pca_test <- prcomp(t(raw_filtered_norm_logcpm_prior_RUVcount_k2_toc1), center=TRUE, scale=TRUE)
pca_df_RUV <- as.data.frame(pca_test$x)
pca_df_RUV <- left_join(rownames_to_column(pca_df_RUV, "title"), metadata, by = "title")
pca_df_RUV$lightdark <- "dark"
pca_df_RUV$lightdark[pca_df_raw$time %in% c("01", "03", "05", "07", "09", "11")] <- "light"

ggplot(pca_df_RUV, aes(PC1, PC2, color = genotype, shape = lightdark, label = time)) +
  geom_point(size = 2) +
  #geom_text(hjust=0, vjust=0, color = "black", size = 2.5) +
  theme_classic() +
  labs(title = "PCA: timepoint & genotype (TMM + RUV, k=2, prior, toc1 vs wt)",
       x = paste0("PC1 (", round(summary(pca_test)$importance[2,1] * 100, 1), "%)"),
       y = paste0("PC2 (", round(summary(pca_test)$importance[2,2] * 100, 1), "%)"))

```

Pretty good i think - not much variances to be corrected by RUV and improvement rather subtle but acceptable!


```{r}
# Limorhyde

period = 24
qvalRhyCutoff = 0.15
qvalDrCutoff = 0.1

### should i use unRUV matrix???? yes
y = raw_filtered_norm_logcpm_prior #_RUVcount_k2

metadata_lmr <- metadata |> dplyr::select(title, genotype, time) |> 
  dplyr::mutate(time = as.integer(time), cond = factor(genotype, levels = c("wt", "lhycca1", "prr7prr9", "toc1"))) |> 
  rownames_to_column("sample_alias")


metadata_lmr <- cbind(metadata_lmr, limorhyde(metadata_lmr$time, 'time_'), W1, W2) |> as.data.table()
```

```{r}
condNow <- "wt"

design = model.matrix(~ time_cos + time_sin, data = metadata_lmr[cond == condNow, ])
fit = lmFit(y[, metadata_lmr$cond == condNow], design)
fit = eBayes(fit, trend = TRUE)
rhyNow = data.table(topTable(fit, coef = 2:3, number = Inf), keep.rownames = TRUE)
setnames(rhyNow, 'rn', 'TAIR')
rhyNow[, cond := condNow]

rhyLimmaSummary = rhyNow[, .(P.Value = min(P.Value)), by = TAIR]
rhyLimmaSummary[, adj.P.Val := p.adjust(P.Value, method = 'BH')]
setorderv(rhyLimmaSummary, 'adj.P.Val')

rhyLimmaSummary[rhyLimmaSummary$adj.P.Val < 0.05, ]


# Extract coefficients from your limma fit
coefs = coef(fit)[, c("(Intercept)", "time_cos", "time_sin")]

# Compute (log2 *relative*) amplitude for each gene
amplitude_lmr = sqrt(coefs[, "time_cos"]^2 + coefs[, "time_sin"]^2)
mean = coefs[, "(Intercept)"]

# Compute (absolute) amplitude
amplitude_lmr_abs = 2^(mean + amplitude_lmr) - 2^mean

cbind(amplitude_lmr, coefs[, "(Intercept)"])

hist(amplitude_lmr, ylim = c(0,500))
```
four genotypes pooled - reasonable longer tail!


### JTK
```{r}
metadata
metadata$title[metadata$genotype == "wt"]
RUVs_output_prior$normalizedCounts[,metadata$title[metadata$genotype == "wt"]]

# log-TMM & RUV normalised matrix, control only
JTK_matrix_RUV <- as.data.frame(RUVs_output_prior$normalizedCounts[,metadata$title[metadata$genotype == "wt"]])
# JTK_matrix_RUV[rowSums(JTK_matrix_RUV)>0,]
# JTK_matrix_RUV[raw_count_WT$gene_name, ]

JTK_matrix_RUV <- JTK_matrix_RUV |> rownames_to_column('Probe')

write.table(JTK_matrix_RUV, file = 'JTK_Cycle/Mehta_JTK/Mehta_JTK_RUV_data.txt', sep = '\t', row.names = FALSE, col.names = TRUE, quote=FALSE)

write.table(JTK_matrix_RUV['Probe'], file = 'JTK_Cycle/Mehta_JTK/Mehta_JTK_RUV_annot.txt', sep = '/t', row.names = FALSE, col.names = TRUE, quote=FALSE)
```

--- running JTK cycle ---

```{r}
JTK_output_RUV <- read.table(file = 'JTK_Cycle/Mehta_JTK/JTK.Mehta_JTK_RUV_run.txt', header = TRUE ) |> dplyr::rename('TAIR' = 'Probe')

# Lable as rythmic & non-rhythmic using limorhyde's amplitude

JTK_output_RUV$amp_lmr <- amplitude_lmr[JTK_output_RUV$TAIR]
JTK_output_RUV$amp_lmr_abs <- amplitude_lmr_abs[JTK_output_RUV$TAIR]
JTK_output_RUV <- dplyr::relocate(JTK_output_RUV, amp_lmr, amp_lmr_abs, .after = AMP)

JTK_output_RUV$Rhythmicity_lmrAMP <- 'Ambiguous'
JTK_output_RUV$Rhythmicity_lmrAMP[JTK_output_RUV$BH.Q > 0.1 & JTK_output_RUV$amp_lmr < 0.58] <- 'Non-rhythmic'
JTK_output_RUV$Rhythmicity_lmrAMP[JTK_output_RUV$BH.Q < 0.05 & JTK_output_RUV$amp_lmr > 0.58] <- 'Rhythmic'
JTK_output_RUV <- dplyr::relocate(JTK_output_RUV, Rhythmicity_lmrAMP, .before = BH.Q)
table(JTK_output_RUV$Rhythmicity_lmrAMP)

# nonsense threshold for absolute values...
# JTK_output_RUV$Rhythmicity_lmrAMP_abs <- 'Non-rhythmic'
# JTK_output_RUV$Rhythmicity_lmrAMP_abs[JTK_output_RUV$BH.Q < 0.05 & JTK_output_RUV$amp_lmr_abs > 0.2] <- 'Rhythmic'
# JTK_output_RUV <- dplyr::relocate(JTK_output_RUV, Rhythmicity_lmrAMP_abs, .before = BH.Q)
# table(JTK_output_RUV$Rhythmicity_lmrAMP_abs)


```

   Ambiguous Non-rhythmic     Rhythmic 
        4842         8504         4276

WHAT IS THIS?????? cuz 2 hr interval????????

_____________________________________________________________________________

# a. lhycca1


```{r}
y_lhycca1 <- y[, metadata[metadata$genotype == "wt" | metadata$genotype == "lhycca1", ]$title]
metadata_lhycca1 <- metadata |> dplyr::filter(genotype %in% c("lhycca1", "wt"))
metadata_lmr_lhycca1 <- metadata_lmr |> 
  dplyr::filter(genotype %in% c("lhycca1", "wt")) |> 
  dplyr::mutate(cond = droplevels(cond))

```

```{r}
metadata_dt <- as.data.table(metadata_lhycca1)

geneIdsNow = c(deLimma_lhycca1$TAIR[1L:10L])  # requires later deLimma chunk ran
geneSymbolsNow = annotations[match(geneIdsNow, annotations$TAIR),]$SYMBOL
geneIdsNow <- geneIdsNow[!is.na(geneSymbolsNow)]
geneSymbolsNow <- geneSymbolsNow[!is.na(geneSymbolsNow)]

df = data.table(t(y[geneIdsNow, ]))
setnames(df, geneSymbolsNow)
df[, title := colnames(y[geneIdsNow, ])]

df = merge(df, metadata_dt[, .(title, genotype, time)], by = 'title')
df = melt(df, measure.vars = geneSymbolsNow, variable.name = 'symbol',
          value.name = 'expr')
df[, symbol := factor(symbol, geneSymbolsNow)]
df[, time := as.integer(time)]
df[, genotype := factor(genotype, levels = c("wt", "lhycca1"))]

ggplot(df) +
  facet_grid(symbol ~ genotype, scales = 'free_y') +
  geom_point(aes(x = time, y = expr, shape = genotype, color = symbol), size = 2) +
  labs(x = 'Zeitgeber time (h)', y = 'Expression (log)') +
  scale_shape(solid = FALSE, guide = 'none') +
  scale_color_brewer(type = 'qual', palette = 'Dark2', guide = 'none') +
  scale_x_continuous(limits = c(0, 24), breaks = seq(0, 24, 4))

deLimma_lhycca1
```

```{r}
geneIdsNow
```


```{r}
geneSymbolsNow = c("LHY", "CCA1", "PRR9", "PRR7", "PRR5", "TOC1")
geneIdsNow = annotations[match(geneSymbolsNow, annotations$SYMBOL),]$TAIR

df = data.table(t(y[geneIdsNow, ]))
setnames(df, geneSymbolsNow)
df[, title := colnames(y[geneIdsNow, ])]

df = merge(df, metadata_dt[, .(title, genotype, time)], by = 'title')
df = melt(df, measure.vars = geneSymbolsNow, variable.name = 'symbol',
          value.name = 'expr')
df[, symbol := factor(symbol, geneSymbolsNow)]
df[, time := as.integer(time)]
df[, genotype := factor(genotype, levels = c("wt", "lhycca1"))]

ggplot(df) +
  facet_grid(symbol ~ genotype, scales = 'free_y') +
  geom_point(aes(x = time, y = expr, shape = genotype, color = symbol), size = 2) +
  labs(x = 'Zeitgeber time (h)', y = 'Expression (log)') +
  scale_shape(solid = FALSE, guide = 'none') +
  scale_color_brewer(type = 'qual', palette = 'Dark2', guide = 'none') +
  scale_x_continuous(limits = c(0, 24), breaks = seq(0, 24, 4))

```

HUHHHHHHHHH?????????? why LHY & CCA1 still rhythmic - check KO method 

```{r}
geneSymbolsNow = c("ELF3", "ELF4", "LUX", "ACT2", "ACT8", "GAPC2")
geneIdsNow = annotations[match(geneSymbolsNow, annotations$SYMBOL),]$TAIR

df = data.table(t(y[geneIdsNow, ]))
setnames(df, geneSymbolsNow)
df[, title := colnames(y[geneIdsNow, ])]

df = merge(df, metadata_dt[, .(title, genotype, time)], by = 'title')
df = melt(df, measure.vars = geneSymbolsNow, variable.name = 'symbol',
          value.name = 'expr')
df[, symbol := factor(symbol, geneSymbolsNow)]
df[, time := as.integer(time)]
df[, genotype := factor(genotype, levels = c("wt", "lhycca1"))]

ggplot(df) +
  facet_grid(symbol ~ genotype, scales = 'free_y') +
  geom_point(aes(x = time, y = expr, shape = genotype, color = symbol), size = 2) +
  labs(x = 'Zeitgeber time (h)', y = 'Expression (log)') +
  scale_shape(solid = FALSE, guide = 'none') +
  scale_color_brewer(type = 'qual', palette = 'Dark2', guide = 'none') +
  scale_x_continuous(limits = c(0, 24), breaks = seq(0, 24, 4))
```


### Limorhyde x Limma DE analysis

```{r}
# with interaction

design = model.matrix(~ cond + time_cos + time_sin + cond:time_cos + cond:time_sin + W1 + W2, data = metadata_lmr_lhycca1)
colnames(design)

qr(design)$rank  # check if W1/2 are colinear with cond
cor(metadata_lmr_lhycca1[, c("W1", "W2", "time_cos", "time_sin")])  # check if W1/2 is eating away time information


fit = lmFit(y_lhycca1, design)
fit = eBayes(fit, trend = TRUE)
deLimma_lhycca1 = data.table(topTable(fit, coef = 2, number = Inf), keep.rownames = TRUE)
setnames(deLimma_lhycca1, 'rn', 'TAIR')

# record mean level of WT and KO!
mean_wt <- fit$coefficients[,1]
mean_ko <- fit$coefficients[,1] + fit$coefficients[,2]

deLimma_lhycca1[, adj.P.Val := p.adjust(P.Value, method = 'BH')]
setorderv(deLimma_lhycca1, 'adj.P.Val')

deLimma_lhycca1

deLimma_lhycca1$diffExpressed <- 'NO'
deLimma_lhycca1$diffExpressed[deLimma_lhycca1$logFC>1 & deLimma_lhycca1$adj.P.Val<0.05] <- 'UP'
deLimma_lhycca1$diffExpressed[deLimma_lhycca1$logFC<(-1) & deLimma_lhycca1$adj.P.Val<0.05] <- 'DOWN'
table(deLimma_lhycca1$diffExpressed)
```

 DOWN    NO    UP 
  110 17351   161 
  
really not a lot lol


## Rhythmicity vs DE

```{r}
JTK_output_RUV

deLimma_lhycca1

JTK_deLimma_lhycca1 <- left_join(deLimma_lhycca1, JTK_output_RUV, by = "TAIR") |> 
  dplyr::mutate(Rhythmicity_fac = factor(Rhythmicity_lmrAMP)) |>
  dplyr::mutate(diffExpressed_fac = factor(diffExpressed)) |> 
  dplyr::relocate(Rhythmicity_fac, .after = "TAIR") |> 
  dplyr::relocate(diffExpressed_fac, .after = "TAIR")

JTK_deLimma_lhycca1$DE <- JTK_deLimma_lhycca1$diffExpressed != "NO"
lhycca1_table <- table(JTK_deLimma_lhycca1[, c('DE', 'Rhythmicity_lmrAMP')])
lhycca1_table[,2:3]
chisq.test(lhycca1_table[,2:3])
table(JTK_deLimma_lhycca1[, c('diffExpressed', 'Rhythmicity_lmrAMP')])
```
             Rhythmicity_lmrAMP
diffExpressed Ambiguous Non-rhythmic Rhythmic
         DOWN        20           36       54
         NO        4793         8408     4150
         UP          29           60       72

	Pearson's Chi-squared test with Yates' continuity correction

data:  lhycca1_table[, 2:3]
X-squared = 54.022, df = 1, p-value = 1.982e-13


```{r}
table <- table(JTK_deLimma_lhycca1[, c('diffExpressed', 'Rhythmicity_lmrAMP')])

DEratio <- (table[1, ] + table[3,]) / colSums(table)
DEratio
```


   Ambiguous Non-rhythmic     Rhythmic 
  0.01011979   0.01128881   0.02946679 
  
  


```{r}
volcano_data_rhythmic <- JTK_deLimma_lhycca1 |> dplyr::filter(Rhythmicity_lmrAMP=='Rhythmic')
volcano_data_nonrhythmic <- JTK_deLimma_lhycca1 |> dplyr::filter(Rhythmicity_lmrAMP=='Non-rhythmic')

ggplot(data=volcano_data_rhythmic, aes(x=logFC, y=-log10(adj.P.Val), col = diffExpressed_fac)) +
  geom_vline(xintercept = c(-0.8, 0.8), col='gray', linetype='dashed') +
  geom_hline(yintercept=c(-log10(0.05)), col='gray', linetype='dashed') +
  geom_point(size=1) +
  scale_color_manual(values = c("#00AFBB", "grey", "#FC4E07"), 
                     labels = c("Downregulated", "Not significant", "Upregulated")) +
  ggtitle("Rhythmic genes") +
  xlim(-5, 5) +
  ylim(-1, 50) #+
  # geom_label_repel(data = dplyr::filter(volcano_data_rhythmic, abs(logFC) > 1.8 | -log10(adj.P.Val) > 25), aes(, label = SYMBOL), hjust=0, vjust=0, color = "black", size = 2.5)
  

ggplot(data=volcano_data_nonrhythmic, aes(x=logFC, y=-log10(adj.P.Val
), col = diffExpressed_fac)) +
  geom_vline(xintercept = c(-0.8, 0.8), col='gray', linetype='dashed') +
  geom_hline(yintercept=c(-log10(0.05)), col='gray', linetype='dashed') +
  geom_point(size=1) +
  scale_color_manual(values = c("#00AFBB", "grey", "#FC4E07"), 
                     labels = c("Downregulated", "Not significant", "Upregulated")) +
  ggtitle("Non-rhythmic genes") +
  xlim(-5, 5) +
  ylim(-1, 50) #+
  # geom_label_repel(data = dplyr::filter(volcano_data_nonrhythmic, logFC < -2 | logFC > 1.2 | -log10(adj.P.Val) > 35), aes(, label = gene_name), hjust=0, vjust=0, color = "black", size = 2.5)

```

### Annotation of DE-rhythmic genes

```{r}
DE_rhythmic_genes_lhycca1 <- JTK_deLimma_lhycca1[JTK_deLimma_lhycca1$BH.Q < 0.05 & 
                                   JTK_deLimma_lhycca1$amp_lmr > 0.58 & 
                                   abs(JTK_deLimma_lhycca1$logFC) > 1.0 & 
                                   JTK_deLimma_lhycca1$adj.P.Val < 0.05]

DE_rhythmic_genes_annot_lhycca1 <- right_join(annotations[match(DE_rhythmic_genes_lhycca1$TAIR, annotations$TAIR),], DE_rhythmic_genes_lhycca1) |>
  arrange(desc(abs(logFC))) |> as.matrix()

write.table(DE_rhythmic_genes_annot_lhycca1, file = "../../results/Mehta_results/Mehta_DE_rhythmic_genes_annot_lhycca1.tsv", sep = "\t", row.names = FALSE)

sum(!rownames(y) %in% annotations$TAIR)
sum(!DE_rhythmic_genes_lhycca1$TAIR %in% annotations$TAIR)
length(rownames(y))

```

????? why so many NA genes... (ok only 2 are not mapped, others are mapped but many don't have SYMBOL/genename)


```{r}
JTK_deLimma_lhycca1
ggplot(JTK_deLimma_lhycca1, aes(x = amp_lmr, y = logFC)) +
  geom_point(alpha = 0.5) +
  theme_minimal() +
  xlab("Rhythmic amplitude (from Limorhyde)") +
  ylab("log2 Fold Change (DE)") +
  ggtitle("DE vs rhythmic amplitude")


cor(JTK_deLimma_lhycca1$logFC, JTK_deLimma_lhycca1$amp_lmr)
```

```{r}
ggplot(JTK_deLimma_lhycca1[JTK_deLimma_lhycca1$Rhythmicity_fac == "Rhythmic", ], aes(x = amp_lmr, y = logFC, color = Rhythmicity_fac)) +
  geom_point(alpha = 0.6) +
  scale_color_manual(values = c(
    "Non-rhythmic" = "transparent",
    "Rhythmic" = "royalblue"
  )) +
  theme_minimal(base_size = 14) +
  xlab("Rhythmic absolute amplitude") +
  ylab("log2 Fold Change") +
  xlim(0, 3) +
  ggtitle("logFC vs rhythmic amplitude by rhythmicity") +
  theme(
    legend.title = element_text(face = "bold"),
    legend.position = "right"
  ) # +
  # geom_label_repel(data = dplyr::filter(JTK_deLimma_lhycca1, (amp_lmr > 1.3 | abs(logFC) > 2.5) & amp_lmr > 0.3), aes(, label = gene_name), 
  #           hjust=0, vjust=0, color = "black", size = 2)
```

Seems not as obvious trend...


### 3D plot!

```{r}

AMP_threshold_list <- seq(0, 2, by=0.05)
logFC_threshold_list <- seq(0, 5, by=0.1)


threshold_result <- data.frame(
  AMP = NA,
  logFC = NA,
  ratio = NA,
  rhythmic_DE = NA,
  DE = NA,
  rhythmic = NA,
  log_rhythmic_DE = NA
)


# 3d threshold plot data
for (logFC_threshold in logFC_threshold_list){
  for (AMP_threshold in AMP_threshold_list){
    N_rhythmic_DE <- nrow(JTK_deLimma_lhycca1[abs(JTK_deLimma_lhycca1$logFC) > logFC_threshold & JTK_deLimma_lhycca1$adj.P.Val < 0.05 & JTK_deLimma_lhycca1$BH.Q < 0.05 & JTK_deLimma_lhycca1$amp_lmr > AMP_threshold])
    N_rhythmic <- nrow(JTK_deLimma_lhycca1[JTK_deLimma_lhycca1$BH.Q < 0.05 & JTK_deLimma_lhycca1$amp_lmr > AMP_threshold])
    # print(c(AMP_threshold, N_rhythmic))
    N_DE <- nrow(JTK_deLimma_lhycca1[abs(JTK_deLimma_lhycca1$logFC) > logFC_threshold & JTK_deLimma_lhycca1$adj.P.Val < 0.05])
    
    if (N_rhythmic == 0){
      DE_ratio = NA
    } 
    else{
      DE_ratio <- N_rhythmic_DE/N_rhythmic
    }
    threshold_result <- rbind(threshold_result, c(AMP = AMP_threshold, logFC = logFC_threshold, ratio = DE_ratio, rhythmic_DE = N_rhythmic_DE, DE = N_DE, rhythmic = N_rhythmic, log10_rhythmic_DE = log10(N_rhythmic_DE)))
  }
}

threshold_result <- drop_na(threshold_result)
threshold_result_finite <- threshold_result[is.finite(threshold_result$log_rhythmic_DE), ]


threshold_result[threshold_result$AMP == 0.55 & threshold_result$logFC == 1.0, ]
```
```{r}
### Relative 3D plot (log-count coloured)

fig <- plot_ly(threshold_result_finite, x = ~logFC, y = ~AMP, z = ~ratio, color = ~log_rhythmic_DE, size = I(30), colors = rainbow(5))
fig <- fig %>% add_markers()
fig <- fig %>% layout(title = list(text="lhycca1: Relative DE ratio for various logFC and relative AMP thresholds", y = 0.95, x = 0.5, xanchor = 'center', yanchor = "top"),
  scene = list(xaxis = list(title = 'log2 Fold Change <br> Magnitude',
                                                range = c(0, 2.5)),
               yaxis = list(title = 'Amplitude (relative)',
                                                range = c(0, 1.0)),
               zaxis = list(title = 'Ratio of DE genes : Rhythmic genes',
                                                range = c(0, 0.9)),
               aspectmode = "cube"))

colorbar(fig, title = "# rhythmic DE genes", tickmode = "array", tickvals = c(0,1,2,3), ticktext = c(1, 10, 100, 1000))


```
So smooth and subtle??????


__________________________________________________

# b. prr7prr9


```{r}
y_prr7prr9 <- y[, metadata[metadata$genotype == "wt" | metadata$genotype == "prr7prr9", ]$title]
metadata_prr7prr9 <- metadata |> dplyr::filter(genotype %in% c("prr7prr9", "wt"))
metadata_lmr_prr7prr9 <- metadata_lmr |> 
  dplyr::filter(genotype %in% c("prr7prr9", "wt")) |> 
  dplyr::mutate(cond = droplevels(cond))

```

```{r}
metadata_dt <- as.data.table(metadata_prr7prr9)

geneIdsNow = c(deLimma_prr7prr9$TAIR[1L:10L])  # requires later deLimma chunk ran
geneSymbolsNow = annotations[match(geneIdsNow, annotations$TAIR),]$SYMBOL
geneIdsNow <- geneIdsNow[!is.na(geneSymbolsNow)]
geneSymbolsNow <- geneSymbolsNow[!is.na(geneSymbolsNow)]

df = data.table(t(y[geneIdsNow, ]))
setnames(df, geneSymbolsNow)
df[, title := colnames(y[geneIdsNow, ])]

df = merge(df, metadata_dt[, .(title, genotype, time)], by = 'title')
df = melt(df, measure.vars = geneSymbolsNow, variable.name = 'symbol',
          value.name = 'expr')
df[, symbol := factor(symbol, geneSymbolsNow)]
df[, time := as.integer(time)]
df[, genotype := factor(genotype, levels = c("wt", "prr7prr9"))]

ggplot(df) +
  facet_grid(symbol ~ genotype, scales = 'free_y') +
  geom_point(aes(x = time, y = expr, shape = genotype, color = symbol), size = 2) +
  labs(x = 'Zeitgeber time (h)', y = 'Expression (log)') +
  scale_shape(solid = FALSE, guide = 'none') +
  scale_color_brewer(type = 'qual', palette = 'Dark2', guide = 'none') +
  scale_x_continuous(limits = c(0, 24), breaks = seq(0, 24, 4))

deLimma_prr7prr9
```

```{r}
geneIdsNow
```


```{r}
geneSymbolsNow = c("LHY", "CCA1", "PRR9", "PRR7", "PRR5", "TOC1")
geneIdsNow = annotations[match(geneSymbolsNow, annotations$SYMBOL),]$TAIR

df = data.table(t(y[geneIdsNow, ]))
setnames(df, geneSymbolsNow)
df[, title := colnames(y[geneIdsNow, ])]

df = merge(df, metadata_dt[, .(title, genotype, time)], by = 'title')
df = melt(df, measure.vars = geneSymbolsNow, variable.name = 'symbol',
          value.name = 'expr')
df[, symbol := factor(symbol, geneSymbolsNow)]
df[, time := as.integer(time)]
df[, genotype := factor(genotype, levels = c("wt", "prr7prr9"))]

ggplot(df) +
  facet_grid(symbol ~ genotype, scales = 'free_y') +
  geom_point(aes(x = time, y = expr, shape = genotype, color = symbol), size = 2) +
  labs(x = 'Zeitgeber time (h)', y = 'Expression (log)') +
  scale_shape(solid = FALSE, guide = 'none') +
  scale_color_brewer(type = 'qual', palette = 'Dark2', guide = 'none') +
  scale_x_continuous(limits = c(0, 24), breaks = seq(0, 24, 4))

```

HUHHHHHHHHH?????????? why LHY & CCA1 still rhythmic - check KO method 

```{r}
geneSymbolsNow = c("ELF3", "ELF4", "LUX", "ACT2", "ACT8", "GAPC2")
geneIdsNow = annotations[match(geneSymbolsNow, annotations$SYMBOL),]$TAIR

df = data.table(t(y[geneIdsNow, ]))
setnames(df, geneSymbolsNow)
df[, title := colnames(y[geneIdsNow, ])]

df = merge(df, metadata_dt[, .(title, genotype, time)], by = 'title')
df = melt(df, measure.vars = geneSymbolsNow, variable.name = 'symbol',
          value.name = 'expr')
df[, symbol := factor(symbol, geneSymbolsNow)]
df[, time := as.integer(time)]
df[, genotype := factor(genotype, levels = c("wt", "prr7prr9"))]

ggplot(df) +
  facet_grid(symbol ~ genotype, scales = 'free_y') +
  geom_point(aes(x = time, y = expr, shape = genotype, color = symbol), size = 2) +
  labs(x = 'Zeitgeber time (h)', y = 'Expression (log)') +
  scale_shape(solid = FALSE, guide = 'none') +
  scale_color_brewer(type = 'qual', palette = 'Dark2', guide = 'none') +
  scale_x_continuous(limits = c(0, 24), breaks = seq(0, 24, 4))
```

Is the oscillation of ACT8 & GAPC2 normal?


### Limorhyde x Limma DE analysis

```{r}
# with interaction

design = model.matrix(~ cond + time_cos + time_sin + cond:time_cos + cond:time_sin + W1 + W2, data = metadata_lmr_prr7prr9)
colnames(design)

qr(design)$rank  # check if W1/2 are colinear with cond
cor(metadata_lmr_prr7prr9[, c("W1", "W2", "time_cos", "time_sin")])  # check if W1/2 is eating away time information


fit = lmFit(y_prr7prr9, design)
fit = eBayes(fit, trend = TRUE)
deLimma_prr7prr9 = data.table(topTable(fit, coef = 2, number = Inf), keep.rownames = TRUE)
setnames(deLimma_prr7prr9, 'rn', 'TAIR')

# record mean level of WT and KO!
mean_wt <- fit$coefficients[,1]
mean_ko <- fit$coefficients[,1] + fit$coefficients[,2]

deLimma_prr7prr9[, adj.P.Val := p.adjust(P.Value, method = 'BH')]
setorderv(deLimma_prr7prr9, 'adj.P.Val')

deLimma_prr7prr9

deLimma_prr7prr9$diffExpressed <- 'NO'
deLimma_prr7prr9$diffExpressed[deLimma_prr7prr9$logFC>1 & deLimma_prr7prr9$adj.P.Val<0.05] <- 'UP'
deLimma_prr7prr9$diffExpressed[deLimma_prr7prr9$logFC<(-1) & deLimma_prr7prr9$adj.P.Val<0.05] <- 'DOWN'
table(deLimma_prr7prr9$diffExpressed)
```

 DOWN    NO    UP 
  237 17209   176 
  
really not a lot too


## Rhythmicity vs DE

```{r}
JTK_output_RUV

deLimma_prr7prr9

JTK_deLimma_prr7prr9 <- left_join(deLimma_prr7prr9, JTK_output_RUV, by = "TAIR") |> 
  dplyr::mutate(Rhythmicity_fac = factor(Rhythmicity_lmrAMP)) |>
  dplyr::mutate(diffExpressed_fac = factor(diffExpressed)) |> 
  dplyr::relocate(Rhythmicity_fac, .after = "TAIR") |> 
  dplyr::relocate(diffExpressed_fac, .after = "TAIR")

JTK_deLimma_prr7prr9$DE <- JTK_deLimma_prr7prr9$diffExpressed != "NO"
prr7prr9_table <- table(JTK_deLimma_prr7prr9[, c('DE', 'Rhythmicity_lmrAMP')])
prr7prr9_table[,2:3]
chisq.test(prr7prr9_table[,2:3])
table(JTK_deLimma_prr7prr9[, c('diffExpressed', 'Rhythmicity_lmrAMP')])
```
       Rhythmicity_lmrAMP
DE      Non-rhythmic Rhythmic
  FALSE         8416     4006
  TRUE            88      270

	Pearson's Chi-squared test with Yates' continuity correction

data:  prr7prr9_table[, 2:3]
X-squared = 289.34, df = 1, p-value < 2.2e-16

             Rhythmicity_lmrAMP
diffExpressed Ambiguous Non-rhythmic Rhythmic
         DOWN        48           66      123
         NO        4787         8416     4006
         UP           7           22      147

```{r}
table <- table(JTK_deLimma_prr7prr9[, c('diffExpressed', 'Rhythmicity_lmrAMP')])

DEratio <- (table[1, ] + table[3,]) / colSums(table)
DEratio
```


   Ambiguous Non-rhythmic     Rhythmic 
  0.01135894   0.01034807   0.06314312 


```{r}
volcano_data_rhythmic <- JTK_deLimma_prr7prr9 |> dplyr::filter(Rhythmicity_lmrAMP=='Rhythmic')
volcano_data_nonrhythmic <- JTK_deLimma_prr7prr9 |> dplyr::filter(Rhythmicity_lmrAMP=='Non-rhythmic')

ggplot(data=volcano_data_rhythmic, aes(x=logFC, y=-log10(adj.P.Val), col = diffExpressed_fac)) +
  geom_vline(xintercept = c(-0.8, 0.8), col='gray', linetype='dashed') +
  geom_hline(yintercept=c(-log10(0.05)), col='gray', linetype='dashed') +
  geom_point(size=1) +
  scale_color_manual(values = c("#00AFBB", "grey", "#FC4E07"), 
                     labels = c("Downregulated", "Not significant", "Upregulated")) +
  ggtitle("Rhythmic genes") +
  xlim(-5, 5) +
  ylim(-1, 50) #+
  # geom_label_repel(data = dplyr::filter(volcano_data_rhythmic, abs(logFC) > 1.8 | -log10(adj.P.Val) > 25), aes(, label = SYMBOL), hjust=0, vjust=0, color = "black", size = 2.5)
  

ggplot(data=volcano_data_nonrhythmic, aes(x=logFC, y=-log10(adj.P.Val
), col = diffExpressed_fac)) +
  geom_vline(xintercept = c(-0.8, 0.8), col='gray', linetype='dashed') +
  geom_hline(yintercept=c(-log10(0.05)), col='gray', linetype='dashed') +
  geom_point(size=1) +
  scale_color_manual(values = c("#00AFBB", "grey", "#FC4E07"), 
                     labels = c("Downregulated", "Not significant", "Upregulated")) +
  ggtitle("Non-rhythmic genes") +
  xlim(-5, 5) +
  ylim(-1, 50) #+
  # geom_label_repel(data = dplyr::filter(volcano_data_nonrhythmic, logFC < -2 | logFC > 1.2 | -log10(adj.P.Val) > 35), aes(, label = gene_name), hjust=0, vjust=0, color = "black", size = 2.5)

```

### Annotation of DE-rhythmic genes

```{r}
DE_rhythmic_genes_prr7prr9 <- JTK_deLimma_prr7prr9[JTK_deLimma_prr7prr9$BH.Q < 0.05 & 
                                   JTK_deLimma_prr7prr9$amp_lmr > 0.58 & 
                                   abs(JTK_deLimma_prr7prr9$logFC) > 1.0 & 
                                   JTK_deLimma_prr7prr9$adj.P.Val < 0.05]

DE_rhythmic_genes_annot_prr7prr9 <- right_join(annotations[match(DE_rhythmic_genes_prr7prr9$TAIR, annotations$TAIR),], DE_rhythmic_genes_prr7prr9) |>
  arrange(desc(abs(logFC))) |> as.matrix()

write.table(DE_rhythmic_genes_annot_prr7prr9, file = "../../results/Mehta_results/Mehta_DE_rhythmic_genes_annot_prr7prr9.tsv", sep = "\t", row.names = FALSE)

sum(!rownames(y) %in% annotations$TAIR)
sum(!DE_rhythmic_genes_prr7prr9$TAIR %in% annotations$TAIR)
length(rownames(y))

```

????? why so many NA genes... (ok only 2 are not mapped, others are mapped but many don't have SYMBOL/genename)


```{r}
JTK_deLimma_prr7prr9
ggplot(JTK_deLimma_prr7prr9, aes(x = amp_lmr, y = logFC)) +
  geom_point(alpha = 0.5) +
  theme_minimal() +
  xlab("Rhythmic amplitude (from Limorhyde)") +
  ylab("log2 Fold Change (DE)") +
  ggtitle("DE vs rhythmic amplitude")


cor(JTK_deLimma_prr7prr9$logFC, JTK_deLimma_prr7prr9$amp_lmr)
```

```{r}
ggplot(JTK_deLimma_prr7prr9[JTK_deLimma_prr7prr9$Rhythmicity_fac == "Rhythmic", ], aes(x = amp_lmr, y = logFC, color = Rhythmicity_fac)) +
  geom_point(alpha = 0.6) +
  scale_color_manual(values = c(
    "Non-rhythmic" = "transparent",
    "Rhythmic" = "royalblue"
  )) +
  theme_minimal(base_size = 14) +
  xlab("Rhythmic absolute amplitude") +
  ylab("log2 Fold Change") +
  xlim(0, 3) +
  ggtitle("logFC vs rhythmic amplitude by rhythmicity") +
  theme(
    legend.title = element_text(face = "bold"),
    legend.position = "right"
  ) # +
  # geom_label_repel(data = dplyr::filter(JTK_deLimma_prr7prr9, (amp_lmr > 1.3 | abs(logFC) > 2.5) & amp_lmr > 0.3), aes(, label = gene_name), 
  #           hjust=0, vjust=0, color = "black", size = 2)
```

Seems not as obvious trend...


### 3D plot!

```{r}

AMP_threshold_list <- seq(0, 2, by=0.05)
logFC_threshold_list <- seq(0, 5, by=0.1)


threshold_result <- data.frame(
  AMP = NA,
  logFC = NA,
  ratio = NA,
  rhythmic_DE = NA,
  DE = NA,
  rhythmic = NA,
  log_rhythmic_DE = NA
)


# 3d threshold plot data
for (logFC_threshold in logFC_threshold_list){
  for (AMP_threshold in AMP_threshold_list){
    N_rhythmic_DE <- nrow(JTK_deLimma_prr7prr9[abs(JTK_deLimma_prr7prr9$logFC) > logFC_threshold & JTK_deLimma_prr7prr9$adj.P.Val < 0.05 & JTK_deLimma_prr7prr9$BH.Q < 0.05 & JTK_deLimma_prr7prr9$amp_lmr > AMP_threshold])
    N_rhythmic <- nrow(JTK_deLimma_prr7prr9[JTK_deLimma_prr7prr9$BH.Q < 0.05 & JTK_deLimma_prr7prr9$amp_lmr > AMP_threshold])
    # print(c(AMP_threshold, N_rhythmic))
    N_DE <- nrow(JTK_deLimma_prr7prr9[abs(JTK_deLimma_prr7prr9$logFC) > logFC_threshold & JTK_deLimma_prr7prr9$adj.P.Val < 0.05])
    
    if (N_rhythmic == 0){
      DE_ratio = NA
    } 
    else{
      DE_ratio <- N_rhythmic_DE/N_rhythmic
    }
    threshold_result <- rbind(threshold_result, c(AMP = AMP_threshold, logFC = logFC_threshold, ratio = DE_ratio, rhythmic_DE = N_rhythmic_DE, DE = N_DE, rhythmic = N_rhythmic, log10_rhythmic_DE = log10(N_rhythmic_DE)))
  }
}

threshold_result <- drop_na(threshold_result)
threshold_result_finite <- threshold_result[is.finite(threshold_result$log_rhythmic_DE), ]


threshold_result[threshold_result$AMP == 0.55 & threshold_result$logFC == 1.0, ]
```
```{r}
### Relative 3D plot (log-count coloured)

fig <- plot_ly(threshold_result_finite, x = ~logFC, y = ~AMP, z = ~ratio, color = ~log_rhythmic_DE, size = I(30), colors = rainbow(5))
fig <- fig %>% add_markers()
fig <- fig %>% layout(title = list(text="prr7prr9: Relative DE ratio for various logFC and relative AMP thresholds", y = 0.95, x = 0.5, xanchor = 'center', yanchor = "top"),
  scene = list(xaxis = list(title = 'log2 Fold Change <br> Magnitude',
                                                range = c(0, 2.5)),
               yaxis = list(title = 'Amplitude (relative)',
                                                range = c(0, 1.0)),
               zaxis = list(title = 'Ratio of DE genes : Rhythmic genes',
                                                range = c(0, 0.9)),
               aspectmode = "cube"))

colorbar(fig, title = "# rhythmic DE genes", tickmode = "array", tickvals = c(0,1,2,3), ticktext = c(1, 10, 100, 1000))


```




__________________________________________________

# c. toc1


```{r}
y_toc1 <- y[, metadata[metadata$genotype == "wt" | metadata$genotype == "toc1", ]$title]
metadata_toc1 <- metadata |> dplyr::filter(genotype %in% c("toc1", "wt"))
metadata_lmr_toc1 <- metadata_lmr |> 
  dplyr::filter(genotype %in% c("toc1", "wt")) |> 
  dplyr::mutate(cond = droplevels(cond))

```

```{r}
metadata_dt <- as.data.table(metadata_toc1)

geneIdsNow = c(deLimma_toc1$TAIR[1L:10L])  # requires later deLimma chunk ran
geneSymbolsNow = annotations[match(geneIdsNow, annotations$TAIR),]$SYMBOL
geneIdsNow <- geneIdsNow[!is.na(geneSymbolsNow)]
geneSymbolsNow <- geneSymbolsNow[!is.na(geneSymbolsNow)]

df = data.table(t(y[geneIdsNow, ]))
setnames(df, geneSymbolsNow)
df[, title := colnames(y[geneIdsNow, ])]

df = merge(df, metadata_dt[, .(title, genotype, time)], by = 'title')
df = melt(df, measure.vars = geneSymbolsNow, variable.name = 'symbol',
          value.name = 'expr')
df[, symbol := factor(symbol, geneSymbolsNow)]
df[, time := as.integer(time)]
df[, genotype := factor(genotype, levels = c("wt", "toc1"))]

ggplot(df) +
  facet_grid(symbol ~ genotype, scales = 'free_y') +
  geom_point(aes(x = time, y = expr, shape = genotype, color = symbol), size = 2) +
  labs(x = 'Zeitgeber time (h)', y = 'Expression (log)') +
  scale_shape(solid = FALSE, guide = 'none') +
  scale_color_brewer(type = 'qual', palette = 'Dark2', guide = 'none') +
  scale_x_continuous(limits = c(0, 24), breaks = seq(0, 24, 4))

deLimma_toc1
```

```{r}
geneIdsNow
```


```{r}
geneSymbolsNow = c("LHY", "CCA1", "PRR9", "PRR7", "PRR5", "TOC1")
geneIdsNow = annotations[match(geneSymbolsNow, annotations$SYMBOL),]$TAIR

df = data.table(t(y[geneIdsNow, ]))
setnames(df, geneSymbolsNow)
df[, title := colnames(y[geneIdsNow, ])]

df = merge(df, metadata_dt[, .(title, genotype, time)], by = 'title')
df = melt(df, measure.vars = geneSymbolsNow, variable.name = 'symbol',
          value.name = 'expr')
df[, symbol := factor(symbol, geneSymbolsNow)]
df[, time := as.integer(time)]
df[, genotype := factor(genotype, levels = c("wt", "toc1"))]

ggplot(df) +
  facet_grid(symbol ~ genotype, scales = 'free_y') +
  geom_point(aes(x = time, y = expr, shape = genotype, color = symbol), size = 2) +
  labs(x = 'Zeitgeber time (h)', y = 'Expression (log)') +
  scale_shape(solid = FALSE, guide = 'none') +
  scale_color_brewer(type = 'qual', palette = 'Dark2', guide = 'none') +
  scale_x_continuous(limits = c(0, 24), breaks = seq(0, 24, 4))

```

HUHHHHHHHHH?????????? why TOC1 still rhythmic - check KO method 

```{r}
geneSymbolsNow = c("ELF3", "ELF4", "LUX", "ACT2", "ACT8", "GAPC2")
geneIdsNow = annotations[match(geneSymbolsNow, annotations$SYMBOL),]$TAIR

df = data.table(t(y[geneIdsNow, ]))
setnames(df, geneSymbolsNow)
df[, title := colnames(y[geneIdsNow, ])]

df = merge(df, metadata_dt[, .(title, genotype, time)], by = 'title')
df = melt(df, measure.vars = geneSymbolsNow, variable.name = 'symbol',
          value.name = 'expr')
df[, symbol := factor(symbol, geneSymbolsNow)]
df[, time := as.integer(time)]
df[, genotype := factor(genotype, levels = c("wt", "toc1"))]

ggplot(df) +
  facet_grid(symbol ~ genotype, scales = 'free_y') +
  geom_point(aes(x = time, y = expr, shape = genotype, color = symbol), size = 2) +
  labs(x = 'Zeitgeber time (h)', y = 'Expression (log)') +
  scale_shape(solid = FALSE, guide = 'none') +
  scale_color_brewer(type = 'qual', palette = 'Dark2', guide = 'none') +
  scale_x_continuous(limits = c(0, 24), breaks = seq(0, 24, 4))
```

Is the oscillation of ACT8 & GAPC2 normal?


### Limorhyde x Limma DE analysis

```{r}
# with interaction

design = model.matrix(~ cond + time_cos + time_sin + cond:time_cos + cond:time_sin + W1 + W2, data = metadata_lmr_toc1)
colnames(design)

qr(design)$rank  # check if W1/2 are colinear with cond
cor(metadata_lmr_toc1[, c("W1", "W2", "time_cos", "time_sin")])  # check if W1/2 is eating away time information


fit = lmFit(y_toc1, design)
fit = eBayes(fit, trend = TRUE)
deLimma_toc1 = data.table(topTable(fit, coef = 2, number = Inf), keep.rownames = TRUE)
setnames(deLimma_toc1, 'rn', 'TAIR')

# record mean level of WT and KO!
mean_wt <- fit$coefficients[,1]
mean_ko <- fit$coefficients[,1] + fit$coefficients[,2]

deLimma_toc1[, adj.P.Val := p.adjust(P.Value, method = 'BH')]
setorderv(deLimma_toc1, 'adj.P.Val')

deLimma_toc1

deLimma_toc1$diffExpressed <- 'NO'
deLimma_toc1$diffExpressed[deLimma_toc1$logFC>1 & deLimma_toc1$adj.P.Val<0.05] <- 'UP'
deLimma_toc1$diffExpressed[deLimma_toc1$logFC<(-1) & deLimma_toc1$adj.P.Val<0.05] <- 'DOWN'
table(deLimma_toc1$diffExpressed)
```

 DOWN    NO    UP 
  113 17419    90
  
really not a lot too


## Rhythmicity vs DE

```{r}
JTK_output_RUV

deLimma_toc1

JTK_deLimma_toc1 <- left_join(deLimma_toc1, JTK_output_RUV, by = "TAIR") |> 
  dplyr::mutate(Rhythmicity_fac = factor(Rhythmicity_lmrAMP)) |>
  dplyr::mutate(diffExpressed_fac = factor(diffExpressed)) |> 
  dplyr::relocate(Rhythmicity_fac, .after = "TAIR") |> 
  dplyr::relocate(diffExpressed_fac, .after = "TAIR")

JTK_deLimma_toc1$DE <- JTK_deLimma_toc1$diffExpressed != "NO"
toc1_table <- table(JTK_deLimma_toc1[, c('DE', 'Rhythmicity_lmrAMP')])
toc1_table[,2:3]
chisq.test(toc1_table[,2:3])
table(JTK_deLimma_toc1[, c('diffExpressed', 'Rhythmicity_lmrAMP')])
```

       Rhythmicity_lmrAMP
DE      Non-rhythmic Rhythmic
  FALSE         8425     4191
  TRUE            79       85

	Pearson's Chi-squared test with Yates' continuity correction

data:  toc1_table[, 2:3]
X-squared = 24.354, df = 1, p-value = 8.016e-07

             Rhythmicity_lmrAMP
diffExpressed Ambiguous Non-rhythmic Rhythmic
         DOWN        26           47       40
         NO        4803         8425     4191
         UP          13           32       45


```{r}
table <- table(JTK_deLimma_toc1[, c('diffExpressed', 'Rhythmicity_lmrAMP')])

DEratio <- (table[1, ] + table[3,]) / colSums(table)
DEratio
```


   Ambiguous Non-rhythmic     Rhythmic 
 0.008054523  0.009289746  0.019878391


```{r}
volcano_data_rhythmic <- JTK_deLimma_toc1 |> dplyr::filter(Rhythmicity_lmrAMP=='Rhythmic')
volcano_data_nonrhythmic <- JTK_deLimma_toc1 |> dplyr::filter(Rhythmicity_lmrAMP=='Non-rhythmic')

ggplot(data=volcano_data_rhythmic, aes(x=logFC, y=-log10(adj.P.Val), col = diffExpressed_fac)) +
  geom_vline(xintercept = c(-0.8, 0.8), col='gray', linetype='dashed') +
  geom_hline(yintercept=c(-log10(0.05)), col='gray', linetype='dashed') +
  geom_point(size=1) +
  scale_color_manual(values = c("#00AFBB", "grey", "#FC4E07"), 
                     labels = c("Downregulated", "Not significant", "Upregulated")) +
  ggtitle("Rhythmic genes") +
  xlim(-5, 5) +
  ylim(-1, 50) #+
  # geom_label_repel(data = dplyr::filter(volcano_data_rhythmic, abs(logFC) > 1.8 | -log10(adj.P.Val) > 25), aes(, label = SYMBOL), hjust=0, vjust=0, color = "black", size = 2.5)
  

ggplot(data=volcano_data_nonrhythmic, aes(x=logFC, y=-log10(adj.P.Val
), col = diffExpressed_fac)) +
  geom_vline(xintercept = c(-0.8, 0.8), col='gray', linetype='dashed') +
  geom_hline(yintercept=c(-log10(0.05)), col='gray', linetype='dashed') +
  geom_point(size=1) +
  scale_color_manual(values = c("#00AFBB", "grey", "#FC4E07"), 
                     labels = c("Downregulated", "Not significant", "Upregulated")) +
  ggtitle("Non-rhythmic genes") +
  xlim(-5, 5) +
  ylim(-1, 50) #+
  # geom_label_repel(data = dplyr::filter(volcano_data_nonrhythmic, logFC < -2 | logFC > 1.2 | -log10(adj.P.Val) > 35), aes(, label = gene_name), hjust=0, vjust=0, color = "black", size = 2.5)

```

### Annotation of DE-rhythmic genes

```{r}
DE_rhythmic_genes_toc1 <- JTK_deLimma_toc1[JTK_deLimma_toc1$BH.Q < 0.05 & 
                                   JTK_deLimma_toc1$amp_lmr > 0.58 & 
                                   abs(JTK_deLimma_toc1$logFC) > 1.0 & 
                                   JTK_deLimma_toc1$adj.P.Val < 0.05]

DE_rhythmic_genes_annot_toc1 <- right_join(annotations[match(DE_rhythmic_genes_toc1$TAIR, annotations$TAIR),], DE_rhythmic_genes_toc1) |>
  arrange(desc(abs(logFC))) |> as.matrix()

write.table(DE_rhythmic_genes_annot_toc1, file = "../../results/Mehta_results/Mehta_DE_rhythmic_genes_annot_toc1.tsv", sep = "\t", row.names = FALSE)

sum(!rownames(y) %in% annotations$TAIR)
sum(!DE_rhythmic_genes_toc1$TAIR %in% annotations$TAIR)
length(rownames(y))

```

????? why so many NA genes... (ok only 2 are not mapped, others are mapped but many don't have SYMBOL/genename)


```{r}
JTK_deLimma_toc1
ggplot(JTK_deLimma_toc1, aes(x = amp_lmr, y = logFC)) +
  geom_point(alpha = 0.5) +
  theme_minimal() +
  xlab("Rhythmic amplitude (from Limorhyde)") +
  ylab("log2 Fold Change (DE)") +
  ggtitle("DE vs rhythmic amplitude")


cor(JTK_deLimma_toc1$logFC, JTK_deLimma_toc1$amp_lmr)
```

```{r}
ggplot(JTK_deLimma_toc1[JTK_deLimma_toc1$Rhythmicity_fac == "Rhythmic", ], aes(x = amp_lmr, y = logFC, color = Rhythmicity_fac)) +
  geom_point(alpha = 0.6) +
  scale_color_manual(values = c(
    "Non-rhythmic" = "transparent",
    "Rhythmic" = "royalblue"
  )) +
  theme_minimal(base_size = 14) +
  xlab("Rhythmic absolute amplitude") +
  ylab("log2 Fold Change") +
  xlim(0, 3) +
  ggtitle("logFC vs rhythmic amplitude by rhythmicity") +
  theme(
    legend.title = element_text(face = "bold"),
    legend.position = "right"
  ) # +
  # geom_label_repel(data = dplyr::filter(JTK_deLimma_toc1, (amp_lmr > 1.3 | abs(logFC) > 2.5) & amp_lmr > 0.3), aes(, label = gene_name), 
  #           hjust=0, vjust=0, color = "black", size = 2)
```

Seems not as obvious trend...


### 3D plot!

```{r}

AMP_threshold_list <- seq(0, 2, by=0.05)
logFC_threshold_list <- seq(0, 5, by=0.1)


threshold_result <- data.frame(
  AMP = NA,
  logFC = NA,
  ratio = NA,
  rhythmic_DE = NA,
  DE = NA,
  rhythmic = NA,
  log_rhythmic_DE = NA
)


# 3d threshold plot data
for (logFC_threshold in logFC_threshold_list){
  for (AMP_threshold in AMP_threshold_list){
    N_rhythmic_DE <- nrow(JTK_deLimma_toc1[abs(JTK_deLimma_toc1$logFC) > logFC_threshold & JTK_deLimma_toc1$adj.P.Val < 0.05 & JTK_deLimma_toc1$BH.Q < 0.05 & JTK_deLimma_toc1$amp_lmr > AMP_threshold])
    N_rhythmic <- nrow(JTK_deLimma_toc1[JTK_deLimma_toc1$BH.Q < 0.05 & JTK_deLimma_toc1$amp_lmr > AMP_threshold])
    # print(c(AMP_threshold, N_rhythmic))
    N_DE <- nrow(JTK_deLimma_toc1[abs(JTK_deLimma_toc1$logFC) > logFC_threshold & JTK_deLimma_toc1$adj.P.Val < 0.05])
    
    if (N_rhythmic == 0){
      DE_ratio = NA
    } 
    else{
      DE_ratio <- N_rhythmic_DE/N_rhythmic
    }
    threshold_result <- rbind(threshold_result, c(AMP = AMP_threshold, logFC = logFC_threshold, ratio = DE_ratio, rhythmic_DE = N_rhythmic_DE, DE = N_DE, rhythmic = N_rhythmic, log10_rhythmic_DE = log10(N_rhythmic_DE)))
  }
}

threshold_result <- drop_na(threshold_result)
threshold_result_finite <- threshold_result[is.finite(threshold_result$log_rhythmic_DE), ]


threshold_result[threshold_result$AMP == 0.55 & threshold_result$logFC == 1.0, ]
```
```{r}
### Relative 3D plot (log-count coloured)

fig <- plot_ly(threshold_result_finite, x = ~logFC, y = ~AMP, z = ~ratio, color = ~log_rhythmic_DE, size = I(30), colors = rainbow(5))
fig <- fig %>% add_markers()
fig <- fig %>% layout(title = list(text="toc1: Relative DE ratio for various logFC and relative AMP thresholds", y = 0.95, x = 0.5, xanchor = 'center', yanchor = "top"),
  scene = list(xaxis = list(title = 'log2 Fold Change <br> Magnitude',
                                                range = c(0, 2.5)),
               yaxis = list(title = 'Amplitude (relative)',
                                                range = c(0, 1.0)),
               zaxis = list(title = 'Ratio of DE genes : Rhythmic genes',
                                                range = c(0, 0.9)),
               aspectmode = "cube"))

colorbar(fig, title = "# rhythmic DE genes", tickmode = "array", tickvals = c(0,1,2,3), ticktext = c(1, 10, 100, 1000))


```

*????????????????? no increase?????????????????????*