---
title: "Yang Pipeline Analysis"
author: "Elysia Luo"
date: "2026-02-21"
output: html_document
---

```{r}
mamba_folder <- "/Users/elysialuo/miniforge3"
env_name <- "circadian"
.libPaths(paste(mamba_folder, "envs", env_name, "lib/R/library", sep = "/"))

library(tidyverse)
library(dryR)
library(stringr)
library(ggplot2)
library(ggpubr)

library(MASS)
library(janitor)
library(naniar)
library(patchwork)
library(scales)
library(cluster)

library(RColorBrewer)
library(ggrepel)

library(GEOquery)
library(stringr)
library(tidyr)

library(AnnotationHub)
library(AnnotationDbi)
library(ensembldb)

library(limma)
library(edgeR)
library(RUVSeq)

library(limorhyde)
library(annotate)
library(data.table)
library(org.Mm.eg.db)
library(foreach)

library(VennDiagram)

library('org.Mm.eg.db')
library(GEOquery)
library(stringr)
library(tidyr)
library(plotly)
library("RColorBrewer")
```


```{r}
getwd()
```

```{r}
ah <- AnnotationHub()

# search for id of desired dataset

dbResult <- query(ah, c("EnsDb", "Mus musculus", "102")) 
dbResult[3]
str(dbResult)

# load mouse transcript dataset
MouseEnsDb <- ah[[dbResult$ah_id]]
str(MouseEnsDb)

# output data of interest in form of data.frame
annotations <- genes(MouseEnsDb, return.type = "data.frame")
annotations

# subset(ah, species == "Mus musculus")

# dbResult <- query(ah, c("OrgDb", "org.Mm.eg.db"))
# dbResult
# str(dbResult)
# 
# annotations <- genes(MouseOrgDb, return.type = "data.frame")
```


```{r}
# read in raw count matrix from Aviram
gse <- getGEO(filename = "../../data/Yang_raw_data/GSE70497_series_matrix.txt", GSEMatrix = TRUE, getGPL = FALSE)

metadata <- pData(gse) |> 
  dplyr::mutate("sample collection time:ch1" = gsub("\\D", "", `sample collection time:ch1`)) |> 
  dplyr::mutate("title" = gsub("mouse ", "", title)) |>
  dplyr::rename(genotype = "genotype/variation:ch1", time = "sample collection time:ch1") |>
  dplyr::mutate("time" = str_pad(time, 2, pad="0")) |>
  dplyr::select(title, geo_accession, genotype, time, description) |>
  dplyr::mutate("title" = paste(    ## BRUH THE NAMING IS SO BAD
    ifelse(genotype == "WT", "WT", "KO"),
    "_CT",
    time,
    "_",
    description,
    sep = ""))

metadata <- metadata[order(as.integer(metadata$time)),]
metadata <- metadata[order(metadata$genotype, decreasing = TRUE),]

raw_count <- read.table("../../data/Yang_raw_data/GSE70497_FINAL_master_list_of_genes_counts_MIN.Guangrui_WT_v_KO_timecourse.txt", header = TRUE) |> 
  dplyr::mutate("id" = gsub("gene:", "", `id`)) |>
  dplyr::select(-geneCoordinate, -geneSymbol) |> column_to_rownames("id")

colnames(raw_count) <- metadata[match(colnames(raw_count), metadata$title), ]$geo_accession  # change rownames from title to accession number


test <- pData(gse)
```


Pre-filtered count matrix


### Normalisation

```{r}
# 1. Convert to CPM

# Compute CPM
raw_count_cpm <- raw_count |>
  dplyr::mutate(across(.cols = everything(), .fns = ~ . * 1000000 / sum(.), .names = "{col}"))

# Filter for at least CPM >= 1 in one sample
raw_count_cpm_filtered <- raw_count_cpm_filtered <- raw_count_cpm[rowSums(raw_count_cpm >= 1) >= 2, ]
raw_count_cpm_filtered

```


```{r}
colSums(is.na(raw_count))
```


```{r}
# 2. TMM normalisation

# Filter raw count based on cpm
raw_count_filtered <- raw_count[rownames(raw_count) %in% rownames(raw_count_cpm_filtered),]
nrow(raw_count_filtered)

# Normalise library size of filtered raw count
raw_count_filtered_norm <- DGEList(counts=raw_count_filtered) |> normLibSizes()
str(raw_count_filtered_norm)
raw_count_filtered_norm$samples

# log2 cpm with normalised library size
raw_filtered_norm_logcpm <- cpm(raw_count_filtered_norm, log=TRUE)
raw_filtered_norm_logcpm_prior <- cpm(raw_count_filtered_norm, log=TRUE, prior.count=2)

```

```{r}
cov <- sd(raw_count_filtered_norm$samples$lib.size) / mean(raw_count_filtered_norm$samples$lib.size) * 100
cov
```

WOWWW lib size so consistent!!


```{r}
### RUV
x <- str_c(metadata$genotype, str_pad(metadata$time, 2, pad="0"), sep="_")

differences <- makeGroups(x)  # -1 assigned to missing sample (PerKO20 and PerKO8)
controls <- rownames(raw_filtered_norm_logcpm_prior)

# k=2, prior
RUVs_output_prior <- RUVs(raw_filtered_norm_logcpm_prior, controls, k=2, differences, isLog=TRUE)
raw_filtered_norm_logcpm_prior_RUVcount_k2 <- RUVs_output_prior$normalizedCounts
as.data.frame(raw_filtered_norm_logcpm_prior)
as.data.frame(raw_filtered_norm_logcpm_prior_RUVcount_k2)
W1 <- RUVs_output_prior$W[,1]
W2 <- RUVs_output_prior$W[,2]

```

```{r}
pdf("../../results/Plots/Yang_RLE.pdf", width = 9, height = 2.5)

par(mfrow = c(1,3))

## filtered, unnormalised
plotRLE(as.matrix(raw_count_filtered), outline=FALSE, exprs_logged = FALSE,
        ylim=c(-0.9, 0.9), main = 'Unnormalised', xlab = "Samples", ylab = "Relative log", cex.main = 1.2)

## filtered, TMM normalised
plotRLE(raw_filtered_norm_logcpm_prior, outline=FALSE, exprs_logged = TRUE,
        ylim=c(-0.9, 0.9), main = 'TMM normalised', xlab = "Samples", ylab = "Relative log", cex.main = 1.2)

## filtered, TMM & RUV normalised
plotRLE(RUVs_output_prior$normalizedCounts, outline=FALSE, exprs_logged = TRUE,
        ylim=c(-0.9, 0.9), main = 'TMM & RUV normalised', xlab = "Samples", ylab = "Relative log", cex.main = 1.2)

dev.off()
```

```{r}
as.data.frame(raw_filtered_norm_logcpm_prior)
```


```{r}
# PCA plot: CPM filtered raw count
pca_test_raw <- prcomp(t(raw_count_filtered), center=TRUE, scale=TRUE)
pca_df_raw <- as.data.frame(pca_test_raw$x)
pca_df_raw <- left_join(rownames_to_column(pca_df_raw, "geo_accession"), metadata, by = "geo_accession")

ggplot(pca_df_raw, aes(PC1, PC2, color = genotype, shape = time)) +
  geom_point(size = 2) +
  theme_classic() +
  labs(title = "PCA: timepoint & genotype (raw count)",
       x = paste0("PC1 (", round(summary(pca_test_raw)$importance[2,1] * 100, 1), "%)"),
       y = paste0("PC2 (", round(summary(pca_test_raw)$importance[2,2] * 100, 1), "%)"))


## PCA plot: TMM_log_cpm normalised

pca_test <- prcomp(t(raw_filtered_norm_logcpm_prior), center=TRUE, scale=TRUE)
pca_df_TMM <- as.data.frame(pca_test$x)
pca_df_TMM <- left_join(rownames_to_column(pca_df_TMM, "geo_accession"), metadata, by = "geo_accession")

ggplot(pca_df_TMM, aes(PC1, PC2, color = genotype, shape = time)) +
  geom_point(size = 2) +
  theme_classic() +
  labs(title = "PCA: timepoint & genotype (TMM normalised)",
       x = paste0("PC1 (", round(summary(pca_test)$importance[2,1] * 100, 1), "%)"),
       y = paste0("PC2 (", round(summary(pca_test)$importance[2,2] * 100, 1), "%)"))


## PCA plot: RUV normalised, k=2, prior

pca_test <- prcomp(t(raw_filtered_norm_logcpm_prior_RUVcount_k2), center=TRUE, scale=TRUE)
pca_df_RUV <- as.data.frame(pca_test$x)
pca_df_RUV <- left_join(rownames_to_column(pca_df_RUV, "geo_accession"), metadata, by = "geo_accession")

ggplot(pca_df_RUV, aes(PC1, PC2, color = genotype, shape = time)) +
  geom_point(size = 2) +
  theme_classic() +
  labs(title = "PCA: timepoint & genotype (TMM + RUV, k=2, prior)",
       x = paste0("PC1 (", round(summary(pca_test)$importance[2,1] * 100, 1), "%)"),
       y = paste0("PC2 (", round(summary(pca_test)$importance[2,2] * 100, 1), "%)"))
```


Bmal1KO can't cluster as expected, WT separates timepoints?



```{r}
# PCA plot: CPM filtered raw count
pca_test_raw <- prcomp(t(raw_count_filtered), center=TRUE, scale=TRUE)
pca_df_raw <- as.data.frame(pca_test_raw$x)
pca_df_raw <- left_join(rownames_to_column(pca_df_raw, "geo_accession"), metadata, by = "geo_accession") |> dplyr::mutate(genotype = ifelse(genotype == "WT", "WT", "Bmal1KO")) |> dplyr::mutate(genotype_fac = factor(genotype, levels = c("WT", "Bmal1KO")))
pca_df_raw$lightdark <- "light"
pca_df_raw$lightdark[pca_df_raw$time %in% c("0", "16", "20")] <- "dark"

ggplot(pca_df_raw, aes(PC1, PC2, color = lightdark, shape = genotype_fac, label = time)) +
  geom_point(size = 2) +
  geom_text(hjust=-0.2, vjust=-0.2, color = "black", size = 2.5) +
  theme_bw() +
  theme(
    panel.grid = element_blank(),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.6),
    axis.text = element_text(colour = "black"),
    axis.title = element_text(colour = "black", size = 12),
    plot.title = element_text(colour = "black", size = 12, hjust = 0),
    legend.key = element_blank()
  ) + 
  labs(title = "PCA: Yang unnormalised, filtered",
       x = paste0("PC1 (", round(summary(pca_test_raw)$importance[2,1] * 100, 1), "%)"),
       y = paste0("PC2 (", round(summary(pca_test_raw)$importance[2,2] * 100, 1), "%)"))

ggsave(
  filename = "../../results/Plots/Yang_volcano_PCA_raw.svg",
  plot = get_last_plot(),
  width = 8, 
  height = 5, 
  units = "in"
)

## PCA plot: TMM_log_cpm normalised

pca_test <- prcomp(t(raw_filtered_norm_logcpm_prior), center=TRUE, scale=TRUE)
pca_df_TMM <- as.data.frame(pca_test$x)
pca_df_TMM <- left_join(rownames_to_column(pca_df_TMM, "geo_accession"), metadata, by = "geo_accession") |> dplyr::mutate(genotype = ifelse(genotype == "WT", "WT", "Bmal1KO")) |> dplyr::mutate(genotype_fac = factor(genotype, levels = c("WT", "Bmal1KO")))
pca_df_TMM$lightdark <- "light"
pca_df_TMM$lightdark[pca_df_raw$time %in% c("0", "16", "20")] <- "dark"

ggplot(pca_df_TMM, aes(PC1, PC2, color = lightdark, shape = genotype_fac, label = time)) +
  geom_point(size = 2) +
  geom_text(hjust=-0.2, vjust=-0.2, color = "black", size = 2.5) +
  theme_bw() +
  theme(
    panel.grid = element_blank(),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.6),
    axis.text = element_text(colour = "black"),
    axis.title = element_text(colour = "black", size = 12),
    plot.title = element_text(colour = "black", size = 12, hjust = 0),
    legend.key = element_blank()
  ) + 
  labs(title = "PCA: Yang TMM normalised, filtered",
       x = paste0("PC1 (", round(summary(pca_test)$importance[2,1] * 100, 1), "%)"),
       y = paste0("PC2 (", round(summary(pca_test)$importance[2,2] * 100, 1), "%)"))

ggsave(
  filename = "../../results/Plots/Yang_volcano_PCA_TMM.svg",
  plot = get_last_plot(),
  width = 8, 
  height = 5, 
  units = "in"
)

## PCA plot: RUV normalised, k=2, prior

pca_test <- prcomp(t(raw_filtered_norm_logcpm_prior_RUVcount_k2), center=TRUE, scale=TRUE)
pca_df_RUV <- as.data.frame(pca_test$x)
pca_df_RUV <- left_join(rownames_to_column(pca_df_RUV, "geo_accession"), metadata, by = "geo_accession") |> dplyr::mutate(genotype = ifelse(genotype == "WT", "WT", "Bmal1KO")) |> dplyr::mutate(genotype_fac = factor(genotype, levels = c("WT", "Bmal1KO")))
pca_df_RUV$lightdark <- "light"
pca_df_RUV$lightdark[pca_df_raw$time %in% c("0", "16", "20")] <- "dark"

ggplot(pca_df_RUV, aes(PC1, PC2, color = lightdark, shape = genotype_fac, label = time)) +
  geom_point(size = 2) +
  geom_text(hjust=-0.2, vjust=-0.2, color = "black", size = 2.5) +
  theme_bw() +
  theme(
    panel.grid = element_blank(),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.6),
    axis.text = element_text(colour = "black"),
    axis.title = element_text(colour = "black", size = 12),
    plot.title = element_text(colour = "black", size = 12, hjust = 0),
    legend.key = element_blank()
  ) + 
  labs(title = "PCA: Yang TMM & RUV normalised (k=2), filtered ",
       x = paste0("PC1 (", round(summary(pca_test)$importance[2,1] * 100, 1), "%)"),
       y = paste0("PC2 (", round(summary(pca_test)$importance[2,2] * 100, 1), "%)"))

ggsave(
  filename = "../../results/Plots/Yang_volcano_PCA_RUV.svg",
  plot = get_last_plot(),
  width = 8, 
  height = 5, 
  units = "in"
)
```


```{r}
# Limorhyde

period = 24
qvalRhyCutoff = 0.15
qvalDrCutoff = 0.1

### should i use unRUV matrix???? yes
y = raw_filtered_norm_logcpm_prior #_RUVcount_k2

metadata_lmr <- metadata |> dplyr::select(title, genotype, time) |> 
  dplyr::mutate(time = as.integer(time), cond = factor(genotype, levels = c("WT", "Bmal1 KO"))) |> 
  rownames_to_column("geo_accession")


metadata_lmr <- cbind(metadata_lmr, limorhyde(metadata_lmr$time, 'time_'), W1, W2) |> as.data.table()


```


```{r}
condNow <- "Bmal1 KO"

design = model.matrix(~ time_cos + time_sin, data = metadata_lmr[cond == condNow, ])
fit = lmFit(y[, metadata_lmr$cond == condNow], design)
fit = eBayes(fit, trend = TRUE)
rhyNow = data.table(topTable(fit, coef = 2:3, number = Inf), keep.rownames = TRUE)
setnames(rhyNow, 'rn', 'gene_id')
rhyNow[, cond := condNow]

rhyLimmaSummary = rhyNow[, .(P.Value = min(P.Value)), by = gene_id]
rhyLimmaSummary[, adj.P.Val := p.adjust(P.Value, method = 'BH')]
setorderv(rhyLimmaSummary, 'adj.P.Val')

rhyLimmaSummary[rhyLimmaSummary$adj.P.Val < 0.05, ]


# Extract coefficients from your limma fit
coefs = coef(fit)[, c("(Intercept)", "time_cos", "time_sin")]

# Compute (log2 *relative*) amplitude for each gene
amplitude_lmr = sqrt(coefs[, "time_cos"]^2 + coefs[, "time_sin"]^2)
mean = coefs[, "(Intercept)"]

# Compute (absolute) amplitude
amplitude_lmr_abs = 2^(mean + amplitude_lmr) - 2^mean

cbind(amplitude_lmr, coefs[, "(Intercept)"])

hist(amplitude_lmr, ylim = c(0,100))
```

```{r}
as.data.frame(raw_filtered_norm_logcpm_prior[amplitude_lmr > 1,])
amplitude_lmr[amplitude_lmr > 1]
sum(amplitude_lmr > 0.58)
```



### Plot to check

```{r}
metadata_dt <- as.data.table(metadata)

geneSymbolsNow = c("Per1", "Per2", "Per3", "Cry1", "Cry2", "Arntl")
#, "Clock", "Npas2", "Nr1d1","Dbp", "Actb", "Kcnj15"
geneIdsNow = annotations[match(geneSymbolsNow, annotations$gene_name),]$gene_id


df = data.table(t(y[geneIdsNow, ]))
setnames(df, geneSymbolsNow)
df[, geo_accession := colnames(y[geneIdsNow, ])]

df = merge(df, metadata_dt[, .(geo_accession, genotype, time)], by = 'geo_accession')
df = melt(df, measure.vars = geneSymbolsNow, variable.name = 'symbol',
          value.name = 'expr')
df[, symbol := factor(symbol, geneSymbolsNow)]
df[, time := as.integer(time)]
df[, genotype := factor(genotype, levels = c("WT", "Bmal1 KO"), labels=c("Bmal1_WT", "Bmal1_KO"))]

ggplot(df) +
  facet_grid(symbol ~ genotype, scales = 'free_y') +
  geom_point(aes(x = time, y = expr, shape = genotype, color = symbol), size = 1.5) +
  labs(x = 'Circadian time (h)', y = 'Expression (log)') +
  scale_shape(solid = FALSE, guide = 'none') +
  scale_color_brewer(type = 'qual', palette = 'Dark2', guide = 'none') +
  scale_x_continuous(limits = c(0, 20), breaks = seq(0, 20, 4))

ggsave(
  filename = "../../results/Plots/Yang_oscillations1.svg",
  height = 4 ,
  width = 5.5,
  plot = get_last_plot()
)

```

```{r}

geneSymbolsNow = c("Clock", "Npas2", "Nr1d1","Dbp", "Actb", "Rplp0")
geneIdsNow = annotations[match(geneSymbolsNow, annotations$gene_name),]$gene_id


df = data.table(t(y[geneIdsNow, ]))
setnames(df, geneSymbolsNow)
df[, geo_accession := colnames(y[geneIdsNow, ])]

df = merge(df, metadata_dt[, .(geo_accession, genotype, time)], by = 'geo_accession')
df = melt(df, measure.vars = geneSymbolsNow, variable.name = 'symbol',
          value.name = 'expr')
df[, symbol := factor(symbol, geneSymbolsNow)]
df[, time := as.integer(time)]
df[, genotype := factor(genotype, levels = c("WT", "Bmal1 KO"), labels=c("Bmal1_WT", "Bmal1_KO"))]

ggplot(df) +
  facet_grid(symbol ~ genotype, scales = 'free_y') +
  geom_point(aes(x = time, y = expr, shape = genotype, color = symbol), size = 1.5) +
  labs(x = 'Circadian time (h)', y = 'Expression (log)') +
  scale_shape(solid = FALSE, guide = 'none') +
  scale_color_brewer(type = 'qual', palette = 'Dark2', guide = 'none') +
  scale_x_continuous(limits = c(0, 20), breaks = seq(0, 20, 4))

ggsave(
  filename = "../../results/Plots/Yang_oscillations2.svg",
  height = 3.5,
  width = 4.5,
  plot = get_last_plot()
)
```

*Why is the KO still quite rhythmic....???*

What's wrong with Kcnj15...



### JTK
```{r}
library(dataPreparation)
metadata
table(metadata$time[metadata$genotype == "WT"])

# log-TMM & RUV normalised matrix, control only
RUVs_output_prior$normalizedCounts[, metadata[metadata$genotype == "WT", ]$geo_accession]

JTK_matrix_RUV <- as.data.frame(RUVs_output_prior$normalizedCounts[, metadata[metadata$genotype == "WT", ]$geo_accession])
# JTK_matrix_RUV[rowSums(JTK_matrix_RUV)>0,]
# JTK_matrix_RUV[raw_count_WT$gene_name, ]

JTK_matrix_RUV <- JTK_matrix_RUV |> rownames_to_column('Probe')

write.table(JTK_matrix_RUV, file = 'JTK_Cycle/Yang_JTK/Yang_JTK_RUV_data.txt', sep = '\t', row.names = FALSE, col.names = TRUE, quote=FALSE)

write.table(JTK_matrix_RUV['Probe'], file = 'JTK_Cycle/Yang_JTK/Yang_JTK_RUV_annot.txt', sep = '/t', row.names = FALSE, col.names = TRUE, quote=FALSE)

```

--- running JTK cycle ---

```{r}
JTK_output_RUV <- read.table(file = 'JTK_Cycle/Yang_JTK/JTK.Yang_JTK_RUV_run.txt', header = TRUE ) |> dplyr::rename('gene_id' = 'Probe')

# Lable as rythmic & non-rhythmic using limorhyde's amplitude

JTK_output_RUV$amp_lmr <- amplitude_lmr[JTK_output_RUV$gene_id]
JTK_output_RUV$amp_lmr_abs <- amplitude_lmr_abs[JTK_output_RUV$gene_id]
JTK_output_RUV <- dplyr::relocate(JTK_output_RUV, amp_lmr, amp_lmr_abs, .after = AMP)

JTK_output_RUV$Rhythmicity_lmrAMP <- 'Ambiguous'
JTK_output_RUV$Rhythmicity_lmrAMP[JTK_output_RUV$BH.Q > 0.1 & JTK_output_RUV$amp_lmr < 0.58] <- 'Non-rhythmic'
JTK_output_RUV$Rhythmicity_lmrAMP[JTK_output_RUV$BH.Q < 0.05 & JTK_output_RUV$amp_lmr > 0.58] <- 'Rhythmic'
JTK_output_RUV <- dplyr::relocate(JTK_output_RUV, Rhythmicity_lmrAMP, .before = BH.Q)
table(JTK_output_RUV$Rhythmicity_lmrAMP)

# nonsense threshold for absolute values...
# JTK_output_RUV$Rhythmicity_lmrAMP_abs <- 'Non-rhythmic'
# JTK_output_RUV$Rhythmicity_lmrAMP_abs[JTK_output_RUV$BH.Q < 0.05 & JTK_output_RUV$amp_lmr_abs > 0.2] <- 'Rhythmic'
# JTK_output_RUV <- dplyr::relocate(JTK_output_RUV, Rhythmicity_lmrAMP_abs, .before = BH.Q)
# table(JTK_output_RUV$Rhythmicity_lmrAMP_abs)


```

   Ambiguous Non-rhythmic     Rhythmic 
        5825         6973          603
        

bruhh why so many rhythic genes......


### JTK (mutant)

```{r}
table(metadata$time[metadata$genotype == "Bmal1 KO"])

# log-TMM & RUV normalised matrix, control only
RUVs_output_prior$normalizedCounts[, metadata[metadata$genotype == "Bmal1 KO", ]$geo_accession]

JTK_matrix_RUV_mut <- as.data.frame(RUVs_output_prior$normalizedCounts[, metadata[metadata$genotype == "Bmal1 KO", ]$geo_accession]) |>
  rownames_to_column('Probe')


write.table(JTK_matrix_RUV_mut, file = 'JTK_Cycle/Yang_JTK/Yang_QC_JTK_RUV_mut_data.txt', sep = '\t', row.names = FALSE, col.names = TRUE, quote=FALSE)

write.table(JTK_matrix_RUV_mut['Probe'], file = 'JTK_Cycle/Yang_JTK/Yang_QC_JTK_RUV_mut_annot.txt', sep = '/t', row.names = FALSE, col.names = TRUE, quote=FALSE)

str(RUVs_output_prior$normalizedCounts)
metadata
```

--- running JTK cycle ---

```{r}
JTK_output_RUV_mut <- read.table(file = 'JTK_Cycle/Yang_JTK/JTK.Yang_JTK_RUV_mut_run.txt', header = TRUE ) |> dplyr::rename('gene_id' = 'Probe')

# Lable as rythmic & non-rhythmic using limorhyde's amplitude

JTK_output_RUV_mut$amp_lmr <- amplitude_lmr[JTK_output_RUV_mut$gene_id]
JTK_output_RUV_mut$amp_lmr_abs <- amplitude_lmr_abs[JTK_output_RUV_mut$gene_id]
JTK_output_RUV_mut <- dplyr::relocate(JTK_output_RUV_mut, amp_lmr, amp_lmr_abs, .after = AMP)

# JTK_output_RUV_mut$Rhythmicity <- 'Non-rhythmic'
# JTK_output_RUV_mut$Rhythmicity[JTK_output_RUV_mut$BH.Q < 0.05 & JTK_output_RUV_mut$AMP > 0.3] <- 'Rhythmic'
# JTK_output_RUV_mut <- dplyr::relocate(JTK_output_RUV_mut, Rhythmicity, .before = BH.Q)
# table(JTK_output_RUV_mut$Rhythmicity)

JTK_output_RUV_mut$Rhythmicity_lmrAMP <- 'Ambiguous'
JTK_output_RUV_mut$Rhythmicity_lmrAMP[JTK_output_RUV_mut$BH.Q > 0.1 & JTK_output_RUV_mut$amp_lmr < 0.58] <- 'Non-rhythmic'
JTK_output_RUV_mut$Rhythmicity_lmrAMP[JTK_output_RUV_mut$BH.Q < 0.05 & JTK_output_RUV_mut$amp_lmr > 0.58] <- 'Rhythmic'
JTK_output_RUV_mut <- dplyr::relocate(JTK_output_RUV_mut, Rhythmicity_lmrAMP, .before = BH.Q)
table(JTK_output_RUV_mut$Rhythmicity_lmrAMP)

# JTK_output_RUV_mut$Rhythmicity_lmrAMP_abs <- 'Non-rhythmic'
# JTK_output_RUV_mut$Rhythmicity_lmrAMP_abs[JTK_output_RUV_mut$BH.Q < 0.05 & JTK_output_RUV_mut$amp_lmr_abs > 0.2] <- 'Rhythmic'
# JTK_output_RUV_mut <- dplyr::relocate(JTK_output_RUV_mut, Rhythmicity_lmrAMP_abs, .before = BH.Q)
# table(JTK_output_RUV_mut$Rhythmicity_lmrAMP_abs)

sum(JTK_output_RUV_mut$Rhythmicity_lmrAMP_abs == 'Rhythmic' & JTK_output_RUV_mut$Rhythmicity_lmrAMP == 'Rhythmic')

```

```{r}
## Visualise overlap (WT vs mutant rhythmic genes)

JTK_RUV_rhythmic_list_wt <- JTK_output_RUV$gene_id[JTK_output_RUV$Rhythmicity == 'Rhythmic'] 
JTK_RUV_rhythmic_list_mut <- JTK_output_RUV_mut$gene_id[JTK_output_RUV_mut$Rhythmicity == 'Rhythmic'] 

length(JTK_RUV_rhythmic_list_wt)
length(JTK_RUV_rhythmic_list_mut)

vlist <- list(JTK_RUV_rhythmic_list_wt, JTK_RUV_rhythmic_list_mut)
names(vlist) <- c("JTK_RUV_wt", "JTK_RUV_mut")
myCol <- brewer.pal(3, "Pastel2")

venn.plot <- venn.diagram(
  x = vlist,
  category.names = c("WT", "Bmal1 KO"),
  filename = NULL,
  output = FALSE,
  # main = "Weger: WT vs Mutant rhythmic genes", # Title
  
  # Output features
  imagetype="png" ,
  height = 480 , 
  width = 480 , 
  resolution = 300,
  compression = "lzw",
  
  # Circles
  lwd = 2,
  lty = 'blank',
  fill = myCol[1:2],
  
  # Numbers
  cex = 1.5,
  fontface = "bold",
  fontfamily = "sans",
  
  # Set names
  cat.cex = 1.5,
  cat.fontface = "bold",
  cat.default.pos = "outer",
  cat.pos = c(-30, 30),
  cat.dist = c(0.03, 0.035),
  cat.fontfamily = "sans",
  cat.col = myCol[1:2],
  rotation.degree = 1,
  
  # The --3 label
  # scaled = FALSE,
  ext.pos = 181.5,
  ext.dist = -0.035,
  ext.length = 0.84
)

pdf("Yang_JTK_RUV_WTvsMut.pdf", width = 4, height = 4)
grid::grid.newpage()
grid::grid.draw(venn.plot)
dev.off()
```

### Limorhyde x Limma DE analysis

```{r}
# with interaction

design = model.matrix(~ cond + time_cos + time_sin + cond:time_cos + cond:time_sin + W1 + W2, data = metadata_lmr)

qr(design)$rank  # check if W1/2 are colinear with cond
cor(metadata_lmr[, c("W1", "W2", "time_cos", "time_sin")])  # check if W1/2 is eating away time information


fit = lmFit(y, design)
fit = eBayes(fit, trend = TRUE)
plotSA(fit)
deLimma = data.table(topTable(fit, coef = 2, number = Inf), keep.rownames = TRUE)
setnames(deLimma, 'rn', 'gene_id')

# record mean level of WT and KO!
mean_wt <- fit$coefficients[,1]
mean_ko <- fit$coefficients[,1] + fit$coefficients[,2]

# deLimma = deLimma[!(gene_id %in% drLimma[adj.P.Val <= qvalDrCutoff]$gene_id)] ## don't ignore differentially rhythmic genes
deLimma[, adj.P.Val := p.adjust(P.Value, method = 'BH')]
setorderv(deLimma, 'adj.P.Val')

deLimma

deLimma$diffExpressed <- 'NO'
deLimma$diffExpressed[deLimma$logFC>1 & deLimma$adj.P.Val<0.05] <- 'UP'
deLimma$diffExpressed[deLimma$logFC<(-1) & deLimma$adj.P.Val<0.05] <- 'DOWN'
table(deLimma$diffExpressed)
```

## Rhythmicity vs DE

```{r}
JTK_output_RUV

deLimma

JTK_deLimma <- left_join(deLimma, JTK_output_RUV, by = "gene_id") |> 
  dplyr::mutate(Rhythmicity_fac = factor(Rhythmicity_lmrAMP)) |>
  dplyr::mutate(diffExpressed_fac = factor(diffExpressed)) |> 
  dplyr::relocate(Rhythmicity_fac, .after = "gene_id") |> 
  dplyr::relocate(diffExpressed_fac, .after = "gene_id") |>
  left_join(annotations[, 1:2], by = "gene_id") |> 
  dplyr::relocate(gene_name, .after = gene_id)

table(JTK_deLimma[, c('diffExpressed', 'Rhythmicity_lmrAMP')])
```

             Rhythmicity_lmrAMP
diffExpressed Ambiguous Non-rhythmic Rhythmic
         DOWN         1           10        3
         NO        5815         6937      596
         UP           9           26        4

```{r}
table <- table(JTK_deLimma[, c('diffExpressed', 'Rhythmicity_lmrAMP')])

DEratio <- (table[1, ] + table[3,]) / colSums(table)
DEratio
```


```{r}
volcano_data_rhythmic <- JTK_deLimma |> dplyr::filter(Rhythmicity_lmrAMP=='Rhythmic')
volcano_data_nonrhythmic <- JTK_deLimma |> dplyr::filter(Rhythmicity_lmrAMP=='Non-rhythmic')

ggplot(data=volcano_data_rhythmic, aes(x=logFC, y=-log10(adj.P.Val), col = diffExpressed_fac)) +
  geom_vline(xintercept = c(-1, 1), col='gray', linetype='dashed') +
  geom_hline(yintercept=c(-log10(0.05)), col='gray', linetype='dashed') +
  geom_point(size=1) +
  scale_color_manual(values = c("#00AFBB", "grey", "#FC4E07"), 
                     labels = c("Downregulated", "Not significant", "Upregulated")) +
  ggtitle("Rhythmic genes") +
  xlim(-4, 4) +
  ylim(-1, 10) +
  geom_text(data = dplyr::filter(volcano_data_rhythmic, abs(logFC) > 1 | -log10(adj.P.Val) > 25), aes(, label = gene_name), hjust=0, vjust=0, color = "black", size = 2.5)
  

ggplot(data=volcano_data_nonrhythmic, aes(x=logFC, y=-log10(adj.P.Val
), col = diffExpressed_fac)) +
  geom_vline(xintercept = c(-0.8, 0.8), col='gray', linetype='dashed') +
  geom_hline(yintercept=c(-log10(0.05)), col='gray', linetype='dashed') +
  geom_point(size=1) +
  scale_color_manual(values = c("#00AFBB", "grey", "#FC4E07"), 
                     labels = c("Downregulated", "Not significant", "Upregulated")) +
  ggtitle("Non-rhythmic genes") +
  xlim(-5, 5) +
  ylim(-1, 10) +
  geom_label_repel(data = dplyr::filter(volcano_data_nonrhythmic, abs(logFC) > 1 | -log10(adj.P.Val) > 35), aes(, label = gene_name), hjust=0, vjust=0, color = "black", size = 2.5)
```

HUH why is the p values here also squashed???


```{r}
volcano_data_rhythmic <- JTK_deLimma |> dplyr::filter(Rhythmicity_lmrAMP=='Rhythmic')
volcano_data_nonrhythmic <- JTK_deLimma |> dplyr::filter(Rhythmicity_lmrAMP=='Non-rhythmic')

ggplot(data=volcano_data_rhythmic, aes(x=logFC, y=-log10(adj.P.Val), col = diffExpressed_fac)) +
  geom_vline(xintercept = c(-1, 1), col='gray', linetype='dashed') +
  geom_hline(yintercept=c(-log10(0.05)), col='gray', linetype='dashed') +
  geom_point(size=1) +
  scale_color_manual(values = c("#00AFBB", "grey", "#FC4E07", "black"), 
                     labels = c("Downregulated", "Not significant", "Upregulated", "black")) +
  ggtitle("Volcano plot of rhythmic genes in wild type") +
  labs(colour = "Differential expression")+ 
  xlim(-6, 6) +
  ylim(-1, 55) +
  labs(
    x = expression(log[2]~fold~change),
    y = expression(-log[10]~adjusted~p~value),
    colour = NULL) +
  geom_text_repel(data = dplyr::filter(volcano_data_rhythmic, abs(logFC) > 1 | -log10(adj.P.Val) > 30), aes(, label = gene_name), hjust=0, vjust=0, color = "black", size = 2.5) + 
  theme_bw(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.6),
    axis.text = element_text(colour = "black"),
    axis.title = element_text(colour = "black", size = 12),
    plot.title = element_text(colour = "black", size = 12, hjust = 0),
    legend.key = element_blank()
  )

ggsave(
  filename = "../../results/Plots/Yang_volcano_plot_rhythmic.svg",
  plot = get_last_plot(),
  width = 8, 
  height = 5, 
  units = "in"
)


ggplot(data=volcano_data_nonrhythmic, aes(x=logFC, y=-log10(adj.P.Val
), col = diffExpressed_fac)) +
  geom_vline(xintercept = c(-1, 1), col='gray', linetype='dashed') +
  geom_hline(yintercept=c(-log10(0.05)), col='gray', linetype='dashed') +
  geom_point(size=1) +
  scale_color_manual(values = c("#00AFBB", "grey", "#FC4E07"), 
                     labels = c("Downregulated", "Not significant", "Upregulated")) +
  ggtitle("Volcano plot of non-rhythmic genes in wild type") +
  labs(colour = "Differential expression")+ 
  xlim(-6, 6) +
  ylim(-1, 55) +
  labs(
    x = expression(log[2]~fold~change),
    y = expression(-log[10]~adjusted~p~value),
    colour = NULL) +
  geom_text_repel(data = dplyr::filter(volcano_data_nonrhythmic, abs(logFC) > 2.5 | -log10(adj.P.Val) > 40), aes(, label = gene_name), hjust=0, vjust=0, color = "black", size = 2.5) +
  theme_bw(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.6),
    axis.text = element_text(colour = "black"),
    axis.title = element_text(colour = "black", size = 12),
    plot.title = element_text(colour = "black", size = 12, hjust = 0),
    legend.key = element_blank()
  )

ggsave(
  filename = "../../results/Plots/Yang_volcano_plot_nonrhythmic.svg",
  plot = get_last_plot(),
  width = 8, 
  height = 5, 
  units = "in"
)
```

### Annotation of rhythmic genes

```{r}
DE_rhythmic_genes <- JTK_deLimma[JTK_deLimma$BH.Q < 0.05 & 
                                   JTK_deLimma$amp_lmr > 0.58 & 
                                   abs(JTK_deLimma$logFC) > 1.0 & 
                                   JTK_deLimma$adj.P.Val < 0.05]

DE_rhythmic_genes_annot <- right_join(annotations[match(DE_rhythmic_genes$gene_name, annotations$gene_name),], DE_rhythmic_genes) |>
  arrange(desc(abs(logFC))) |> as.matrix()

write.table(DE_rhythmic_genes_annot, file = "../../results/Yang_DE_rhythmic_genes_annot.tsv", sep = "\t", row.names = FALSE)



```

```{r}
JTK_deLimma
ggplot(JTK_deLimma, aes(x = amp_lmr, y = logFC)) +
  geom_point(alpha = 0.5) +
  theme_minimal() +
  xlab("Rhythmic amplitude (from Limorhyde)") +
  ylab("log2 Fold Change (DE)") +
  ggtitle("DE vs rhythmic amplitude")


cor(JTK_deLimma$logFC, JTK_deLimma$amp_lmr)
```

fold change is sqaushed... why???
and there are more rhythmic genes tooo


```{r}
ggplot(JTK_deLimma[JTK_deLimma$Rhythmicity_fac == "Rhythmic", ], aes(x = amp_lmr, y = logFC, color = Rhythmicity_fac)) +
  geom_point(alpha = 0.6) +
  scale_color_manual(values = c(
    "Non-rhythmic" = "transparent",
    "Rhythmic" = "royalblue"
  )) +
  theme_minimal(base_size = 14) +
  xlab("Rhythmic absolute amplitude") +
  ylab("log2 Fold Change") +
  xlim(0, 3) +
  ggtitle("logFC vs rhythmic amplitude by rhythmicity") +
  theme(
    legend.title = element_text(face = "bold"),
    legend.position = "right"
  ) # +
  # geom_label_repel(data = dplyr::filter(JTK_deLimma, (amp_lmr > 1.3 | abs(logFC) > 2.5) & amp_lmr > 0.3), aes(, label = gene_name), 
  #           hjust=0, vjust=0, color = "black", size = 2)
```







```{r}
AMP_threshold_list <- seq(0, 2, by=0.05)
logFC_threshold_list <- seq(0, 5, by=0.1)


threshold_result <- data.frame(
  AMP = NA,
  logFC = NA,
  ratio = NA,
  rhythmic_DE = NA,
  DE = NA,
  rhythmic = NA,
  log_rhythmic_DE = NA
)


# 3d threshold plot data
for (logFC_threshold in logFC_threshold_list){
  for (AMP_threshold in AMP_threshold_list){
    N_rhythmic_DE <- nrow(JTK_deLimma[abs(JTK_deLimma$logFC) > logFC_threshold & JTK_deLimma$adj.P.Val < 0.05 & JTK_deLimma$BH.Q < 0.05 & JTK_deLimma$amp_lmr > AMP_threshold])
    N_rhythmic <- nrow(JTK_deLimma[JTK_deLimma$BH.Q < 0.05 & JTK_deLimma$amp_lmr > AMP_threshold])
    # print(c(AMP_threshold, N_rhythmic))
    N_DE <- nrow(JTK_deLimma[abs(JTK_deLimma$logFC) > logFC_threshold & JTK_deLimma$adj.P.Val < 0.05])
    
    if (N_rhythmic == 0){
      DE_ratio = NA
    } 
    else{
      DE_ratio <- N_rhythmic_DE/N_rhythmic
    }
    threshold_result <- rbind(threshold_result, c(AMP = AMP_threshold, logFC = logFC_threshold, ratio = DE_ratio, rhythmic_DE = N_rhythmic_DE, DE = N_DE, rhythmic = N_rhythmic, log10_rhythmic_DE = log10(N_rhythmic_DE)))
  }
}

threshold_result <- drop_na(threshold_result)
threshold_result_finite <- threshold_result[is.finite(threshold_result$log_rhythmic_DE), ]

threshold_result[threshold_result$AMP == 0.55 & threshold_result$logFC == 1.0, ]

```

```{r}
plot_df <- threshold_result |>
  filter(
    abs(logFC - 0.3) < 1e-6 |
    abs(logFC - 0.5) < 1e-6 |
    abs(logFC - 0.7) < 1e-6 |
    abs(logFC - 1) < 1e-6
  ) |>
  mutate(logFC_label = paste0("|logFC| ≥ ", round(logFC, 2)))

plot_df

ggplot(plot_df, aes(x = AMP, y = ratio, color = log_rhythmic_DE)) +
  geom_point(size = 1) +
  facet_wrap(~ logFC_label, ncol = 2, nrow = 2) +
  labs(
    x = "Relative amplitude threshold",
    y = "Proportion of rhythmic genes that are DE",
    color = "# rhythmic DE genes"
  ) +
  ylim(0, 0.9) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.6),
    axis.text = element_text(colour = "black"),
    axis.title = element_text(colour = "black", size = 12),
    axis.title.x = element_text(margin = margin(t = 10)),
    axis.title.y = element_text(margin = margin(r = 10)),
    plot.title = element_text(colour = "black", size = 12, hjust = 0),
    legend.key = element_blank(),
    legend.title.align = 0.5
  ) +
  scale_color_viridis_c(
    breaks = c(0, 1, 2, 3),
    labels = c("1", "10", "100", "1000"),
    values = scales::rescale(c(0, 1, 2, 3)),
    limits = c(0, 3)
  )

ggsave(
  filename = "../../results/Plots/Yang_AMP_threshold_fixedFC.svg",
  plot = get_last_plot(),
  width = 8, 
  height = 5, 
  units = "in"
)
```


```{r}
fig <- plot_ly(threshold_result_finite, x = ~logFC, y = ~AMP, z = ~ratio, color = ~log_rhythmic_DE, size = I(30), colors = rainbow(5))
fig <- fig %>% add_markers()
fig <- fig %>% layout(title = list(text="(Relative) DE ratio for various logFC and relative AMP thresholds", y = 0.95, x = 0.5, xanchor = 'center', yanchor = "top"),
  scene = list(xaxis = list(title = 'log2 Fold Change <br> Magnitude',
                                                range = c(0, 3.5)),
                                   yaxis = list(title = 'Amplitude (relative)',
                                                range = c(0, 1.5)),
                                   zaxis = list(title = 'Ratio of DE genes : Rhythmic genes',
                                                range = c(0, 0.9)),
               aspectmode = "cube"))

fig <- colorbar(fig, title = "# rhythmic DE genes", tickmode = "array", tickvals = c(0,1,2,3), ticktext = c(1, 10, 100, 1000))

htmlwidgets::saveWidget(fig, "../../results/Plots/Yang_AMP_FC_threshold_3D.html", selfcontained = TRUE)
```

```{r}
JTK_deLimma[JTK_deLimma$gene_name == "Actb" | JTK_deLimma$gene_name == "Kcnj15" | JTK_deLimma$gene_name == "Rplp0",]
```

